---
title: "Tertiary Analysis of AML Hematology Atlas"
author: "Matt Ung and Alyssa Pyclik"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: yeti
    toc: true
    number_sections: true
---

```{r setup, include = FALSE}

library(flowCore)
library(SeuratObject)
library(BPCells)
library(Seurat)
library(SeuratDisk)
library(dplyr)
library(aws.s3)
library(tibble)
library(tidyr)
library(dotenv)
library(ggplot2)
library(ggpubr)
library(glue)
library(DoubletFinder)
library(tidyseurat)
library(pzfx)
library(preprocessCore)
library(pals)
library(EnhancedVolcano)
library(SeuratWrappers)
library(alphahull)
library(igraph)
library(reticulate)
library(viridis)
library(msigdbr)
library(GSVA)
library(escape)
library(dittoSeq)
library(fgsea)
library(GSEABase)
library(ggbeeswarm)
library(circlize)
library(pheatmap)
library(biomaRt)
library(SingleCellExperiment)
library(Matrix)
library(stringr)
library(ComplexHeatmap)
library(future)
library(circlize)
library(msigdbr)
library(survival)
library(survminer)
library(broom)
library(scales)
library(lubridate)
library(future)
library(ggsci)
library(bazar)
library(flowCore)
library(mclust)
library(doParallel)
library(foreach)
library(scales)
library(tictoc)
library(randomForest)
library(e1071)
library(survminer)
library(robCompositions)
library(dsb)
library(table1)
library(gtsummary)

options(Seurat.object.assay.version = "v5")
options(future.globals.maxSize = 1e9)

load_dot_env(file = "/home/rstudio/data/bitbucket/.env")

knitr::opts_knit$set(root.dir = "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas")

Sys.setenv(
  "AWS_ACCESS_KEY_ID" = Sys.getenv("key"),
  "AWS_SECRET_ACCESS_KEY" = Sys.getenv("secret"),
  "AWS_DEFAULT_REGION" = "us-east-1"
)

cl <- makeCluster(12)
registerDoParallel(cl)


# unregister <- function() {
#   env <- foreach:::.foreachGlobals
#   rm(list=ls(name=env), pos=env)
# }


```

```{r functions}

dsb_ebayes <- function (cell_protein_matrix, empty_drop_matrix, denoise.counts = TRUE, 
          use.isotype.control = TRUE, isotype.control.name.vec = NULL, 
          define.pseudocount = FALSE, pseudocount.use, quantile.clipping = FALSE, 
          quantile.clip = c(0.001, 0.9995), scale.factor = c("standardize", 
                                                             "mean.subtract")[1], return.stats = FALSE) 
{
  if (is.null(empty_drop_matrix)) {
    stop(paste0("to use DSBNormalizeProtein specify `empty_drop_matrix`", 
                "to normalize ADT data without using empty droplets use", 
                "the function `ModelNegativeADTnorm`"))
  }
  a = isotype.control.name.vec
  b = rownames(empty_drop_matrix)
  cm = rownames(cell_protein_matrix)
  if (!isTRUE(all.equal(cm, b))) {
    diff = c(setdiff(cm, b), setdiff(b, cm))
    if (!isTRUE(all.equal(nrow(cell_protein_matrix), nrow(empty_drop_matrix)))) {
      stop(paste0("the number of proteins in `cell_protein_matrix` and `empty_drop_matrix` do not match check: ", 
                  diff))
    }
    if (length(diff) > 0) {
      stop(paste0("rows of cell and background matrices have mis-matching names: \n", 
                  diff))
    }
    if (length(diff < 0)) {
      rmatch = match(x = rownames(cell_protein_matrix), 
                     table = rownames(empty_drop_matrix))
      empty_drop_matrix = empty_drop_matrix[rmatch, ]
      warning(paste0("rows (proteins) of cell_protein_matrix and `empty_drop_matrix`", 
                     "were not in the same order. dsb reordered `empty_drop_matrix` rows", 
                     "to match `cell_protein_matrix` rows"))
    }
  }
  iso_detect = cm[grepl(pattern = "sotype|Iso|iso|control|CTRL|ctrl|Ctrl|ontrol", 
                        x = cm)]
  if (!is.null(a) & !isTRUE(all(a %in% b)) & !isTRUE(all(a %in% 
                                                         cm))) {
    stop(paste0("some elements of isotype.control.name.vec are not in input data rownames: \n", 
                "cell_protein_matrix - ", setdiff(a, cm), " \nempty_drop_matrix - ", 
                setdiff(a, b)))
  }
  if (isFALSE(denoise.counts)) {
    print(paste0("Not running dsb step II (removal of cell to cell technical noise)", 
                 " Setting use.isotype.control and isotype.control.name.vec to FALSE and NULL"))
    use.isotype.control = FALSE
    isotype.control.name.vec = NULL
    if (length(iso_detect) > 0) {
      print("potential isotype controls detected: ")
      print(iso_detect)
    }
  }
  if (isTRUE(use.isotype.control) & is.null(isotype.control.name.vec)) {
    if (length(iso_detect) > 0) {
      print("potential isotype controls detected: ")
      print(iso_detect)
    }
    stop("if use.isotype.control = TRUE, set isotype.control.name.vec to rownames of isotype controls")
  }
  if (isTRUE(denoise.counts) & isFALSE(use.isotype.control)) {
    warning(paste0("`use.isotype.control` = FALSE is not recommended if setting `denoise.counts` = TRUE", 
                   " \nwhen isotype controls are available.\n If data include isotype controls", 
                   " set `denoise.counts` = TRUE `use.isotype.control` = TRUE", 
                   " \and set `isotype.control.name.vec` to a vector of isotype control rownames"))
    if (length(iso_detect) > 0) {
      print("potential isotype controls detected: ")
      print(iso_detect)
    }
  }
  adt = cell_protein_matrix %>% as.matrix()
  adtu = empty_drop_matrix %>% as.matrix()
  if (isTRUE(define.pseudocount)) {
    adtu_log = log(adtu + pseudocount.use)
    adt_log = log(adt + pseudocount.use)
  }
  else {
    adtu_log = log(adtu + 10)
    adt_log = log(adt + 10)
  }
  print("correcting ambient protein background noise")
  mu_u = apply(adtu_log, 1, mean)
  sd_u = apply(adtu_log, 1, sd)
  
  
  # use empirical bayes to shrink background means [ADD-ON]
  mu_u = coef(ebnm::ebnm(x = mu_u, s = sd_u, mode = "estimate", prior_family = "normal"))
  
  
  if (scale.factor == "standardize") {
    if (any(sd_u < 0.05)) {
      print(paste0("some proteins with low background variance detected", 
                   " check raw and normalized distributions. ", 
                   " protein stats can be returned with return.stats = TRUE"))
      print(names(which(sd_u < 0.05)))
      
      
      # add small variance to features with little background variation [ADD-ON]
      sd_u[sd_u < 0.05] <- min(sd_u[sd_u > 0.05])
      
      
    }
    norm_adt = apply(adt_log, 2, function(x) (x - mu_u)/sd_u)
  }
  if (scale.factor == "mean.subtract") {
    norm_adt = apply(adt_log, 2, function(x) (x - mu_u))
  }
  if (isTRUE(denoise.counts)) {
    print(paste0("fitting models to each cell for dsb technical component and", 
                 " removing cell to cell technical noise"))
    cellwise_background_mean = apply(norm_adt, 2, function(x) {
      g = mclust::Mclust(x, G = 2, warn = FALSE, verbose = FALSE)
      return(g$parameters$mean[1])
    })
    gc()
    if (isTRUE(use.isotype.control)) {
      noise_matrix = rbind(norm_adt[isotype.control.name.vec, 
      ], cellwise_background_mean)
      get_noise_vector = function(noise_matrix) {
        g = stats::prcomp(t(noise_matrix), scale = TRUE)
        return(g$x[, 1])
      }
      noise_vector = get_noise_vector(noise_matrix)
      norm_adt = limma::removeBatchEffect(norm_adt, covariates = noise_vector)
    }
    else {
      noise_vector = cellwise_background_mean
      norm_adt = limma::removeBatchEffect(norm_adt, covariates = noise_vector)
    }
  }
  if (isTRUE(quantile.clipping)) {
    ql = apply(norm_adt, 1, FUN = stats::quantile, quantile.clip[1])
    qh = apply(norm_adt, 1, FUN = stats::quantile, quantile.clip[2])
    for (i in 1:nrow(norm_adt)) {
      norm_adt[i, ] = ifelse(norm_adt[i, ] < ql[i], ql[i], 
                             norm_adt[i, ])
      norm_adt[i, ] = ifelse(norm_adt[i, ] > qh[i], qh[i], 
                             norm_adt[i, ])
    }
  }
  if (isTRUE(return.stats)) {
    print("returning results in a list; normalized matrix accessed x$dsb_normalized_matrix")
    protein_stats = list(`raw cell matrix stats` = data.frame(cell_mean = apply(adt_log, 
                                                                                1, mean), cell_sd = apply(adt_log, 1, sd)), `dsb normalized matrix stats` = data.frame(dsb_mean = apply(norm_adt, 
                                                                                                                                                                                        1, mean), dsb_sd = apply(adt_log, 1, sd)))
    if (isTRUE(denoise.counts)) {
      if (isTRUE(use.isotype.control)) {
        technical_stats = cbind(t(noise_matrix), dsb_technical_component = noise_vector)
      }
      else {
        technical_stats = data.frame(dsb_technical_component = noise_vector)
      }
    }
    else {
      technical_stats = NULL
    }
    if (is.null(empty_drop_matrix)) {
      background.stats = NULL
    }
    else {
      background.stats = data.frame(background_mean = mu_u, 
                                    background_sd = sd_u)
      protein_stats = c(protein_stats, background.stats)
    }
    ret_obj = list(dsb_normalized_matrix = norm_adt, technical_stats = technical_stats, 
                   protein_stats = protein_stats)
    return(ret_obj)
  }
  else {
    return(norm_adt)
  }
}

```

Before reading a Seurat v5 .Rds *in-memory*, a folder containing the **object .RDS** & **BPCells matrices** must be downloaded *on-disk* and accessible in your working directory.

*In-memory*, Seurat accesses the *on-disk* matricies by the pre-defined "destination directory" path (pointing to the BPCells directories) and stores them in the assay level data.

```{r load_data}

orig_data <- SeuratObject::LoadSeuratRds(file = "seurat_2024-04-03/aml_atlas_layered_seurat.rds") 

dim(orig_data)
orig_data@assays

```

# QC

In Seurat v5, objects are either "layered", with raw counts for each dataset stored in individual matrices, or "joined", with all datasets collapsed and stored in one matrix. Seurat requires layers to be joined before downstream analysis.

Before joining the layers, we can perform pre-processing steps on the layered dataset to process each sample independently and use the selected features for downstream analysis. Because we can't filter features in a layered object, the selected features have to be filtered before joining.

## Cell counts across samples prior to filtering

```{r}
 
# total cells by sample_id
all_total <- table(orig_data$sample_id) %>%
  as.data.frame() %>% 
  dplyr::rename("Donor" = "Var1", "total_number_of_cells" = "Freq") %>%
  dplyr::arrange(desc(total_number_of_cells))

DT::datatable(all_total, rownames = FALSE)

hist(all_total$total_number_of_cells, breaks = 20)

```

## Library size & complexity distributions
**RNA**

```{r, fig.height=5, fig.width=10}

nfeat_bottom_cutoff <- 200
nfeat_top_cutoff <- 7000

ncount_top_cutoff <- 30000
ncount_bottom_cutoff <- 1000

FeatureScatter(orig_data, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size = 0.5) + 
  geom_hline(yintercept = nfeat_bottom_cutoff, lty = 2) +
  geom_hline(yintercept = nfeat_top_cutoff, lty = 2) +
  geom_vline(xintercept = ncount_bottom_cutoff, lty = 2) +
  geom_vline(xintercept = ncount_top_cutoff, lty = 2)

hist(orig_data$nFeature_RNA, main = "Complexity distribution", breaks = 50)
abline(v = nfeat_bottom_cutoff, lty = 2)

hist(orig_data$nCount_RNA, main = "Library size distribution",  breaks = 50)
abline(v = ncount_top_cutoff, lty = 2)

```

**ADT**

```{r, fig.height = 10, fig.width = 15}

DefaultAssay(orig_data) <- "ADT"
Idents(orig_data) <- "sample_id"

# Visualizing library size & complexity
VlnPlot(orig_data, features = c("nFeature_ADT", "nCount_ADT"), ncol = 1, pt.size = 0)
RidgePlot(orig_data, features = "nCount_ADT", group.by = "sample_id", log = TRUE) + NoLegend()
FeatureScatter(orig_data, feature1 = "nCount_ADT", feature2 = "nFeature_ADT") + NoLegend()
hist(orig_data$nCount_ADT, breaks = 100)

```

## Filtering

**nCount_RNA, nFeature_RNA, and nCount_ADT**

```{r}

# RNA
DefaultAssay(orig_data) <- "RNA"
# orig_data <- dplyr::filter(orig_data, 
#                            nCount_RNA < ncount_top_cutoff, 
#                            nCount_RNA > ncount_bottom_cutoff,
#                            nFeature_RNA > nfeat_bottom_cutoff,
#                            nFeature_RNA < nfeat_top_cutoff)

orig_data <- dplyr::filter(orig_data, 
                           nCount_RNA > ncount_bottom_cutoff,
                           nFeature_RNA > nfeat_bottom_cutoff)

# ADT
DefaultAssay(orig_data) <- "ADT"
#orig_data <- dplyr::filter(orig_data, nCount_ADT < 100000)

```

**Summary of mitochondrial composition**

Calculating percentage of cells that pass \<10% mitochondrial composition filter for each sample

```{r mito_analysis}

mito_filt_thr <- 15

mito_summary <- orig_data@meta.data %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(perc_cells_pass_mito_filter = signif(sum(percent.mt < mito_filt_thr) / n() * 100, 2)) %>%
  dplyr::ungroup() %>%
  dplyr::select(sample_id, perc_cells_pass_mito_filter) %>%
  dplyr::distinct()

DT::datatable(mito_summary, rownames = FALSE)

orig_data <- dplyr::filter(orig_data, percent.mt <= mito_filt_thr)

```


## Integrate data

Seurat's CLR normalization will not work on IterableMatrix object, it's converted to sparse matrix assay (v4 class assay) and processed as before.

### Healthy
```{r integrate_data_healthy, eval = TRUE}

healthy_data <- dplyr::filter(orig_data, time %in% "Healthy")

healthy_data <- dplyr::mutate(healthy_data, 
                              experiment_batch = gsub("vorbio-analysis/single_cell/internal/", "", key),
                              experiment_batch = gsub("\\/.*", "", experiment_batch))

tmp <- JoinLayers(healthy_data, assay = "RNA")
feat_sums <- rowSums(tmp@assays$RNA@layers$counts)
healthy_data <- subset(healthy_data, features = c(rownames(healthy_data@assays$ADT), rownames(tmp@assays$RNA)[feat_sums > 0]))
rm(tmp)

# Normalize, scale, and find variable features 
healthy_data <- NormalizeData(healthy_data, normalization.method = "LogNormalize", assay = "RNA")
healthy_data <- FindVariableFeatures(healthy_data, assay = "RNA", nfeatures = 2500)
healthy_data <- ScaleData(healthy_data, assay = "RNA") 
healthy_data <- RunPCA(healthy_data, npcs = 50, assay = "RNA", seed.use = 905217) 
healthy_data2 <- healthy_data

ElbowPlot(healthy_data, ndims = 50)

set.seed(87564)
healthy_data <- IntegrateLayers(
  object = healthy_data, 
  assay = "RNA",
  method = FastMNNIntegration,
  layers = "data",
  auto.merge = TRUE,
  verbose = TRUE,
  features = VariableFeatures(healthy_data@assays$RNA)
)

healthy_data2 <- IntegrateLayers(
  object = healthy_data2,
  assay = "RNA",
  group.by = "sample_id",
  method = HarmonyIntegration,
  orig.reduction = "pca",
  layers = "data",
  verbose = TRUE
)

healthy_data@reductions$harmony <- healthy_data2@reductions$harmony
rm(healthy_data2)

# healthy_data <- JoinLayers(healthy_data, assay = "RNA")
# healthy_data_split <- SplitObject(healthy_data, split.by = "sample_id")
# anchors <- FindIntegrationAnchors(healthy_data_split, assay = rep("RNA", length(healthy_data_split)), reduction = "cca", anchor.features = 500) 
# healthy_data <- IntegrateData(anchorset = anchors)

healthy_data <- FindNeighbors(healthy_data, 
                              reduction = "integrated.mnn", 
                              k.param = 15,
                              graph.name = c("nn_mnn", "snn_mnn"), 
                              dims = 1:50)

healthy_data <- FindNeighbors(healthy_data, 
                              reduction = "harmony", 
                              k.param = 15,
                              graph.name = c("nn_harmony", "snn_harmony"), 
                              dims = 1:50)

healthy_data <- FindClusters(healthy_data, 
                             graph.name = "snn_mnn", 
                             cluster.name = "seurat_clusters_mnn", 
                             resolution = 1.5, 
                             method = "igraph", 
                             random.seed = 14692)

healthy_data <- FindClusters(healthy_data, 
                             graph.name = "snn_harmony", 
                             cluster.name = "seurat_clusters_harmony", 
                             resolution = 1.5, 
                             method = "igraph", 
                             random.seed = 1291)

healthy_data <- RunUMAP(healthy_data, 
                        reduction = "integrated.mnn", 
                        dims = 1:50, 
                        reduction.key = "UMAP_mnn_", 
                        reduction.name = "umap_mnn", 
                        repulsion.strength = 3, 
                        local.connectivity = 5,
                        n.neighbors = 30,
                        spread = 4,
                        return.model = TRUE,
                        seed.use = 259814)

healthy_data <- RunUMAP(healthy_data, 
                        reduction = "harmony", 
                        dims = 1:50, 
                        reduction.key = "UMAP_harmony_", 
                        reduction.name = "umap_harmony", 
                        repulsion.strength = 1, 
                        local.connectivity = 2,
                        n.neighbors = 30,
                        spread = 5,
                        seed.use = 286)

# Join ADT layers
healthy_data <- JoinLayers(healthy_data, assay = "ADT")

DefaultAssay(healthy_data) <- "ADT"

# convert ADT counts to sparse matrix 
healthy_data[["ADT"]] <- as(object = healthy_data[["ADT"]], Class = "Assay")

# ignore variation from isotype controls
adt_feats <- rownames(healthy_data@assays$ADT)
adt_feats <- adt_feats[!grepl("isotype", adt_feats, ignore.case = TRUE)]

healthy_data <- healthy_data %>%
  NormalizeData(normalization.method = "CLR", assay = "ADT", margin = 2) %>%
  ScaleData(assay = "ADT", features = adt_feats, vars.to.regress = "sample_id") %>%
  RunPCA(assay = "ADT", reduction.name = "pca_adt", features = adt_feats, seed.use = 65187) %>%
  FindNeighbors(reduction = "pca_adt", graph.name = "snn_adt") %>%
  FindClusters(graph.name = "snn_adt", cluster.name = "seurat_clusters_adt", random.seed = 11293) %>%
  RunUMAP(reduction = "pca_adt", reduction.name = "umap_adt", seed.use = 99698, dims = 1:10)

# healthy_data[["ADT"]] <- split(healthy_data[["ADT"]], f = healthy_data$sample_id)
# 
# healthy_data <- IntegrateLayers(
#   object = healthy_data, 
#   assay = "ADT",
#   features = adt_feats,
#   method = FastMNNIntegration,
#   layers = "data",
#   new.reduction = "integrated.mnn_adt",
#   reduction.key = "mnn_adt_",
#   reconstructed.assay = "mnn.reconstructed_adt",
#   auto.merge = TRUE,
#   verbose = TRUE
# )

healthy_data <- FindMultiModalNeighbors(
  healthy_data, 
  reduction.list = list("integrated.mnn", "pca_adt"), 
  dims.list = list(1:50, 1:5), 
  modality.weight.name = "RNA.weight"
)

healthy_data <- RunUMAP(healthy_data, 
                        nn.name = "weighted.nn", 
                        reduction.name = "umap_wnn", 
                        reduction.key = "wnnUMAP_",
                        repulsion.strength = 2, 
                        local.connectivity = 1,
                        n.neighbors = 20,
                        spread = 5,
                        seed.use = 8357)



DotPlot(healthy_data, features = c("CD3", "CD19", "CD14", "CD45", "CD34"), group.by = "seurat_clusters_mnn", assay = "ADT")

FeaturePlot(healthy_data, features = c("CD34", "CD33", "CD3", "CD19"), reduction = "umap_mnn")

feature_cluster <- hclust(dist(healthy_data@assays$ADT@scale.data))

DoHeatmap(healthy_data, 
          features = rownames(healthy_data@assays$ADT@scale.data)[feature_cluster$order], 
          assay = "ADT", 
          group.by = "seurat_clusters_mnn", 
          cells = sample(healthy_data$barcode, 1000, replace = FALSE))

Idents(healthy_data) <- "seurat_clusters_mnn"
cluster_markers <- FindAllMarkers(healthy_data, 
                                  assay = "ADT", 
                                  logfc.threshold = log2(1.1), 
                                  random.seed = 3611, 
                                  only.pos = FALSE, 
                                  test.use = "negbinom", 
                                  latent.vars = "sample_id") %>%
  dplyr::arrange(cluster, desc(avg_log2FC)) %>%
  dplyr::group_split(cluster)

mymarkers <- do.call("rbind", cluster_markers) %>%
  dplyr::pull(gene) %>%
  unique()
mymarkers <- mymarkers[!grepl("ISOTYPE", mymarkers)]

feature_cluster <- hclust(dist(healthy_data@assays$ADT@scale.data[mymarkers, ]))
cell_sample <- unlist(lapply(unique(healthy_data$cell_state_mnn_rna_l1), function(arg) sample(colnames(healthy_data)[healthy_data$cell_state_mnn_rna_l1 %in% arg], 100)))
DoHeatmap(healthy_data, 
          features = rownames(healthy_data@assays$ADT@scale.data[mymarkers, ])[feature_cluster$order], 
          assay = "ADT", 
          group.by = "seurat_clusters_mnn", 
          cells = cell_sample)


cluster_markers_rna <- FindAllMarkers(JoinLayers(healthy_data, assay = "RNA"), 
                                      assay = "RNA", 
                                      logfc.threshold = log2(1.1), 
                                      latent.vars = "sample_id",
                                      random.seed = 4860, 
                                      only.pos = TRUE) %>%
  dplyr::arrange(cluster, desc(avg_log2FC)) %>%
  dplyr::group_split(cluster)

# annotate cells
# naive_bcell <- c(12, 16, 20, 32)
# memory_bcell <- c(36)
# plasma_cell <- c(30)
# bcell_precursor <- c(15, 27)
# cd4_naive_tcells <- c(0)
# cd4_memory_tcells <- c(1, 2)
# cd8_naive_tcells <- c(3)
# cd8_memory_tcells <- c(5)
# monocytes <- c(7, 8)
# granulocyte <- c(25, 28)
# cdc <- c(21)
# pdc <- c(29)
# nk <- c(6, 13)
# nkt <- c(4)
# neutrophil <- c(14)
# erythroblast <- c(10, 26)
# erythrocyte <- c(9, 19, 22, 23, 31)
# hsc <- c(11)
# gmp <- c(24)
# mep <- c(18)


bcell <- c(12, 15, 19, 24, 30)
bcell_precursor <- c(16)
cd4_naive_tcells <- c(0)
cd4_memory_tcells <- c(1, 2)
cd8_naive_tcells <- c(4)
cd8_memory_tcells <- c(5)
monocytes <- c(7, 8, 21, 28)
cdc <- c(22)
pdc <- c(29)
nk <- c(6, 13)
nkt <- c(3)
erythroblast <- c(10, 20, 27)
erythroid <- c(9, 23, 25)
megakaryocyte_precursor <- c(18)
hsc <- c(11)
gmp <- c(26)
mep <- c(17)
unknown <- c(14, 31)

healthy_data <- healthy_data %>%
  dplyr::mutate(cell_state_mnn_rna_l1 = case_when(seurat_clusters_mnn %in% hsc ~ "HSC",
                                                      seurat_clusters_mnn %in% mep ~ "MEP",
                                                      seurat_clusters_mnn %in% gmp ~ "GMP",
                                                      seurat_clusters_mnn %in% bcell ~ "B cell",
                                                      seurat_clusters_mnn %in% bcell_precursor ~ "Pre/pro B cell",
                                                      seurat_clusters_mnn %in% cd4_naive_tcells ~ "CD4+ naive T",
                                                      seurat_clusters_mnn %in% cd4_memory_tcells ~ "CD4+ memory T",
                                                      seurat_clusters_mnn %in% cd8_naive_tcells ~ "CD8+ naive T",
                                                      seurat_clusters_mnn %in% cd8_memory_tcells ~ "CD8+ memory T",
                                                      seurat_clusters_mnn %in% erythroblast ~ "Erythroblast",
                                                      seurat_clusters_mnn %in% erythroid ~ "Erythroid",
                                                      seurat_clusters_mnn %in% megakaryocyte_precursor ~ "MkP",
                                                      seurat_clusters_mnn %in% monocytes ~ "Monocyte",
                                                      seurat_clusters_mnn %in% cdc ~ "cDC",
                                                      seurat_clusters_mnn %in% pdc ~ "pDC",
                                                      seurat_clusters_mnn %in% nk ~ "NK",
                                                      seurat_clusters_mnn %in% nkt ~ "NKT",
                                                      seurat_clusters_mnn %in% unknown ~ "Unknown")) %>%
  dplyr::mutate(cell_state_mnn_rna_l1 = factor(cell_state_mnn_rna_l1, 
                                                   levels = c("HSC",
                                                              "MEP",
                                                              "GMP",
                                                              "Monocyte",
                                                              "cDC",
                                                              "pDC",
                                                              "NK",
                                                              "B cell",
                                                              "Pre/pro B cell",
                                                              "CD4+ naive T",
                                                              "CD4+ memory T",
                                                              "CD8+ naive T",
                                                              "CD8+ memory T",
                                                              "NKT",
                                                              "Erythroblast",
                                                              "Erythroid",
                                                              "MkP",
                                                              "Unknown")))


# identify subclusters
healthy_data_split <- SplitObject(healthy_data, split.by = "cell_state_mnn_rna_l1")

cluster_list <- NULL
for (i in 1:length(healthy_data_split)) {
  
  cat(i, "\n")
  arg <- healthy_data_split[[i]]
  arg <- NormalizeData(arg)
  arg <- FindVariableFeatures(arg)
  arg <- ScaleData(arg, vars.to.regress = "sample_id")
  arg <- RunPCA(arg)
  
  # arg <- IntegrateLayers(
  #   object = arg, 
  #   assay = "RNA",
  #   method = FastMNNIntegration,
  #   orig.reduction = "pca", 
  #   d = 30,
  #   auto.merge = TRUE,
  #   verbose = TRUE
  # )
  
  arg <- FindNeighbors(arg, 
                       reduction = "pca", 
                       k.param = 15,
                       n.trees = 100,
                       dims = 1:30)
  
  arg <- FindClusters(arg, 
                      cluster.name = "seurat_subclusters", 
                      resolution = 0.5, 
                      method = "igraph", 
                      random.seed = 1291)

  arg <- dplyr::select(arg, barcode, seurat_subclusters)
    
  cluster_list[[names(healthy_data_split)[i]]] <- arg
  
}

subclusters <- do.call("rbind", cluster_list)
stopifnot(healthy_data$barcode %in% subclusters$barcode)
healthy_data <- healthy_data %>% 
  dplyr::left_join(subclusters, by = "barcode") %>%
  dplyr::mutate(seurat_subclusters = paste(seurat_clusters_mnn, seurat_subclusters, sep = "-"))

mycols <- c("firebrick1", "lightcyan3", "seagreen3", "cyan", "purple", "hotpink", "gold1", "mediumturquoise", "yellow2", "sienna3", "springgreen", "olivedrab", "orange1",  "cornflowerblue", "rosybrown1", "maroon2", "slateblue2", "tan2", "peachpuff1", "orchid1", "yellowgreen", "burlywood2")

DimPlot(dplyr::filter(healthy_data, cell_state_mnn_rna_l1 != "Unknown"), 
        group.by = "cell_state_mnn_rna_l1", 
        reduction = "umap_mnn", 
        cols = as.character(mycols), 
        label = TRUE, 
        repel = TRUE,
        label.size = 5,
        alpha = 0.7) +
  NoAxes() +
  NoLegend() 


DimPlot(healthy_data, group.by = "seurat_clusters_mnn", label = TRUE)

healthy_data_subset <- dplyr::filter(healthy_data, cell_state_mnn_rna_l1 != "Unknown")
healthy_data_subset <- lapply(unique(healthy_data_subset$cell_state_mnn_rna_l1), function(arg) {
  sample(colnames(dplyr::filter(healthy_data_subset, cell_state_mnn_rna_l1 == arg)), 100)
}) %>% unlist()
healthy_data_subset <- subset(healthy_data, cells = healthy_data_subset) %>%
  dplyr::mutate(cell_state_mnn_rna_l1 = factor(cell_state_mnn_rna_l1))

rownames(healthy_data_subset@assays$ADT@scale.data) <- case_when(rownames(healthy_data_subset@assays$ADT@scale.data) == "CLL1" ~ "CLL-1",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "EMR2" ~ "ADGRE2",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD205" ~ "DEC-205",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD49D" ~ "ITGA4",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD305" ~ "LAIR1",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD85K" ~ "LILRB4",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD88" ~ "C5AR1",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD120B" ~ "TNFR2",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD135" ~ "FLT3",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD152" ~ "CTLA-4",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD200" ~ "OX2",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD266" ~ "DNAM-1",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD235A" ~ "GlyA",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD273" ~ "PD-L2",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD274" ~ "PD-L1",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD279" ~ "PD-1",
                                                          rownames(healthy_data_subset@assays$ADT@scale.data) == "CD366" ~ "Tim-3",
                                                          TRUE ~ as.character(rownames(healthy_data_subset@assays$ADT@scale.data)))

adt_feats <- rownames(healthy_data_subset@assays$ADT@scale.data)
myclust <- hclust(dist(healthy_data_subset@assays$ADT@scale.data), method = "ward.D2")
adt_feats <- adt_feats[myclust$order]

DoHeatmap(healthy_data_subset, 
          group.by = "cell_state_mnn_rna_l1", 
          assay = "ADT", 
          features = adt_feats, 
          group.colors = mycols,
          size = 4,
          raster = FALSE)


rna_de_feats <- FindAllMarkers(JoinLayers(healthy_data, assay = "RNA"),
                               group.by = "cell_state_mnn_rna_l1",
                               assay = "RNA",
                               min.pct = 0.1,
                               logfc.threshold = 1)

rna_feats <- rna_de_feats %>%
  dplyr::filter(p_val_adj < 0.00001 | is.na(p_val_adj)) %>%
  dplyr::group_by(cluster) %>%
  dplyr::arrange(p_val) %>%
  dplyr::slice(1:5) %>%
  dplyr::ungroup()

rna_feats <- dplyr::bind_rows(rna_feats,
                              data.frame(cluster = "HSC", gene = c("AVP", "MECOM", "HLF")),
                              data.frame(cluster = "NKT", gene = c("ADGRG1", "TRGV5", "TRDC", "TRGC1", "TRAV10", "TRAV25")),
                              data.frame(cluster = "Erythroid", gene = "GYPA"),
                              data.frame(cluster = "pDC", gene = "IL3RA"),
                              data.frame(cluster = "CD4+ naive T", gene = "CD4"),
                              data.frame(cluster = "B cell", gene = "CD19"),
                              data.frame(cluster = "MkP", gene = c("PF4", "GATA1"))) %>%
  dplyr::mutate(cluster = factor(cluster, levels = levels(healthy_data$cell_state_mnn_rna_l1))) %>%
  dplyr::arrange(cluster)

rna_feats <- rna_feats %>%
  dplyr::pull(gene) %>%
  unique()

healthy_data_subset <- ScaleData(healthy_data_subset, features = rna_feats, assay = "RNA")

# myclust <- hclust(dist(LayerData(healthy_data_subset, features = rna_feats, layer = "scale.data", assay = "RNA")))
# rna_feats <- rna_feats[myclust$order]

DoHeatmap(healthy_data_subset, 
          group.by = "cell_state_mnn_rna_l1", 
          assay = "RNA", 
          features = rna_feats, 
          group.colors = mycols,
          size = 4,
          raster = FALSE)









FeaturePlot(healthy_data, features = c("rna_AVP"), reduction = "umap_mnn")
VlnPlot(healthy_data, features = c("TYMS"),  group.by = "seurat_clusters_harmony")
DotPlot(healthy_data, features = c("TYMS", "C1QTNF4", "PIK3R1", "GYPC", "IGLL1", "STMN1", "TUBB", "HNRNPA1", "SLC40A1", "RNF130", "HIST1H4C"),  group.by = "seurat_clusters_harmony", assay = "RNA") + RotatedAxis()

healthy_adj_matrix <- healthy_data@graphs$nn_harmony

```



### AML
#### RNA
##### Harmony
```{r integrate_data_aml}

paired_samples <- c("4", "26", "27", "28", "29", "30", "31", "40", "41", "42", "43", "44", "47", "48", "49", "50", "51", "52", "54", "56", "58", "60", "61", "64", "65", "66", "67", "70")
paired_samples <- paste0("Patient_", paired_samples)

aml_data <- dplyr::filter(orig_data, patient %in% paired_samples)
rm(orig_data)

# Normalize, scale, and find variable features 
aml_data <- FindVariableFeatures(aml_data, assay = "RNA", nfeatures = 2000)
aml_data <- NormalizeData(aml_data, normalization.method = "LogNormalize", assay = "RNA", margin = 2)
aml_data <- ScaleData(aml_data, assay = "RNA") 
aml_data <- RunPCA(aml_data, assay = "RNA", seed.use = 466) 
aml_data <- RunUMAP(aml_data, dims = 1:25, assay = "RNA", seed.use = 59111)

output_path <- "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/seurat_2024-04-03/aml_data_paired_samples_layered.rds"

if (!file.exists(output_path)) {
  saveRDS(aml_data, output_path)
}
  
set.seed(912)
aml_data <- IntegrateLayers(
  object = aml_data, 
  assay = "RNA",
  group.by = "sample_id",
  method = HarmonyIntegration, #other methods don't work with Seurat v5 beta
  orig.reduction = "pca", 
  verbose = TRUE
)

aml_data <- FindNeighbors(aml_data, 
                          reduction = "harmony", 
                          graph.name = c("nn_harmony", "snn_harmony"), 
                          dims = 1:35)

aml_data <- FindClusters(aml_data, 
                         graph.name = "snn_harmony", 
                         cluster.name = "seurat_clusters_harmony", 
                         resolution = 0.9, 
                         method = "igraph", 
                         random.seed = 871)

aml_data <- RunUMAP(aml_data, 
                    reduction = "harmony", 
                    dims = 1:35, 
                    reduction.key = "UMAP_harmony_", 
                    reduction.name = "umap_harmony", 
                    repulsion.strength = 2, 
                    n.neighbors = 15,
                    seed.use = 141)

aml_data <- RunUMAP(aml_data, 
                    reduction = "harmony", 
                    dims = 1:35, 
                    reduction.key = "UMAP_harmony_3d_", 
                    reduction.name = "umap_harmony_3d", 
                    repulsion.strength = 2, 
                    n.neighbors = 15,
                    n.components = 3,
                    seed.use = 141)

```

##### FastMNN
```{r integrate_data_aml}

set.seed(912)
aml_data_tmp <- IntegrateLayers(
  object = aml_data_tmp, 
  assay = "RNA",
  #group.by = "sample_id",
  method = FastMNNIntegration, #other methods don't work with Seurat v5 beta
  orig.reduction = "pca", 
  verbose = TRUE
)

aml_data_tmp <- FindNeighbors(aml_data_tmp, 
                              reduction = "integrated.mnn", 
                              graph.name = c("nn_mnn", "snn_mnn"), 
                              dims = 1:35)

aml_data_tmp <- FindClusters(aml_data_tmp, 
                             graph.name = "snn_mnn", 
                             cluster.name = "seurat_clusters_mnn", 
                             resolution = 0.9, 
                             method = "igraph", 
                             random.seed = 871)

aml_data_tmp <- RunUMAP(aml_data_tmp, 
                        reduction = "integrated.mnn", 
                        dims = 1:35, 
                        reduction.key = "UMAP_mnn_", 
                        reduction.name = "umap_mnn", 
                        repulsion.strength = 1, 
                        n.neighbors = 15,
                        seed.use = 141)

aml_data@reductions$mnn <- aml_data_tmp@reductions$integrated.mnn
aml_data@reductions$umap_mnn <- aml_data_tmp@reductions$umap_mnn
aml_data@assays$RNA_mnn_reconstructed <- aml_data_tmp@assays$mnn.reconstructed
rm(aml_data_tmp)
```

#### ADT
```{r analyze_adt}

# temporary fix; perform formal imputation of missing ADT features later **** these samples have different number of antigens in panel
pat47and48 <- dplyr::filter(aml_data, patient %in% c("Patient_47", "Patient_48"))
aml_data <- dplyr::filter(aml_data, !patient %in% c("Patient_47", "Patient_48"))
# remove cells with no ADT signal 
aml_data <- dplyr::filter(aml_data, nCount_ADT > 0)
# Join ADT layers
aml_data <- JoinLayers(aml_data, assay = "ADT")

DefaultAssay(aml_data) <- "ADT"
# convert ADT counts to sparse matrix
aml_data[["ADT"]] <- as(object = aml_data[["ADT"]], Class = "Assay")

# ignore variation from isotype controls
adt_feats <- rownames(aml_data@assays$ADT)
adt_feats <- adt_feats[!grepl("isotype", adt_feats, ignore.case = TRUE)]

aml_data <- NormalizeData(aml_data, normalization.method = "CLR", assay = "ADT", margin = 2)
aml_data <- ScaleData(aml_data, assay = "ADT", features = adt_feats, vars.to.regress = "sample_id")
aml_data <- RunPCA(aml_data, assay = "ADT", reduction.name = "pca_adt", features = adt_feats, seed.use = 6587)
aml_data <- FindNeighbors(aml_data, reduction = "pca_adt", graph.name = c("nn_adt", "snn_adt"))
aml_data <- FindClusters(aml_data, graph.name = "snn_adt", cluster.name = "seurat_clusters_adt", random.seed = 1193)
aml_data <- RunUMAP(aml_data, reduction = "pca_adt", reduction.name = "umap_adt", seed.use = 698, dims = 1:10)

# background correction using limma to regress out signal from isotype controls
isotype_controls <- rownames(aml_data@assays$ADT)
isotype_controls <- isotype_controls[grepl("isotype", isotype_controls, ignore.case = TRUE)]
model_data <- aml_data@assays$ADT@data
rownames(model_data) <- rownames(aml_data@assays$ADT)
colnames(model_data) <- colnames(aml_data@assays$ADT)

mod_res <- limma::removeBatchEffect(x = model_data[!rownames(model_data) %in% isotype_controls, ], 
                                    covariates = t(model_data[rownames(model_data) %in% isotype_controls, ]))
aml_data[["ADT_limma"]] <- Seurat::CreateAssayObject(data = mod_res)

# background correction using DSB
sample_paths <- unique(aml_data$key)
raw_sample_paths <- gsub("filtered", "raw", sample_paths)
raw_sample_paths <- paste0("s3://", raw_sample_paths)

bg_adt_matrix <- list()
bg_rna_matrix <- list()
for (id in 1:length(raw_sample_paths)) {
  
  cat(raw_sample_paths[id], "\n")
  raw_adt_mat <- aws.s3::s3read_using(FUN = Read10X_h5, object = raw_sample_paths[id])[["Antibody Capture"]]
  raw_rna_mat <- aws.s3::s3read_using(FUN = Read10X_h5, object = raw_sample_paths[id])[["Gene Expression"]]
  tmp_id <- unique(aml_data@meta.data$sample_id[aml_data@meta.data$key %in% sample_paths[id]])
  colnames(raw_adt_mat) <- paste(colnames(raw_adt_mat), tmp_id, sep = "-")
  colnames(raw_rna_mat) <- paste(colnames(raw_rna_mat), tmp_id, sep = "-")
  bg_adt_matrix[[tmp_id]] <- raw_adt_mat[, colSums(raw_adt_mat) > 0]
  bg_rna_matrix[[tmp_id]] <- Matrix::colSums(raw_rna_mat)[Matrix::colSums(raw_rna_mat) > 0]

}


stopifnot(all(sapply(bg_adt_matrix, nrow) == nrow(aml_data@assays$ADT)))
name_list <- lapply(bg_adt_matrix, rownames)
name_list <- sapply(name_list, function(arg) all(rownames(arg) == rownames(name_list[[1]])))
stopifnot(all(name_list))

bg_adt_matrix <- do.call("cbind", bg_adt_matrix)

rownames(bg_adt_matrix) <- toupper(gsub("\\..*", "", rownames(bg_adt_matrix)))
rownames(bg_adt_matrix) <- toupper(gsub("_", "-", rownames(bg_adt_matrix)))
rownames(bg_adt_matrix) <- ifelse(rownames(bg_adt_matrix) == "CD371", "CLL1", rownames(bg_adt_matrix))
stopifnot(all(rownames(bg_adt_matrix) == rownames(aml_data@assays$ADT)))

# select droplets based on RNA and ADT library size
adt_size <- Matrix::colSums(bg_adt_matrix)
adt_complex <- Matrix::colSums(bg_adt_matrix > 0)
rna_size <- unlist(bg_rna_matrix)
names(rna_size) <- gsub(".*\\.", "", names(rna_size))
overlap_bc <- intersect(names(adt_size), names(rna_size))
adt_size <- adt_size[names(adt_size) %in% overlap_bc]
adt_complex <- adt_complex[names(adt_complex) %in% overlap_bc]
rna_size <- rna_size[names(rna_size) %in% overlap_bc]
stopifnot(all(names(adt_size) == names(rna_size)))

qc_df <- data.frame(rna_size = rna_size,
                    adt_size = adt_size) %>%
  dplyr::filter(rna_size > 0 & adt_size > 0)

myclust <- mclust::Mclust(data = qc_df, G = 3)
qc_df <- dplyr::mutate(qc_df, 
                       cluster = as.factor(myclust$classification),
                       uncertainty = myclust$uncertainty)
mysample <- qc_df[sample(1:nrow(qc_df), 5000), ]

ggscatter(data = mysample,
          title = "Random sample of 5,000 empty droplets",
          x = "adt_size",
          y = "rna_size",
          xlab = "ADT library size",
          ylab = "mRNA library size",
          color = "cluster",
          alpha = 0.1) +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 10000)) +
  scale_y_log10() +
  stat_density2d()

# keep only empty droplets for background matrix
bg_adt_matrix <- bg_adt_matrix[, !colnames(bg_adt_matrix) %in% colnames(aml_data)]
# remove droplets with zero signal
bg_adt_matrix <- bg_adt_matrix[, colSums(bg_adt_matrix) > 50]
# remove droplets near filter threshold based on histogram of droplet count sums
bg_adt_matrix <- bg_adt_matrix[, colSums(bg_adt_matrix) < 1000]
# remove droplets that have large library sizes but low complexity indicating just a few features are saturating the background signal
#bg_adt_matrix <- bg_adt_matrix[, colSums(bg_adt_matrix > 0) >= 50]
# remove droplets that have relatively high RNA expression (caution: there are less cells with RNA size info)
rna_filter <- names(rna_size)[rna_size < 100]
bg_adt_matrix <- bg_adt_matrix[, colnames(bg_adt_matrix) %in% rna_filter]


# plot mean and variance trend for each feature
fmean <- rowMeans(log1p(bg_adt_matrix))
fsd <- apply(log1p(bg_adt_matrix), 1, sd)
full_droplet_means <- rowMeans(log1p(aml_data@assays$ADT@counts))
full_droplet_means <- full_droplet_means[names(fmean)]
meanvar_df <- data.frame(feature_mean = fmean,
                         full_droplet_mean = full_droplet_means,
                         feature_sd = fsd,
                         feature_var = fsd^2) %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::mutate(label = ifelse(gene %in% c("CD33", "CLL1", "CD123", "EMR2", "CD45", "CD99", "CD44"), gene, NA),
                coef_of_variation = feature_sd / feature_mean * 100,
                corrected_value = (full_droplet_mean - feature_mean) / feature_sd) %>%
  dplyr::arrange(desc(coef_of_variation))

p <- ggscatter(data = meanvar_df,
               title = "Mean-variance trend in log-space",
               x = "feature_mean",
               y = "feature_sd",
               xlab = "Antigen expression mean",
               ylab = "Antigen expression SD",
               add = "loess",
               add.params = list(color = "red", size = 0.5))  # note: no label or repel argument

# Now add customized labels:
p + geom_text_repel(aes(label = label),
                    box.padding = 1,    # extra padding around the label box
                    point.padding = 0.1,  # extra space around the data point
                    segment.size = 0.5, # thickness of the connecting line
                    segment.color = "grey50",
                    nudge_y = 0.5)    # nudge labels vertically (adjust as needed)

ggbarplot(meanvar_df,
          x = "gene",
          y = "coef_of_variation",
          x.text.angle = 45,
          fill = "label")

ggscatter(meanvar_df,
          x = "full_droplet_mean",
          y = "feature_mean",
          xlab = "Mean counts in full droplets",
          ylab = "Mean counts in empty droplets",
          add = "loess",
          add.params = list(color = "red", size = 0.5)) + 
  geom_text_repel(aes(label = label),
  box.padding = 1,    # extra padding around the label box
  point.padding = 0.1,  # extra space around the data point
  segment.size = 0.5, # thickness of the connecting line
  segment.color = "grey50",
  nudge_y = 0.5) +
  scale_x_log10() +
  scale_y_log10()

# DSB normalization using aggregated background droplet matrix
cell_adt_matrix <- aml_data@assays$ADT@counts
# normalize and de-noise with dsb
stopifnot(isotype_controls %in% rownames(cell_adt_matrix))
stopifnot(isotype_controls %in% rownames(bg_adt_matrix))
set.seed(35600)
cell_sample <- sample(1:ncol(cell_adt_matrix), 2000)
dsb_norm_mat <- dsb::DSBNormalizeProtein(
  
  cell_protein_matrix = cell_adt_matrix, 
  empty_drop_matrix = bg_adt_matrix, 
  denoise.counts = TRUE, 
  use.isotype.control = TRUE, 
  scale.factor = "standardize",
  isotype.control.name.vec = isotype_controls
  
)

aml_data[["ADT_dsb_qc_filter"]] <- CreateAssayObject(data = dsb_norm_mat$dsb_normalized_matrix)

aml_data_ebayes_list <- NULL
set.seed(7350)
for (samp in unique(aml_data$sample_id)) {
  
  barcodes <- colnames(bg_adt_matrix)
  barcodes <- barcodes[gsub(".*-", "", barcodes) %in% samp]
  
  qc_df_sample <- qc_df[grepl(samp, rownames(qc_df)), colnames(qc_df) %in% c("rna_size", "adt_size")]
  myclust <- mclust::Mclust(data = qc_df_sample, G = 3)
  
  qc_df_sample <- qc_df_sample %>%
    dplyr::mutate(cluster = myclust$classification,
                  uncertainty = myclust$uncertainty)
  
  bg_cluster <- which(myclust$parameters$mean["adt_size", ] == median(myclust$parameters$mean["adt_size", ]))
  top_empty_droplets <- qc_df_sample[rownames(qc_df_sample) %in% barcodes & qc_df_sample$cluster == bg_cluster, ]
  top_empty_droplets <- top_empty_droplets %>%
    dplyr::arrange(uncertainty) %>%
    dplyr::slice(1:1000)
  top_empty_droplets <- rownames(top_empty_droplets)
  
  bg <- bg_adt_matrix[, colnames(bg_adt_matrix) %in% top_empty_droplets]
  
  aml_data_ebayes <- dsb_ebayes(
    
    cell_protein_matrix = dplyr::filter(aml_data, sample_id %in% samp)@assays$ADT@counts, 
    empty_drop_matrix = bg, 
    denoise.counts = TRUE, 
    scale.factor = "standardize",
    isotype.control.name.vec = isotype_controls,
    use.isotype.control = TRUE,
    return.stats = TRUE
    
  )
  
  aml_data_ebayes_list[[samp]] <- aml_data_ebayes

}

aml_data_ebayes_list <- lapply(aml_data_ebayes_list, function(arg) arg$dsb_normalized_matrix)
aml_data_ebayes <- do.call("cbind", aml_data_ebayes_list)

aml_data[["ADT_dsb_ebayes"]] <- CreateAssayObject(data = aml_data_ebayes)

# remove objects to clear up memory
rm("qc_df")
rm("rna_size")
rm("adt_size")
rm("bg_rna_matrix")
rm("raw_bc")

dsb_norm_mat_no_iso <- dsb::DSBNormalizeProtein(
  
  cell_protein_matrix = aml_data@assays$ADT@counts, 
  empty_drop_matrix = bg_adt_matrix, 
  denoise.counts = FALSE, 
  pseudocount.use = 1
  
)

aml_data[["ADT_dsb_no_iso"]] <- CreateAssayObject(data = dsb_norm_mat_no_iso)

```


### Cell state annotation
```{r cell_state_annotation_for_aml_samples}

aml_data <- JoinLayers(aml_data, assay = "RNA")

Idents(aml_data) <- "seurat_clusters_harmony"
cell_type_markers <- FindAllMarkers(aml_data, assay = "ADT")

# broad cell type markers
DotPlot(aml_data, 
        features = c("CD34", "CD38", "CD3", "CD4", "CD8", "CD19", "CD22", "CD14", "CD15", "CD16", "CD11C", "CD11B", "CD9", "CD56", "CD244"), 
        group.by = "seurat_clusters_harmony",
        assay = "ADT_limma") + 
  rotate_x_labels()

# erythroiod rna markers
DotPlot(aml_data, features = c("GYPA", "GYPB", "HBB", "CD4", "CD8A", "CD8B", "PDCD1", "HAVCR2", "LAG3", "CXCR3"), 
        group.by = "seurat_clusters_harmony", 
        assay = "RNA") + 
  rotate_x_labels()

# nk cell rna markers
DotPlot(aml_data, 
        features = c("CD244", "CTSW", "FASLG", "GZMA", "GZMB", "GZMH", "IL18RAP", "IL2RB", "KIR2DL4", "KLRB1", "KLRC3", "KLRD1", "KLRF1", "KLRK1", "NCR1", "NKG7", "PRF1", "XCL1", "XCL2"), 
        group.by = "seurat_clusters_harmony", 
        assay = "RNA") + 
  rotate_x_labels()

aml_data <- aml_data %>%
  dplyr::mutate(cell_state_harmony_rna = case_when(seurat_clusters_harmony %in% c(23) ~ "T cell Erythroid doublet",
                                                   seurat_clusters_harmony %in% c(6) ~ "CD4 T cell",
                                                   seurat_clusters_harmony %in% c(5, 27) ~ "CD8 T cell",
                                                   seurat_clusters_harmony %in% c(24) ~ "DP T cell",
                                                   seurat_clusters_harmony %in% c(22) ~ "NKT cell",
                                                   seurat_clusters_harmony %in% c(25, 21) ~ "B cell",
                                                   seurat_clusters_harmony %in% c(19) ~ "Dendritic cell",
                                                   seurat_clusters_harmony %in% c(3) ~ "Monocyte",
                                                   seurat_clusters_harmony %in% c(18, 9) ~ "Macrophage",
                                                   seurat_clusters_harmony %in% c(8, 28, 20, 17, 15, 13, 11) ~ "Erythroid",
                                                   TRUE ~ "Blast"),
    
                cell_state_harmony_rna = ifelse(sample_id %in% c("Patient_31_Diagnosis", "Patient_41_Relapse", "Patient_44_Relapse") & seurat_clusters_harmony %in% "Monocyte", 
                                                "Blast", 
                                                as.character(cell_state_harmony_rna)), # AMoL samples identified by flow
                
                cell_state_harmony_rna = factor(cell_state_harmony_rna, levels = c("Blast", 
                                                                                   "Monocyte",
                                                                                   "Macrophage",
                                                                                   "Dendritic cell", 
                                                                                   "CD4 T cell", 
                                                                                   "CD8 T cell", 
                                                                                   "DP T cell", 
                                                                                   "NKT cell",
                                                                                   "B cell", 
                                                                                   "Erythroid", 
                                                                                   "T cell Erythroid doublet")))

# monocyte_data <- filter(aml_data, cell_state_harmony_rna == "Monocyte")
# 
# monocyte_data <- FindNeighbors(monocyte_data, 
#                                reduction = "harmony", 
#                                graph.name = c("nn_harmony", "snn_harmony"), 
#                                dims = 1:10)
# 
# monocyte_data <- FindClusters(monocyte_data, 
#                               graph.name = "snn_harmony", 
#                               cluster.name = "seurat_clusters_harmony_monocyte", 
#                               resolution = 0.3, 
#                               method = "igraph", 
#                               random.seed = 8791)
# 
# monocyte_data <- RunUMAP(monocyte_data, 
#                     reduction = "harmony", 
#                     dims = 1:20, 
#                     reduction.key = "UMAP_harmony_monocyte", 
#                     reduction.name = "umap_harmony_monocyte", 
#                     repulsion.strength = 5, 
#                     seed.use = 1341)
# 
# 
# DimPlot(monocyte_data, group.by = "seurat_clusters_harmony_monocyte", reduction = "umap_harmony_monocyte")
# DotPlot(monocyte_data, assay = "ADT", features = c("CD34", "CD38", "CD45", "CD14", "CD15", "CD16", "CD64"), group.by = "seurat_clusters_harmony_monocyte")

# plot umap
DimPlot(aml_data, 
        reduction = "umap_harmony", 
        group.by = "cell_state_harmony_rna", 
        label = TRUE, 
        repel = TRUE,
        label.size = 7,
        alpha = 0.2,
        shuffle = TRUE) + 
  NoLegend() + 
  NoAxes() +
  ggsci::scale_color_ucscgb()

# remove doublets and plot umap
aml_data_filt <- dplyr::filter(aml_data, !(cell_state_harmony_rna %in% c("DP T cell", "T cell Erythroid doublet")))
aml_data_filt <- dplyr::mutate(aml_data_filt, cell_state_harmony_rna = as.factor(cell_state_harmony_rna))

DimPlot(aml_data_filt, 
        reduction = "umap_harmony", 
        group.by = "cell_state_harmony_rna", 
        label = TRUE, 
        repel = TRUE,
        label.size = 5,
        alpha = 0.7) + 
  NoLegend() + 
  NoAxes() +
  ggsci::scale_color_ucscgb()

FeaturePlot(aml_data_filt,
            reduction = "umap_harmony",
            features = c("CD33", "CLL1", "CD34", "CD45"),
            order = TRUE,
            cols = c("lightblue", "violetred")) &
NoAxes()

set.seed(264599)
cell_states <- as.character(unique(aml_data_filt$cell_state_harmony_rna))
cell_states <- cell_states[!grepl("doublet", cell_states, ignore.case = TRUE)]
random_cells <- lapply(cell_states, function(arg) {
  sample(aml_data_filt$barcode[aml_data_filt$cell_state_harmony_rna %in% arg], 100, replace = FALSE)
}) %>% unlist()

DoHeatmap(aml_data_filt %>% 
            dplyr::mutate(cell_state_harmony_rna = factor(as.character(cell_state_harmony_rna), 
                                                          levels = c("Blast", "Monocyte", "Dendritic cell", "CD4 T cell", "CD8 T cell", "B cell", "Erythroid"))), 
          assay = "ADT", 
          cells = random_cells , 
          group.by = "cell_state_harmony_rna", 
          features = c("CD34", "CD117", "CD90", "CD133", "CD14", "CD15", "CD16", "CD33", "CLL1", "CD123", "CD11B", "CD11C", "CD3", "CD4", "CD8", "CD19", "CD235A", "CD71", "CD45"),
          size = 4,
          group.colors = pal_ucscgb("default")(26),
          raster = FALSE)
  

aml_data[["ADT"]] <- as(object = aml_data[["ADT"]], Class = "Assay5")

# SeuratObject::SaveSeuratRds(
#   object = aml_data, 
#   file = "aml/aml_atlas_filtered_integrated_seurat_paired_disease_only_20240110.rds"
# )

```

```{r}

sampled_data <- subset(aml_data, cells = sample(colnames(aml_data), 50000, replace = FALSE))
gif_data <- sampled_data@reductions$umap_harmony_3d@cell.embeddings %>% as.data.frame()
gif_data <- dplyr::mutate(gif_data, cell_state = sampled_data$cell_state_harmony_rna)

N <- 100
x <- rnorm(N, mean = 50, sd = 2.3)
y <- runif(N, min = 0, max = 100)
z <- runif(N, min = 4, max = 70)
luci.frame <- data.frame(x, y, z)

Noax <- list(
  title = "",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE,
  showgrid = FALSE
)

fig <- plotly::plot_ly(
  type = "scatter3d",
  mode = "markers",
  data = as.data.frame(gif_data),
  color = ~cell_state,
  colors = c("red", "orange", "yellow", "green", "blue", "aquamarine", "forestgreen","gray", "pink"),
  size = 0.1,
  x = ~ UMAPharmony3d_1,
  y = ~ UMAPharmony3d_2,
  z = ~ UMAPharmony3d_3
) %>%
  plotly::layout(scene = list(xaxis = list(visible = FALSE), 
                              yaxis = list(visible = FALSE),
                              zaxis = list(visible = FALSE),
                              camera = list(
                                eye = list(
                                  x = 1.5,
                                  y = 1.5,
                                  z = 0.5
                                  ),
                                center = list(x = 0,
                                              y = 0,
                                              z = 0)
                                ))) %>%
  onRender("
      function(el, x){
  var id = el.getAttribute('id');
  var gd = document.getElementById(id);
  Plotly.update(id).then(attach);
  function attach() {
    var cnt = 0;
    
    function run() {
      rotate('scene', Math.PI / 200);
      requestAnimationFrame(run);
    } 
    run();
    
    function rotate(id, angle) {
      var eye0 = gd.layout[id].camera.eye
      var rtz = xyz2rtz(eye0);
      rtz.t += angle;
      
      var eye1 = rtz2xyz(rtz);
      Plotly.relayout(gd, id + '.camera.eye', eye1)
    }
    
    function xyz2rtz(xyz) {
      return {
        r: Math.sqrt(xyz.x * xyz.x + xyz.y * xyz.y),
        t: Math.atan2(xyz.y, xyz.x),
        z: xyz.z
      };
    }
    
    function rtz2xyz(rtz) {
      return {
        x: rtz.r * Math.cos(rtz.t),
        y: rtz.r * Math.sin(rtz.t),
        z: rtz.z
      };
    }
  };
}
    ")


htmlwidgets::saveWidget(plotly::partial_bundle(fig), file = "rotate_scatter3d.HTML", selfcontained = TRUE)

utils::browseURL("rotate_scatter3d.HTML")

```


# Everything
```{r integrate_data_all, eval = FALSE}

# Normalize, scale, and find variable features 
mydata <- NormalizeData(orig_data, normalization.method = "LogNormalize", assay = "RNA", margin = 2)
mydata <- FindVariableFeatures(mydata, assay = "RNA")
mydata <- ScaleData(mydata, assay = "RNA") 
mydata <- RunPCA(mydata, assay = "RNA", seed.use = 466) 
mydata <- RunUMAP(mydata, dims = 1:25, assay = "RNA", seed.use = 59111)

set.seed(9912)
mydata <- IntegrateLayers(
  object = mydata, 
  assay = "RNA",
  group.by = "patient",
  method = HarmonyIntegration, #other methods don't work with Seurat v5 beta
  orig.reduction = "pca", 
  verbose = TRUE
)

mydata <- FindNeighbors(mydata, 
                        reduction = "harmony", 
                        graph.name = c("nn_harmony", "snn_harmony"), 
                        dims = 1:50)

mydata <- FindClusters(mydata, 
                       graph.name = "snn_harmony", 
                       cluster.name = "seurat_clusters_harmony", 
                       resolution = 2, 
                       method = "igraph", 
                       random.seed = 8791)

mydata <- RunUMAP(mydata, 
                  reduction = "harmony", 
                  dims = 1:20, 
                  reduction.key = "UMAP_harmony_", 
                  reduction.name = "umap_harmony", 
                  repulsion.strength = 2, 
                  seed.use = 11941)

mydata <- JoinLayers(mydata, assay = "ADT")

DefaultAssay(mydata) <- "ADT"

# convert ADT counts to sparse matrix 
mydata[["ADT"]] <- as(object = mydata[["ADT"]], Class = "Assay")

# ignore variation from isotype controls
adt_feats <- rownames(mydata@assays$ADT)
adt_feats <- adt_feats[!grepl("isotype", adt_feats, ignore.case = TRUE)]

mydata <- mydata %>%
  NormalizeData(normalization.method = "CLR", assay = "ADT", margin = 2) %>%
  ScaleData(assay = "ADT", features = adt_feats) %>%
  RunPCA(assay = "ADT", reduction.name = "pca_adt", features = adt_feats, seed.use = 6587) %>%
  FindNeighbors(reduction = "pca_adt", graph.name = "snn_adt") %>%
  FindClusters(graph.name = "snn_adt", cluster.name = "seurat_clusters_adt", random.seed = 1193) %>%
  RunUMAP(reduction = "pca_adt", reduction.name = "umap_adt", seed.use = 698, dims = 1:10)

DimPlot(mydata, reduction = "umap_harmony")

```

#### Reference mapping
##### Transfer labels from healthy reference to aml atlas
```{r reference_mapping, eval = TRUE}

DefaultAssay(healthy_data) <- "RNA"
DefaultAssay(aml_data) <- "RNA"

#healthy_data <- JoinLayers(healthy_data, assay = "RNA")

# create reference from healthy data for projection
# options(Seurat.object.assay.version = 'v3')
# reference <- CreateSeuratObject(counts = as(LayerData(healthy_data, assay = "RNA", layers = "counts"), "dgCMatrix"), 
#                                 meta.data = healthy_data@meta.data)
# reference@assays$RNA@data <- healthy_data@assays$mnn.reconstructed@data
# reference <- FindVariableFeatures(reference)

# add reductions
# reference@reductions$integrated.mnn <- healthy_data@reductions$integrated.mnn
# reference@reductions$umap_mnn <- healthy_data@reductions$umap_mnn
# reference@reductions$pca <- healthy_data@reductions$pca
# DefaultAssay(reference) <- "RNA"
# mysplit <- SplitObject(reference, split.by = "sample_id")
# mysplit <- lapply(mysplit, function(arg) {
#   arg <- FindVariableFeatures(arg, nfeatures = 500)
#   arg <- NormalizeData(arg)
#   return(arg)
# })
# anchors <- FindIntegrationAnchors(mysplit[c(1, 3)]) 
# healthy_data <- IntegrateData(anchorset = anchors)


query_list <- SplitObject(aml_data, split.by = "sample_id")
cell_state_list <- NULL
for (query in names(query_list)) {
  
  cat("\r", query)
  query_sample <- query_list[[query]]
  query_sample <- CreateSeuratObject(counts = as(LayerData(query_sample, layer = "counts"), "dgCMatrix"), meta.data = query_sample@meta.data)
  query_sample <- query_sample %>%
    FindVariableFeatures() %>%
    NormalizeData() %>%
    ScaleData() %>% 
    RunPCA()
  
  anchors <- FindTransferAnchors(
    reference = healthy_data,
    query = query_sample,
    reference.assay = "RNA",
    query.assay = "RNA",
    normalization.method = "LogNormalize",
    reference.reduction = "pca",
    dims = 1:50,
    project.query = FALSE
  )

  cell_state_list[[query]] <- TransferData(anchorset = anchors, refdata = healthy_data$cell_state_mnn_rna_l1, dims = 1:50)
  
}

cell_state_prediction <- do.call("rbind", cell_state_list) %>%
  janitor::clean_names()
#cell_state_prediction <- cell_state_prediction[match(colnames(aml_data), rownames(cell_state_prediction)), ]
stopifnot(all(gsub(".*\\.", "", rownames(cell_state_prediction)) == colnames(aml_data)))
aml_data <- AddMetaData(aml_data, metadata = cell_state_prediction)

table(aml_data$cell_state_harmony_rna, aml_data$predicted_id)
table(dplyr::filter(aml_data, prediction_score_max > 0.8) %>% dplyr::pull(cell_state_harmony_rna),
      dplyr::filter(aml_data, prediction_score_max > 0.8) %>% dplyr::pull(predicted_id))
mytable <- table(aml_data$cell_state_harmony_rna, aml_data$predicted_id)
pheatmap(mytable, display_numbers = TRUE, number_color = "darkblue", fontsize_number = 9)


```

##### Transfer labels from Triana reference to aml atlas
```{r reference_mapping_triana, eval = TRUE}

triana <- aws.s3::s3readRDS(object = "s3://vorbio-datascience/genomics/external_data/single_cell_rnaseq/triana/WTA_projected.rds")
cells_to_remove <- table(triana$ct)
cells_to_remove <- names(cells_to_remove)[cells_to_remove < 10]
triana <- dplyr::filter(triana, !ct %in% cells_to_remove) %>%
  dplyr::mutate(ct = factor(ct))

query_list <- SplitObject(aml_data, split.by = "sample_id")
cell_state_list <- NULL
for (query in names(query_list)) {
  
  cat("\r", query)
  query_sample <- query_list[[query]]
  query_sample <- CreateSeuratObject(counts = as(LayerData(query_sample, layer = "counts"), "dgCMatrix"), meta.data = query_sample@meta.data)
  query_sample <- query_sample %>%
    FindVariableFeatures() %>%
    NormalizeData() %>%
    ScaleData() %>% 
    RunPCA()
  
  anchors <- FindTransferAnchors(
    reference = triana,
    query = query_sample,
    reference.assay = "RNA",
    query.assay = "RNA",
    normalization.method = "LogNormalize",
    dims = 1:50,
    project.query = FALSE
  )

  cell_state_list[[query]] <- TransferData(anchorset = anchors, refdata = triana$ct, dims = 1:50)
  
}

cell_state_prediction <- do.call("rbind", cell_state_list) %>%
  janitor::clean_names()
colnames(cell_state_prediction) <- paste("triana", colnames(cell_state_prediction), sep = "_")

aml_data <- AddMetaData(aml_data, metadata = cell_state_prediction)

table(aml_data$cell_state_harmony_rna, aml_data$triana_predicted_id)
table(dplyr::filter(aml_data, triana_prediction_score_max > 0.8) %>% dplyr::pull(cell_state_harmony_rna),
      dplyr::filter(aml_data, triana_prediction_score_max > 0.8) %>% dplyr::pull(triana_predicted_id))
mytable <- table(aml_data$cell_state_harmony_rna, aml_data$triana_predicted_id)
pheatmap(mytable, display_numbers = TRUE, number_color = "darkblue", fontsize_number = 9)


```


##### Project aml atlas onto healthy reference UMAP
```{r project_data_to_reference}

DefaultAssay(healthy_data) <- "RNA"
DefaultAssay(aml_data) <- "RNA"

healthy_data <- JoinLayers(healthy_data, assay = "RNA")

# create reference from healthy data for projection
options(Seurat.object.assay.version = 'v3')
reference <- CreateSeuratObject(counts = as(LayerData(healthy_data, assay = "RNA", layers = "counts"), "dgCMatrix"), 
                                meta.data = healthy_data@meta.data)
reference@assays$RNA@data <- healthy_data@assays$mnn.reconstructed@data

# add reductions
reference@reductions$integrated.mnn <- healthy_data@reductions$integrated.mnn
reference@reductions$umap_mnn <- healthy_data@reductions$umap_mnn
reference@reductions$pca <- healthy_data@reductions$pca
DefaultAssay(reference) <- "RNA"

# project aml atlas to healthy reference
anchors <- FindTransferAnchors(
  reference = reference,
  query = aml_data,
  reference.assay = "RNA",
  query.assay = "RNA",
  normalization.method = "LogNormalize",
  reference.reduction = "integrated.mnn",
  dims = 1:50,
  project.query = FALSE,
  features = rownames(reference@reductions$integrated.mnn@feature.loadings)
)

reference[["RNA"]] <- as(object = reference[["RNA"]], Class = "Assay5")

cell_state_prediction <- TransferData(anchorset = anchors, refdata = reference$cell_state_mnn_rna_l1, dims = 1:50)

feature.loadings <- reference@reductions$integrated.mnn@feature.loadings
feature.loadings <- feature.loadings[rownames(feature.loadings) %in% rownames(aml_data@assays$RNA), ]
pca.features <- rownames(feature.loadings)

aml_data <- ScaleData(aml_data, features = pca.features, assay = "RNA")
pca.embeddings <- t(LayerData(aml_data, assay = "RNA", layer = "scale.data")[pca.features, ]) %*% feature.loadings
aml_data[["integrated_mnn_embed"]] <- CreateDimReducObject(embeddings = pca.embeddings, key = "MNN_EMBED_", assay = "RNA")

aml_data <- ProjectUMAP(
  query = aml_data,
  k.param = 30,
  n.ntrees = 100,
  l2.norm = FALSE,
  query.reduction = "integrated_mnn_embed",
  reference = reference,
  reference.reduction = "integrated.mnn",
  reduction.model = "umap_mnn",
)

aml_data <- AddMetaData(aml_data, metadata = janitor::clean_names(cell_state_prediction))


aml_data <- dplyr::mutate(aml_data, predicted_id = factor(predicted_id, levels = c("HSC",
                                                                                    "MEP",
                                                                                    "GMP",
                                                                                    "Monocyte",
                                                                                    "DC",
                                                                                    "NK",
                                                                                    "B cell",
                                                                                    "preB",
                                                                                    "CD4+ naive T",
                                                                                    "CD4+ memory T",
                                                                                    "CD8+ naive T",
                                                                                    "CD8+ memory T",
                                                                                    "Erythroblast",
                                                                                    "Erythroid",
                                                                                    "Unknown")))
                                                          
                                                          
DimPlot(aml_data, group.by = "cell_state_harmony_rna", reduction = "ref.umap", label = TRUE, repel = TRUE) + NoAxes() + NoLegend()

DimPlot(aml_data, reduction = "umap_harmony", group.by = "predicted_id", label = TRUE, repel = TRUE, cols = as.character(mycols)) + 
  scale_x_reverse() +
  NoAxes() +
  NoLegend()

DimPlot(aml_data, reduction = "ref.umap", group.by = "predicted_id", label = TRUE, repel = TRUE, cols = as.character(mycols)) + 
  scale_x_reverse() +
  NoAxes()


```

#### Cell state annotation
```{r fig.height = 9, fig.width = 12, eval = FALSE}

adt_markers <- list(tcell = c("CD2", "CD3", "CD4", "CD8"),
                    bcell = c("CD19", "CD22"),
                    myeloid = c("CD11B", "CD14", "CD15", "CD16", "CD33"),
                    eryth = c("CD71","CD235A"),
                    blast = c("CD45"))

DotPlot(mydata, assay = "ADT", features = as.character(unlist(adt_markers)), group.by = "seurat_clusters_harmony") +
  coord_flip() +
  RotatedAxis() 

DotPlot(mydata, assay = "RNA", features = c("GYPA", "TFRC", "HBD", "SLC25A37"), group.by = "seurat_clusters_harmony") +
  coord_flip() +
  RotatedAxis() 

b_cells <- c(27, 34, 40, 72, 75, 83)
t_cells <- c(7, 8, 21, 37, 43, 57, 62, 66)
erythrocytes <- c(14, 17, 19, 22, 23, 28, 30, 31, 32, 42, 50, 53, 54, 60, 67, 68, 69, 47, 76, 79, 84, 80, 81)
monocytes <- c(2, 18, 12, 44, 48, 41, 78, 82)
multiplets <- c(36)

mydata <- mydata %>%
  dplyr::mutate(cell_state_l1 = case_when(seurat_clusters_harmony %in% b_cells ~ "B cell",
                                          seurat_clusters_harmony %in% t_cells ~ "T cell",
                                          seurat_clusters_harmony %in% erythrocytes ~ "Erythrocyte",
                                          seurat_clusters_harmony %in% monocytes ~ "Monocyte",
                                          seurat_clusters_harmony %in% multiplets ~ "Multiplet",
                                          TRUE ~ "Blast"),
                cell_state_l1 = as.character(cell_state_l1))

DimPlot(mydata, 
        group.by = "cell_state_l1", 
        reduction = "umap_harmony", 
        label = TRUE, 
        order = list("Tcell", "Blast", "B cell", "Monocyte", "Erythrocyte", "Multiplet")) + NoAxes()

```

## Check blast percentage concordance
```{r check_blast_percentage}

blast_counts <- table(aml_data$cell_state_harmony_rna, aml_data$sample_id)

col_sums <- colSums(blast_counts)
blast_counts <- sweep(x = blast_counts, MARGIN = 2, STATS = col_sums, FUN = "/")
blast_percent <- blast_counts["Blast", , drop = FALSE] %>% 
  t() %>%
  as.data.frame.matrix() %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::mutate(Blast = Blast * 100)
                      
blast_metadata <- aws.s3::s3read_using(FUN = readr::read_csv,
                                       object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/aml_atlas_sample_metadata.csv") %>%
  dplyr::select(sample_id, blast_percent_vor) %>%
  dplyr::filter(!grepl("Healthy", sample_id)) %>%
  dplyr::filter(!is.na(blast_percent_vor))

blast_data <- dplyr::inner_join(blast_percent, blast_metadata, by = "sample_id")
model_res <- lm(blast_percent_vor ~ Blast, data = blast_data)$residuals
blast_data <- dplyr::mutate(blast_data, residuals = model_res, abs_dev = abs(Blast - blast_percent_vor)) %>%
  dplyr::mutate(`Model` = ifelse(abs(residuals) < 25, "yes", "no"))

deviant_samples <- blast_data$sample_id[blast_data$Model == "no"]

ggpubr::ggscatter(blast_data,
                  x = "Blast",
                  y = "blast_percent_vor",
                  xlab = "CITEseq blast %",
                  ylab = "Flow blast %",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  #label = "sample_id",
                  cor.coef = TRUE,
                  #label.select = blast_data$sample_id[blast_data$Model == "no"],
                  repel = TRUE,
                  font.label = c(8, "plain"),
                  size = 2) +
  geom_smooth(lty = 2, method = "lm",  lwd = 0.4)

ggpubr::ggscatter(blast_data,
                  x = "Blast",
                  y = "blast_percent_vor",
                  color = "Model",
                  xlab = "CITEseq blast %",
                  ylab = "Flow blast %",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  cor.coef = TRUE,
                  cor.coef.coord = c(0, 100),
                  palette = c("lightgray", "orchid1"),
                  size = 2) +
  geom_abline(slope = 1, lty = 2)


```

# Generate table 1
```{r generate_table1}

patient_age <- aws.s3::s3read_using(FUN = readr::read_csv,
                                    object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/patient_age_at_dx.csv")

tdata <- aml_data@meta.data %>%
  dplyr::select(patient, uhn_patient_id, time, vital_status, date_of_dx, relapse_date_1, initial_tx, response_to_initial_tx, date_of_death_or_last_follow_up, blast_percent_vor, hsct, cytogenetics_1) %>%
  dplyr::distinct() %>%
  dplyr::left_join(patient_age, by = "uhn_patient_id") %>%
  dplyr::select(-patient_manuscript_id, -uhn_patient_id) %>%
  tidyr::pivot_wider(names_from = time, values_from = blast_percent_vor) %>%
  dplyr::mutate(sex = dplyr::case_when(grepl("XX", cytogenetics_1) & !grepl("XY", cytogenetics_1) ~ "Female",
                                       grepl("XY", cytogenetics_1) & !grepl("XX", cytogenetics_1) ~ "Male",
                                       patient == "Patient_49" ~ "Male"),
                date_of_dx = as_date(date_of_dx),
                relapse_date_1 = as_date(relapse_date_1),
                date_of_death_or_last_follow_up = as_date(date_of_death_or_last_follow_up),
                days_to_relapse = relapse_date_1 - date_of_dx,
                days_to_last_follow_up = date_of_death_or_last_follow_up - date_of_dx,
                #initial_tx = ifelse(initial_tx == "3+7", "7+3", initial_tx),
                initial_tx = gsub("Clinical Trial with ", "", initial_tx),
                initial_tx = ifelse(initial_tx == "SGI", "SGI-110", initial_tx),
                hsct = ifelse(hsct == "HSCT", "Yes", "No"),
                vital_status = ifelse(vital_status == "Dead", "Deceased", "Alive")) %>%
  dplyr::select(-patient, -date_of_dx, -relapse_date_1, -date_of_death_or_last_follow_up, -cytogenetics_1) %>%
  dplyr::select(sex, age_at_aml_dx, Diagnosis, Relapse, initial_tx, response_to_initial_tx, hsct, vital_status, days_to_relapse, days_to_last_follow_up) %>%
  dplyr::mutate(response_to_initial_tx = gsub(" \\(CR\\)", "", response_to_initial_tx),
                response_to_initial_tx = gsub("response", "remission", response_to_initial_tx),
                response_to_initial_tx = dplyr::case_when(response_to_initial_tx %in% "Complete remission" ~ "Complete remission (CR)",
                                                          response_to_initial_tx %in% "No remission" ~ "Primary refractory disease (PRD)",
                                                          response_to_initial_tx %in% "Partial remission" ~ "Partial remission (PR)"))

colnames(tdata) <- c("Sex", "Age at diagnosis", "Blast % at diagnosis", "Blast % at relapse", "Initial treatment", "Response to inital treatment", "Prior HCT", "Vital status", "Days to first relapse", "Days to last follow up")
tdata |> tbl_summary() |> modify_caption("**Table 1. Patient Characteristics**") |> bold_labels()


tdata <- dplyr::select(tdata, Sex, `Age at diagnosis`, `Blast % at diagnosis`, `Blast % at relapse`, `Initial treatment`, `Response to inital treatment`, `Prior HCT`)
tdata |> tbl_summary() |> modify_caption("**Table 1. Patient Characteristics**") |> bold_labels()

```

```{r save_objects}

output_path <- "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/seurat_2024-04-03/aml_data_v2.rds"

if (!file.exists(output_path)) saveRDS(aml_data, file = output_path)

output_path <- "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/seurat_2024-04-03/reference_v2.rds"

if (!file.exists(output_path)) saveRDS(aml_data, file = output_path)

```
