---
title: "Transcriptomic analysis of cell states across timepoints"
output: html_document
date: "2024-09-11"
---

```{r setup, include = FALSE}

library(devtools)
library(flowCore)
library(SeuratObject)
library(BPCells)
library(Seurat)
library(SeuratDisk)
library(dplyr)
library(aws.s3)
library(tibble)
library(tidyr)
library(dotenv)
library(ggplot2)
library(ggpubr)
library(glue)
library(DoubletFinder)
library(tidyseurat)
library(pzfx)
library(preprocessCore)
library(pals)
library(EnhancedVolcano)
library(SeuratWrappers)
library(alphahull)
library(igraph)
library(reticulate)
library(viridis)
library(msigdbr)
library(GSVA)
library(escape)
library(dittoSeq)
library(fgsea)
library(GSEABase)
library(ggbeeswarm)
library(circlize)
library(pheatmap)
library(biomaRt)
library(SingleCellExperiment)
library(Matrix)
library(stringr)
library(ComplexHeatmap)
library(future)
library(circlize)
library(msigdbr)
library(survival)
library(survminer)
library(broom)
library(scales)
library(lubridate)
library(future)
library(ggsci)
library(bazar)
library(flowCore)
library(mclust)
library(doParallel)
library(foreach)
library(scales)
library(tictoc)
library(randomForest)
library(e1071)
library(survminer)
library(robCompositions)
library(limma)
library(infercnv)
library(vcfR)
library(edgeR)
library(patchwork)

options(Seurat.object.assay.version = "v3")
options(future.globals.maxSize = 1e9)

load_dot_env(file = "/home/rstudio/data/bitbucket/.env")

knitr::opts_knit$set(root.dir = "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas")

Sys.setenv(
  "AWS_ACCESS_KEY_ID" = Sys.getenv("key"),
  "AWS_SECRET_ACCESS_KEY" = Sys.getenv("secret"),
  "AWS_DEFAULT_REGION" = "us-east-1"
)

cl <- makeCluster(12)
registerDoParallel(cl)

```

# Load data
```{r load_data}

aml_data <- aws.s3::s3readRDS(object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/seurat_v5/seurat_2024-04-03/aml_data_v4.rds")

patient_order <- c("P26", "P44", "P52", "P70", "P61", "P31", "P43", "P30", "P64", "P56", "P54", "P27", "P41", "P49", "P51", "P50", "P66", "P29", "P67", "P42", "P65", "P60", "P58", "P28", "P40", "P4") 
patient_id_map <- data.frame(patient = patient_order, 
                             patient_manuscript_id = paste0("P", 1:length(patient_order))) %>%
  dplyr::mutate(patient = factor(patient, levels = patient_order),
                patient_manuscript_id = ifelse(nchar(patient_manuscript_id) == 2, gsub("P", "P0", patient_manuscript_id), patient_manuscript_id),
                patient_manuscript_id = factor(patient_manuscript_id, levels = patient_manuscript_id))

#healthy_data <- readRDS("/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/seurat_2024-04-03/healthy_data_v2.rds")
healthy_data <- aws.s3::s3readRDS(object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/seurat_v5/seurat_2024-04-03/healthy_data_v4.rds")

blast_object <- dplyr::filter(aml_data, 
                              cell_state_harmony_rna %in% "Blast",
                              !cell_state_harmony_rna %in% "Unknown") %>%
  dplyr::mutate(cell_state_harmony_rna = droplevels(cell_state_harmony_rna))

lsc17 <- list(lsc17 = c("AKR1C3",
                        "ARHGAP22",
                        "TMEM259", # (C19orf177)
                        "CD34",
                        "CDK6",
                        "CPXM1",
                        "DNMT3B",
                        "DPYSL3",
                        "EMP1",
                        "ADGRG1",
                        "FAM30A",
                        "LAPTM4B",
                        "MMRN1",
                        "BEX3",
                        "NYNRIN",
                        "SOCS2",
                        "ZBTB46"))


blast_object <- AddModuleScore(blast_object, 
                               features = lsc17,
                               assay = "RNA",
                               name = "lsc17") %>%
  dplyr::rename(lsc17 = lsc171)

lscs <- blast_object$barcode[blast_object@assays$ADT@data["CD34", ] > 1 & blast_object@assays$ADT@data["CD38", ] < 1 & blast_object$lsc17 > 0]
lsc_object <- dplyr::filter(blast_object, barcode %in% lscs)

blast_object <- blast_object %>%
  dplyr::mutate(lsc = ifelse(barcode %in% lsc_object$barcode, "LSC", "blast"))

mycols <- c("firebrick1", "lightcyan3", "seagreen3", "cyan", "purple", "hotpink", "gold1", "mediumturquoise", "yellow2", "sienna3", "springgreen", "olivedrab", "orange1",  "cornflowerblue", "rosybrown1", "maroon2", "slateblue2", "tan2", "peachpuff1", "orchid1", "yellowgreen", "burlywood2")

```

# Plot blast percentages from flow
```{r blast_percentages_flow}

df <- aml_data@meta.data %>%
  dplyr::select(patient_manuscript_id, time, blast_percent_vor) %>%
  dplyr::distinct() %>%
  dplyr::mutate(percent_remaining = 100 - blast_percent_vor) %>%
  tidyr::pivot_longer(-c(patient_manuscript_id, time), values_to = "value", names_to = "portion")

# Create the pie charts with percentage labels (only on the "percent" slice)
ggplot(df, aes(x = "", y = value, fill = portion)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar("y", clip = "off") +
  facet_grid(time ~ patient_manuscript_id) +   # Arrange all charts in one row
  theme_void() +                  # Removes axes, gridlines, and background
  scale_fill_manual(values = c("bisque1", "white"))

```

# Visualize important features
```{r featureplots}

DefaultAssay(aml_data) <- "ADT"
aml_data <- ScaleData(aml_data, assay = "ADT", features = rownames(aml_data@assays$ADT))
aml_data <- ScaleData(aml_data, assay = "RNA", features = rownames(aml_data@assays$RNA))
FeaturePlot(aml_data, features = c("CD45"), reduction = "umap_harmony", slot = 'scale.data', keep.scale = "all", order = TRUE) + 
  NoAxes() +
  scale_color_gradientn( colours = c('bisque', 'cornflowerblue', 'yellow'),  limits = c(-2, 2))

DefaultAssay(aml_data) <- "RNA"
FeaturePlot(aml_data, features = c("PTPRC"), reduction = "umap_harmony", slot = 'scale.data', keep.scale = "all", order = TRUE) + 
  NoAxes() +
  scale_color_gradientn( colours = c('bisque', 'cornflowerblue', 'yellow'),  limits = c(-2, 2))

DoHeatmap(blast_object, group.by = "cell_state_harmony_rna", assay = "RNA", features = c("GMSCF", "FLT3L", "TNF", "TGFB", "CTLA4", "PDL1", "PD1"))
VlnPlot(blast_object, group.by = "cell_state_harmony_rna", assay = "RNA", features = c("GMSCF", "FLT3L", "TNF", "TGFB", "CTLA4", "CD274", "PDCD1"), pt.size = 0)

```

# Visualize cell state composition
## Blasts
```{r cell_state_composition, fig.width = 12, fig.height = 5}

cell_states <- c("HSC",
                 "GMP",
                 "MEP",
                 "Monocyte",
                 "cDC",
                 "pDC",
                 "NK",
                 "NKT",
                 "B cell",
                 "Pre/pro B cell",
                 "CD4+ naive T",
                 "CD4+ memory T",
                 "CD8+ naive T",
                 "CD8+ memory T",
                 "Erythroblast",
                 "Erythroid",
                 "MkP",
                 "Unknown")

cluster_summary <- blast_object@meta.data

cluster_summary <- cluster_summary %>%
  dplyr::filter(!predicted_id %in% "Unknown") %>%
  dplyr::group_by(time, predicted_id, patient) %>%
  dplyr::summarise(count = n(), 
                   response_to_initial_tx = unique(response_to_initial_tx),
                   date_of_dx = unique(date_of_dx),
                   relapse_date_1 = unique(relapse_date_1),
                   relapse_date_2 = unique(relapse_date_2),
                   vital_status = unique(vital_status),
                   date_of_death_or_last_follow_up = unique(date_of_death_or_last_follow_up),
                   blast_percent_vor = unique(blast_percent_vor),
                   lsc_percentage = unique(lsc_percentage),
                   fab_classification = unique(fab_classification),
                   cytogenetics_1 = unique(cytogenetics_1)) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(patient, time) %>%
  dplyr::mutate(pct = (round((count / sum(count)), 3)) * 100,
                time = factor(as.character(ifelse(time == "Diagnosis", "Dx", "Re")), levels = c("Dx", "Re")),
                patient = gsub("Patient_", "P", patient),
                predicted_id = factor(predicted_id, levels = cell_states)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(`Blast cell state` = predicted_id) %>%
  dplyr::mutate(relapse_status = "Relapsed",
                time_dx_to_re = relapse_date_1 - date_of_dx,
                time_re_to_last_follow_up = date_of_death_or_last_follow_up - relapse_date_1,
                time_dx_to_last_follow_up = date_of_death_or_last_follow_up - date_of_dx,
                sex = ifelse(grepl("XX", cytogenetics_1), "female", "male"))

hsc_rank <- dplyr::filter(cluster_summary, `Blast cell state` == "HSC", time == "Dx") %>%
  dplyr::arrange(desc(pct)) %>%
  dplyr::pull(patient)

hsc_rank <- c(hsc_rank, unique(cluster_summary$patient[!cluster_summary$patient %in% hsc_rank]))

cluster_summary <- cluster_summary %>%
  dplyr::mutate(patient = factor(patient, levels = hsc_rank),
                broad_cell_state = dplyr::case_when(`Blast cell state` %in% c("NK", "NKT", "B cell", "CD4+ T", "CD4+ memory T", "CD8+ T", "CD8+ memory T", "CD4+ naive T", "CD8+ naive T") ~ "Lymphoid", 
                                                    TRUE ~ as.character(`Blast cell state`)),
                broad_cell_state = factor(broad_cell_state, levels = c("HSC", "GMP", "MEP", "Monocyte", "cDC", "pDC", "MkP", "Pre/pro B cell", "Erythroblast", "Erythroid", "Lymphoid"))) %>%
  dplyr::filter(!broad_cell_state %in% c("Erythroid", "Lymphoid"))
levels(cluster_summary$broad_cell_state) <- paste0(levels(cluster_summary$broad_cell_state), "-like")

mycols <- c("firebrick1", "seagreen3", "lightcyan3", "cyan", "purple", "hotpink", "slateblue2", "yellow2", "rosybrown1", "maroon2", "cornflowerblue")
other_colors <- c("gold1", "mediumturquoise",  "sienna3", "springgreen", "olivedrab", "orange1",  "cornflowerblue", "maroon2", "tan2", "peachpuff1", "orchid1", "yellowgreen", "burlywood2")

cluster_summary <- dplyr::mutate(cluster_summary, patient = factor(patient, levels = patient_order)) %>%
  dplyr::left_join(patient_id_map, by = "patient")


ggplot(dplyr::mutate(cluster_summary, time = ifelse(time == "Dx", "D", "R")) %>%
         dplyr::rename(`Blast lineage` = `broad_cell_state`), 
       aes(x = time, y = count, fill = `Blast lineage`)) + 
  geom_bar(position = "fill", stat = "identity", width = 0.7) + 
  xlab("") +
  ylab("Fraction of blasts") +
  facet_grid(~patient_manuscript_id) +
  scale_fill_manual(values = mycols) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14)) 


jfeat <- c("CD274", "GMSCF", "FLT3", "TNF", "CTLA4")
boo <- rownames(aml_data@assays$RNA)
boo <- boo[grepl("^EGF|^FGF|^TGF", boo)]
jfeat <- c(jfeat, boo)


DoHeatmap(subset(aml_data, cells = sample(colnames(aml_data, 300))), features = jfeat, group.by = "cell_state_harmony_rna")

```

## Monocytes
```{r monocytes_m1m2_composition}

monocytes <- dplyr::filter(aml_data, cell_state_harmony_rna %in% "Monocyte")
DefaultAssay(monocytes) <- "RNA"
monocytes@assays$RNA@layers$data <- NULL
monocytes@assays$RNA@layers$scale.data <- NULL
# feats_include <- rownames(monocytes@assays$RNA)[rowSums(monocytes@assays$RNA@layers$counts) > 1]
# feats_include <- c(feats_include, rownames(monocytes@assays$ADT))
# monocytes <- subset(monocytes, features = feats_include)

monocytes[["RNA"]] <- split(monocytes[["RNA"]], f = monocytes$sample_id)
monocytes <- FindVariableFeatures(monocytes, selection.method = "dispersion")
monocytes <- NormalizeData(monocytes)
monocytes <- ScaleData(monocytes)
monocytes <- RunPCA(monocytes)
Idents(monocytes) <- "sample_id"
monocytes <- IntegrateLayers(object = monocytes, assay = "RNA", method = FastMNNIntegration)
monocytes <- FindNeighbors(monocytes, reduction = "integrated.mnn")
monocytes <- FindClusters(monocytes, random.seed = 159828)
monocytes <- RunUMAP(monocytes, reduction = "integrated.mnn", dims = 1:50, seed.use = 1195)

DefaultAssay(monocytes) <- "RNA"
VlnPlot(monocytes, 
        group.by = "seurat_clusters", 
        features = c("CD33", "IL4", "IL6", "CD86", "MRC1", "CD163", "CD80", "IL6", "IL10", "CD274"), 
        pt.size = 0,
        layer = "data")  


Idents(monocytes) <- "seurat_clusters"
tmp_object <- CreateSeuratObject(counts = as(LayerData(JoinLayers(monocytes), assays = "RNA", layer = "counts"), "dgCMatrix"),
                                 meta.data = monocytes@meta.data)
Idents(tmp_object) <- "seurat_clusters"
mono_markers <- FindAllMarkers(tmp_object, latent.vars = "sample_id", test.use = "MAST", only.pos = TRUE, assay = "RNA")




monocytes <- dplyr::mutate(monocytes, m2 = ifelse(seurat_clusters %in% 3, "M2", "other"))
monocytes_re <- dplyr::filter(monocytes, time %in% "Relapse")
monocytes_dx <- dplyr::filter(monocytes, time %in% "Diagnosis")

re_tab <- table(monocytes_re$m2, monocytes_re$response_to_initial_tx)
re_tab <- sweep(re_tab, MARGIN = 2, STATS = colSums(re_tab), FUN = "/")

dx_tab <- table(monocytes_dx$m2, monocytes_dx$response_to_initial_tx)
dx_tab <- sweep(dx_tab, MARGIN = 2, STATS = colSums(dx_tab), FUN = "/")


mac_gs <- immune_gs_list[grepl("GSE5099_CLASSICAL_M1_VS_ALTERNATIVE_M2_MACROPHAGE", names(immune_gs_list))]

m1_genes_nc <- c("S100A8", "S100A9", "S100A12", "TYROBP", "VCAN", 
                "CD68", "MNDA", "CYBB", "STAT1", "FOS", "TREM1", 
                "CD44", "NFKB2", "IL1B", "CD60", "CD80", "CD86", "IFNG")

mac_gs <- c(mac_gs, m1_nc = list(m1_genes_nc))
names(mac_gs) <- c("M2", "M1", "M1_NC")
mac_gs <- mac_gs[c(2, 3, 1)]
monocytes <- AddModuleScore(object = JoinLayers(monocytes), 
                            features = mac_gs, 
                            name = "M", search = TRUE, assay = "RNA") %>%
  dplyr::rename(M1_NC = M2, M2 = M3)

monocytes <- monocytes %>%
  dplyr::mutate(M1_cat = ifelse(M1_NC > median(M1_NC), "M1", "M2"),
                M2_cat = ifelse(M2 > median(M2), "M2", "M1")) %>%
  dplyr::mutate(M_consensus = ifelse(M1_cat == M2_cat, M1_cat, NA))

bardata <- table(monocytes$M_consensus, monocytes$sample_id) %>% 
  t() %>% 
  as.data.frame() %>%
  dplyr::rename(sample_id = Var1, macrophage = Var2, freq = Freq) %>%
  dplyr::mutate(freq = freq + 1) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(M1_M2_ratio = log2(freq[macrophage == "M1"] / freq[macrophage == "M2"])) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(M1_M2_ratio) %>%
  dplyr::mutate(time = ifelse(grepl("Diagnosis", sample_id), "D", "R"),
                patient = gsub("_relapse|_Relapse", "", sample_id),
                patient = gsub("Patient_", "P", patient),
                patient = factor(patient, levels = levels(cluster_summary$patient)))

ggbarplot(data = bardata,
          x = "time",
          y = "freq",
          fill = "macrophage",
          palette = "npg") +
  facet_grid(~patient) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

bardata_ratio <- dplyr::select(bardata, patient, time, M1_M2_ratio) %>% 
  distinct() %>%
  dplyr::arrange(desc(M1_M2_ratio)) 

tmp_order <- dplyr::filter(bardata_ratio, time == "D") %>%
  dplyr::arrange(desc(M1_M2_ratio)) %>%
  dplyr::pull(patient)

bardata_ratio <- dplyr::mutate(bardata_ratio, patient = factor(patient, levels = tmp_order))
# pat_surv <- dplyr::arrange(mutation_data, time_dx_to_last_follow_up) %>% dplyr::pull(patient)
# pat_surv <- gsub("Patient_", "P", pat_surv)
# 
# bardata_ratio <- dplyr::slice(bardata_ratio, match(pat_surv, patient))

ggbarplot(data = bardata_ratio,
          x = "patient",
          y = "M1_M2_ratio",
          fill = "time",
          position = position_dodge(),
          palette = "npg")



```

## CD34+CD38- LSCs
```{r cell_state_composition_lscs, fig.width = 12, fig.height = 5}

cell_states <- c("HSC",
                  "MEP",
                  "GMP",
                  "Monocyte",
                  "DC",
                  "NK",
                  "B cell",
                  "preB",
                  "CD4+ naive T",
                  "CD4+ memory T",
                  "CD8+ naive T",
                  "CD8+ memory T",
                  "Erythroblast",
                  "Erythroid",
                  "Unknown")

cluster_summary <- lsc_object@meta.data
cluster_summary <- cluster_summary %>%
  dplyr::filter(!predicted_id %in% "Unknown") %>%
  dplyr::group_by(time, predicted_id, patient) %>%
  dplyr::summarise(count = n(), 
                   response_to_initial_tx = unique(response_to_initial_tx),
                   date_of_dx = unique(date_of_dx),
                   relapse_date_1 = unique(relapse_date_1),
                   relapse_date_2 = unique(relapse_date_2),
                   vital_status = unique(vital_status),
                   date_of_death_or_last_follow_up = unique(date_of_death_or_last_follow_up),
                   blast_percent_vor = unique(blast_percent_vor),
                   lsc_percentage = unique(lsc_percentage),
                   fab_classification = unique(fab_classification),
                   cytogenetics_1 = unique(cytogenetics_1)) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(patient, time) %>%
  dplyr::mutate(pct = (round((count / sum(count)), 3)) * 100,
                time = factor(as.character(ifelse(time == "Diagnosis", "Dx", "Re")), levels = c("Dx", "Re")),
                patient = gsub("Patient_", "P", patient),
                predicted_id = factor(predicted_id, 
                                      levels = cell_states)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(`Blast cell state` = predicted_id) %>%
  dplyr::mutate(relapse_status = "Relapsed",
                time_dx_to_re = relapse_date_1 - date_of_dx,
                time_re_to_last_follow_up = date_of_death_or_last_follow_up - relapse_date_1,
                time_dx_to_last_follow_up = date_of_death_or_last_follow_up - date_of_dx,
                sex = ifelse(grepl("XX", cytogenetics_1), "female", "male"))

hsc_rank <- dplyr::filter(cluster_summary, `Blast cell state` == "HSC", time == "Dx") %>%
  dplyr::arrange(desc(pct)) %>%
  dplyr::pull(patient)

hsc_rank <- c(hsc_rank, unique(cluster_summary$patient[!cluster_summary$patient %in% hsc_rank]))

cluster_summary <- cluster_summary %>%
  dplyr::mutate(patient = factor(patient, levels = hsc_rank),
                broad_cell_state = ifelse(`Blast cell state` %in% c("NK", "Naive B", "preB", "memory B", "Plasma", "CD4+ T", "CD4+ memory T", "CD8+ T", "CD8+ memory T"), "Lymphocyte", as.character(`Blast cell state`)))

ggplot(dplyr::mutate(cluster_summary, time = ifelse(time == "Dx", "D", "R")) %>%
         dplyr::rename(`Blast lineage` = `Blast cell state`), 
       aes(x = time, y = count, fill = `Blast lineage`)) + 
  geom_bar(position = "fill", stat = "identity", width = 0.7) + 
  xlab("") +
  ylab("Fraction of blasts") +
  facet_grid(~patient) +
  scale_fill_manual(values = mycols) +
  theme_light() +
  theme(axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14)) 

```

## Healthy
```{r healthy_data}

healthy_data_filt <- healthy_data %>%
  dplyr::mutate(broad_cell_state = dplyr::case_when(cell_state_mnn_rna_l1 %in% c("NK", "NKT", "B cell", "CD4+ T", "CD4+ memory T", "CD8+ T", "CD8+ memory T", "CD4+ naive T", "CD8+ naive T") ~ "Lymphoid", 
                                                    TRUE ~ as.character(cell_state_mnn_rna_l1)),
                broad_cell_state = factor(broad_cell_state, levels = c("HSC", "GMP", "MEP", "Monocyte", "cDC", "pDC", "preMegakaryocyte", "preB", "Erythroblast", "Erythroid", "Lymphoid"))) %>%
  dplyr::filter(grepl("JD", patient))

healthy_summary <- table(healthy_data_filt$patient, healthy_data_filt$broad_cell_state) %>%
  as.data.frame() %>%
  dplyr::rename(patient = Var1, `Lineage` = Var2, count = Freq) %>%
  dplyr::mutate(sample_treatment = ifelse(grepl("JD", patient), "Frozen", "Fresh RBC-lysed")) %>%
  dplyr::filter(sample_treatment %in% "Frozen",
                !Lineage %in% c("Unknown", "Erythroid", "Lymphoid")) %>%
  dplyr::group_by(patient) %>%
  dplyr::mutate(fraction = count / sum(count),
                patient = gsub("_Healthy", "", patient),
                patient = gsub("Donor_", "H", patient),
                time = "Healthy") %>%
  dplyr::ungroup()


ggplot(healthy_summary, aes(x = time, y = count, fill = `Lineage`)) + 
  geom_bar(position = "fill", stat = "identity", width = 0.7) + 
  xlab("") +
  ylab("Fraction of cells") +
  facet_grid(~time) +
  scale_fill_manual(values = mycols) +
  theme_light() +
  theme(axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14)) 

```

# Heatmaps of Ags
## Ags across blast lineage states
```{r blast_lineage_states_markers}

DefaultAssay(blast_object) <- "ADT"
Idents(blast_object) <- "predicted_id"

mymarkers <- FindAllMarkers(blast_object, assay = "ADT")
top_feats <- mymarkers$gene[abs(mymarkers$avg_log2FC) > 1.5]
DoHeatmap(subset(blast_object, cells = sample(colnames(blast_object), 1000)),
          group.by = "predicted_id",
          features = top_feats,
          assay = "ADT")

```

## Ag change
```{r heatmap_of_ags, fig,width = 8, fig.height = 12}

blast_object_split <- SplitObject(blast_object, split.by = "patient")

markers_remove <- c("ISOTYPE-IGG1K", "ISOTYPE-RAT", "ISOTYPE-ARMENIAN-HAMSTER", "CD34", "CD4", "CD4B", "CD3", "CD8", "CD8A", "CD19", "CD14", "CD15", "CD16", "CD11C", "CD5", "CD34", "CD38", "CD45", "CD9")

markers_remove <- ""
marker_list <- lapply(blast_object_split, function(patient_object) {
  
  Idents(patient_object) <- "time"
  
  if (length(unique(patient_object$time)) == 2) {
    
    markers <- FindMarkers(patient_object, 
                           ident.1 = "Relapse",
                           ident.2 = "Diagnosis",
                           max.cells.per.ident = min(table(patient_object$time)),
                           assay = "ADT", 
                           logfc.threshold = 0, 
                           min.pct = 0, 
                           return.thresh = 1) %>%
      dplyr::arrange(desc(abs(avg_log2FC))) %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene")
  
    return(markers)
  
  } else {
    
    return(NA)
    
  }
  
})

marker_list <- marker_list[!is.na(marker_list)]
de_adt <- lapply(marker_list, function(arg) {
  dplyr::filter(arg, abs(avg_log2FC) > 1, !grepl("ISOTYPE", gene)) %>% 
  dplyr::pull(gene)
})
de_adt <- table(as.character(unlist(de_adt)))
de_adt_patient <- names(de_adt)[de_adt == 1]
de_adt_multi <- names(de_adt)[de_adt > 1]

filt_adt <- rownames(blast_object@assays$ADT)
filt_adt <- filt_adt[!filt_adt %in% markers_remove]
filt_adt <- filt_adt[!grepl("ISOTYPE", filt_adt)]

# pseudobulk <- AverageExpression(blast_object, assay = "ADT", slot = "data", group.by = c("patient", "time"))$ADT
# pseudobulk <- pseudobulk[rownames(pseudobulk) %in% filt_adt, ]
# pseudo_diag <- pseudobulk[, grepl("Diagnosis", colnames(pseudobulk))]
# pseudo_re <- pseudobulk[, grepl("Relapse", colnames(pseudobulk))]
# heatdata <- log1p(as.matrix(pseudo_re)) - log1p(as.matrix(pseudo_diag))
# colnames(heatdata) <- gsub("_Relapse", "", colnames(heatdata))

heatdata <- lapply(marker_list, function(arg) dplyr::arrange(arg) %>%
                     dplyr::select(gene, avg_log2FC) %>%
                     dplyr::filter(!grepl("ISOTYPE", gene)))

heatdata <- lapply(names(heatdata), function(arg) {
  colnames(heatdata[[arg]]) <- c("gene", arg)
  return(heatdata[[arg]])
})

heatdata <- purrr::reduce(heatdata, .f = full_join) %>%
  tibble::column_to_rownames("gene") %>%
  as.matrix()
heatdata[is.na(heatdata)] <- 0

sample_category <- aws.s3::s3read_using(FUN = readr::read_csv,
                                        object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/aml_atlas_sample_category.csv") %>%
  distinct()

clinical_metadata <- aws.s3::s3read_using(
  FUN = readr::read_csv, 
  object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/aml_atlas_sample_metadata_2024-04-03.csv"
) %>%
  dplyr::left_join(sample_category, by = "uhn_patient_id") %>%
  janitor::clean_names() %>%
  dplyr::select(patient, 
                date_of_dx, 
                hsct, 
                dx, 
                initial_tx, 
                response_to_initial_tx, 
                date_of_death_or_last_follow_up, 
                karyotype,
                npm1_2,
                flt3_itd_2,
                flt3_tkd_2,
                runx1_runx1t1,
                #nup98_rearrangement,
                cebpa,
                other_mutations,
                #kmt2_rearrangement,
                mecom_rearrangement) %>%
  dplyr::filter(!grepl("Donor", patient)) %>%
  dplyr::distinct() %>%
  dplyr::mutate(days_survival = as_date(date_of_death_or_last_follow_up) - as_date(date_of_dx),
                hsct = ifelse(hsct %in% "HSCT", "Yes", NA),
                response_to_initial_tx = gsub(" \\(.*|", "", response_to_initial_tx),
                response_to_initial_tx = factor(response_to_initial_tx, 
                                                levels = c("Complete response", "Partial response", "No response"))) %>%
  dplyr::right_join(data.frame(patient = as.character(colnames(heatdata))))

column_ha <- HeatmapAnnotation(
  
  "Days to last follow up" = anno_barplot(clinical_metadata$days_survival),
  "HSC transplant" = clinical_metadata$hsct,
  "Response to therapy" = clinical_metadata$response_to_initial_tx,
  "Initial therapy" = clinical_metadata$initial_tx,
  "Karyotype" = clinical_metadata$karyotype,
  "MECOM rearrangement" = clinical_metadata$mecom_rearrangement,
  "Other mutations" = clinical_metadata$other_mutations,
  "NPM1" = clinical_metadata$npm1_2,
  "FLT3-ITD" = clinical_metadata$flt3_itd_2,
  "FLT3-TKD" = clinical_metadata$flt3_tkd_2,
  "RUNX1-RUNX1T1" = clinical_metadata$runx1_runx1t1,
  #"NUP98 Rearrangement"= clinical_metadata$nup98_rearrangement,
  "CEBPA" = clinical_metadata$cebpa,
  
  col = list(`HSC transplant` = c("Yes" = "black"),
             `Karyotype` = c("Abnormal" = "orange3", "Complex" = "cyan3", "Normal" = "yellowgreen"),
             `Response to therapy` = c("Complete response" = "palegreen4", "Partial response" = "cornflowerblue", "No response" = "sienna"),
             `NPM1` = c("Yes" = "black"),
             `FLT3-ITD` = c("Yes" = "black"),
             `FLT3-TKD` = c("Yes" = "black"),
             `RUNX1-RUNX1T1` = c("Yes" = "black"),
             `CEBPA` = c("Yes" = "black"),
             `Other mutations` = c("Yes" = "black"),
             `MECOM rearrangement` = c("Yes" = "black"))
)

col_split <- factor(ifelse(colnames(heatdata) %in% clinical_metadata$patient[clinical_metadata$response_to_initial_tx %in% c("No response")], "b", "a"))

col_split <- dplyr::case_when(
  colnames(heatdata) %in% "Patient_29" ~ "a",
  colnames(heatdata) %in% clinical_metadata$patient[clinical_metadata$response_to_initial_tx %in% c("No response", "Partial response")] ~ "b",
  TRUE ~ "c") %>%
  as.factor()

col_split <- factor(ifelse(colnames(heatdata) %in% clinical_metadata$patient[clinical_metadata$patient %in% c("Patient_29")], "b", "a"))

tmp_order <- levels(cluster_summary$patient)
tmp_order <- gsub("^P", "Patient_", tmp_order)
tmp_order <- match(tmp_order, colnames(heatdata))

set.seed(42312)
ComplexHeatmap::Heatmap(
  
  heatdata, 
  clustering_method_rows = "ward.D2", 
  clustering_method_columns = "ward.D2",
  cluster_columns = TRUE,
  top_annotation = column_ha,
  column_order = tmp_order,
  column_names_rot = 90,
  column_gap = unit(3, "mm"),
  column_title = NULL,
  #column_km = 2,
  row_dend_side = "left"
)

```

## Ag pseudobulk
```{r ag_pseudobulk, fig.height = 4, fig.width = 13}

blast_object@assays$ADT_dsb_qc_filter@counts <- blast_object@assays$ADT_dsb_qc_filter@data
blast_object@assays$ADT_limma@counts <- blast_object@assays$ADT_limma@data

myassay <- "ADT_dsb_qc_filter"
sample_pb <- AverageExpression(blast_object,
                               layer = "counts",
                               assay = myassay,
                               group.by = "sample_id",
                               return.seurat = TRUE)

# sample_pb <- AggregateExpression(blast_object,
#                                  assays = myassay,
#                                  group.by = "sample_id",
#                                  return.seurat = TRUE)
# 
# sample_pb <- NormalizeData(sample_pb, assay = myassay, normalization.method = "CLR", margin = 2)

sample_pb <- sample_pb %>% 
  dplyr::mutate(patient = gsub("-Diagnosis|-Relapse", "", sample_id),
                patient = gsub("Patient-", "P", patient),
                patient = factor(patient, levels = levels(cluster_summary$patient)))


DefaultAssay(sample_pb) <- myassay
datamat <- LayerData(sample_pb, assay = myassay, layer = "counts")
datamat <- limma::normalizeQuantiles(datamat)
sample_pb[["ADT_qn"]] <- CreateAssayObject(data = datamat)
adt_matrix <- LayerData(sample_pb, features = c("CD33", "CLL1", "EMR2", "CD123"), assay = "ADT_qn", layer = "data")

tmp_order <- sample_pb@meta.data %>%
  dplyr::arrange(patient) %>%
  dplyr::pull(sample_id) %>%
  rev()

adt_matrix <- adt_matrix[, match(tmp_order, colnames(adt_matrix))] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ag") %>%
  dplyr::mutate(assay = "ADT")

# ag density
mypath <- "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/20250331_Antigen Density Matched Blasts Diagnosis vs  Relapse bone marrow.csv"
flow_ag_data <- aws.s3::s3read_using(readr::read_csv, object = mypath)
colnames(flow_ag_data) <- gsub(" ", "-", colnames(flow_ag_data))
flow_ag_data <- flow_ag_data %>%
  dplyr::mutate(patient = paste0("Patient-", patient)) %>%
  tidyr::pivot_longer(-c(patient), names_to = "sample", values_to = "flow_antigen_density") %>%
  tidyr::separate(sample, into = c("ag", "time"), sep = "-") %>%
  tidyr::unite(sample_id, patient, time, sep = "-") %>%
  dplyr::filter(sample_id %in% tmp_order) %>%
  tidyr::pivot_wider(values_from = flow_antigen_density, names_from = sample_id) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("ag")

flow_ag_data <- flow_ag_data %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ag") %>%
  dplyr::mutate(assay = "QuantiBRITE")

# ag positivity
mypath <- "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/20240425_Percent Positivity Matched Blasts Diagnosis vs  Relapse bone marrow .csv"
flow_pos_data <- aws.s3::s3read_using(readr::read_csv, object = mypath)
colnames(flow_pos_data) <- gsub(" ", "-", colnames(flow_pos_data))
flow_pos_data <- flow_pos_data %>%
  dplyr::mutate(patient = paste0("Patient-", patient)) %>%
  tidyr::pivot_longer(-c(patient), names_to = "sample", values_to = "flow_antigen_density") %>%
  tidyr::separate(sample, into = c("ag", "time"), sep = "-") %>%
  tidyr::unite(sample_id, patient, time, sep = "-", remove = FALSE) %>%
  dplyr::filter(sample_id %in% tmp_order)

myorder <- flow_pos_data %>%
  dplyr::filter(ag == "CD33", time == "Relapse") %>%
  dplyr::arrange(desc(flow_antigen_density)) %>%
  dplyr::pull(sample_id) %>%
  str_replace("-Relapse", "")

myorder <- unlist(lapply(myorder, function(arg) c(paste0(arg, "-Diagnosis"), paste0(arg, "-Relapse"))))

flow_pos_data <- flow_pos_data %>%
  dplyr::group_by(ag) %>%
  dplyr::slice(match(myorder, sample_id)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-patient, -time) %>%
  dplyr::mutate(sample_id = factor(sample_id, levels = myorder)) %>%
  tidyr::pivot_wider(values_from = flow_antigen_density, names_from = sample_id) %>%
  dplyr::mutate(assay = "flow_FMO")

# create heatmaps
heatdata <- dplyr::bind_rows(flow_pos_data, adt_matrix, flow_ag_data) %>%
  dplyr::select(assay, ag, everything())

row_ha <- rowAnnotation(
  "Patient" = gsub("-Diagnosis|-Relapse", "", colnames(heatdata)[-c(1, 2)])
)

heatdata_list <- heatdata %>%
  dplyr::group_by(assay) %>%
  dplyr::group_split(.keep = TRUE) 

list_names <- lapply(heatdata_list, function(arg) unique(arg$assay))
heatdata_list <- lapply(heatdata_list, function(arg) arg %>% dplyr::select(-assay) %>% as.data.frame() %>% tibble::column_to_rownames("ag"))
heatdata_list <- lapply(heatdata_list, t)
names(heatdata_list) <- list_names

heatdata_list[["ADT"]] <- heatdata_list[["ADT"]][, c("CD33", "CLL1", "EMR2", "CD123")]
heatdata_list[["QuantiBRITE"]] <- heatdata_list[["QuantiBRITE"]][, c("CD33", "CLL1", "EMR2", "CD123")]
heatdata_list[["QuantiBRITE"]][heatdata_list[["QuantiBRITE"]] > 5000] <- 5000
heatdata_list[["flow_FMO"]] <- heatdata_list[["flow_FMO"]][, c("CD33", "CLL1", "EMR2", "CD123")]

sample_ids <- colnames(heatdata)[-c(1, 2)]
sample_ids <- sample_ids[grepl("Diagnosis", sample_ids)]

patient_split <- gsub("-Diagnosis|-Relapse", "", myorder)
patient_split <- factor(patient_split, levels = unique(patient_split), ordered = TRUE)

heatmaps <- list()
set.seed(42312)
for (assay in c("ADT", "flow_FMO", "QuantiBRITE")) {
  
  input_matrix <- as.matrix(heatdata_list[[assay]]) %>% t()
  rownames(input_matrix) <- case_when(rownames(input_matrix) %in% "CLL1" ~ "CLL-1",
                                      rownames(input_matrix) %in% "EMR2" ~ "ADGRE2",
                                      TRUE ~ rownames(input_matrix))
  colnames(input_matrix) <- gsub("atient-", "", colnames(input_matrix))
  colnames(input_matrix) <- gsub("iagnosis|elapse|-", "", colnames(input_matrix))
  
    col_fun <- colorRamp2(c(min(input_matrix, na.rm = TRUE),
                            quantile(input_matrix, 0.25, na.rm = TRUE), 
                            quantile(input_matrix, 0.75, na.rm = TRUE),
                            max(input_matrix, na.rm = TRUE)), 
                          c("steelblue3","steelblue3", "yellow1","yellow1"))
  
  # convert sample ID names of sample matrix
  tmp_nam <- colnames(input_matrix)
  tmp_nam <- unlist(lapply(tmp_nam, function(arg) {
    new_id <- patient_id_map$patient_manuscript_id[patient_id_map$patient == gsub("D|R", "", arg)]
    new_id <- paste0(new_id, ifelse(grepl("D$", arg), "D", "R"))
    return(new_id)
  }))
  
  colnames(input_matrix) <- tmp_nam
  
  heatmaps[[assay]] <- ComplexHeatmap::Heatmap(
    
    input_matrix,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_gap = unit(2, "mm"),
    col = col_fun,
    column_split = patient_split,
    na_col = "gray96",
    rect_gp = gpar(col = "black", lwd = 0.5),
    heatmap_legend_param = list(
      direction = "horizontal",  # Horizontal color bar
      legend_width = unit(3, "cm"),
      labels_rot = 45
    )
  )

}

heatmaps[["flow_FMO"]] %v% heatmaps[["QuantiBRITE"]] %v% heatmaps[["ADT"]]

```

## Correlation between CITEseq and flow cytometry
```{r correlation_between_assays}

cor_df <- heatdata %>%
  dplyr::filter(assay %in% c("ADT", "QuantiBRITE")) %>%
  tidyr::pivot_longer(cols = -c(assay, ag), names_to = "sample_id", values_to = "expression") %>%
  tidyr::pivot_wider(names_from = "assay", values_from = "expression") %>%
  dplyr::mutate(ag = case_when(ag %in% "CLL1" ~ "CLL-1", 
                               ag %in% "EMR2" ~ "ADGRE2",
                               TRUE ~ as.character(ag)),
                ag = factor(ag, levels = c("CD33", "CLL-1", "CD123", "ADGRE2"))) %>%
  dplyr::filter(!is.na(QuantiBRITE))

ggscatter(data = cor_df,
          x = "ADT",
          y = "QuantiBRITE",
          xlab = "Mean DSB-normalized ADT per blast (CITE-seq)",
          ylab = "Mean antigen number per blast (Quantitative flow)",
          ylim = c(100, 100000),
          facet.by = "ag",
          cor.coef = TRUE,
          cor.method = "spearman",
          add = "reg.line",
          add.params = list(color = "blue"),
          cor.coeff.args = list(label.x = 0, label.y = 4.9)) +
  scale_y_log10(breaks = c(100, 1000, 10000, 100000), labels = c("100", "1,000", "10,000", "100,000"))

```

## Analysis of high-expressing Ags
```{r analysis_high_expressing_antigens}

sample_pb_adt <- AverageExpression(blast_object, group.by = "sample_id", assays = "ADT_dsb_qc_filter", return.seurat = TRUE)
myfeatures <- rownames(sample_pb_adt@assays$ADT_dsb_qc_filter)
myfeatures <- myfeatures[!grepl("ISOTYPE", myfeatures)]
sample_pb_adt <- NormalizeData(sample_pb_adt, features = myfeatures, normalization.method = "RC", margin = 2)

high_ags <- c("CD33", "CD45", "CD99", "CD44", "CD47", "CD49D")
heatdata <- FetchData(sample_pb_adt, vars = high_ags, layer = "data") %>% t()
heatdata <- heatdata[, order(heatdata["CD33", ], decreasing = TRUE)]

# ADT heatmap
column_ha <- HeatmapAnnotation(
  "CD33>CD45" = factor(ifelse(heatdata["CD33", ] > heatdata["CD45", ], "Yes", "No"), levels = c("Yes", "No")),
  col = list(`CD33>CD45` = c("Yes" = "black", "No" = "lightgray"))
)

Heatmap(
  log10(heatdata),
  name = "Relative\nexpression",  # This sets the main heatmap legend title
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  top_annotation = column_ha
)

# RNA heatmap
sample_pb_adt <- AggregateExpression(blast_object, group.by = "sample_id", assays = "RNA", return.seurat = TRUE)
myfeatures <- rownames(sample_pb_adt@assays$ADT)
sample_pb_adt <- NormalizeData(sample_pb_adt, features = myfeatures, normalization.method = "LogNormalize", margin = 2)

high_ags <- c("CD33", "PTPRC", "CD99", "CD44", "CD47", "ITGA4")
heatdata <- FetchData(sample_pb_adt, vars = high_ags, layer = "data") %>% t()
heatdata <- heatdata[, order(heatdata["CD33", ], decreasing = TRUE)]

# heatmap
column_ha <- HeatmapAnnotation(
  "CD33>CD45" = factor(ifelse(heatdata["CD33", ] > heatdata["PTPRC", ], "Yes", "No"), levels = c("Yes", "No")),
  col = list(`CD33>CD45` = c("Yes" = "black", "No" = "lightgray"))
)

Heatmap(
  log10(heatdata),
  name = "Relative\nexpression",  # This sets the main heatmap legend title
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  top_annotation = column_ha
)

# Surface expression of high expressing antigens in blasts compared to lymphocytes
sample_pb_adt <- AggregateExpression(dplyr::filter(aml_data, cell_state_harmony_rna != "DP T cell"), group.by = "cell_state_harmony_rna", assays = "ADT", return.seurat = TRUE)
myfeatures <- rownames(sample_pb_adt@assays$ADT)
myfeatures <- myfeatures[!grepl("ISOTYPE", myfeatures)]
sample_pb_adt <- NormalizeData(sample_pb_adt, features = myfeatures, normalization.method = "RC", margin = 2)

high_ags <- c("CD33", "CD45", "CD99", "CD44", "CD47", "CD49D")
heatdata <- FetchData(sample_pb_adt, vars = high_ags, layer = "data") %>% t()
heatdata <- heatdata[, order(heatdata["CD33", ], decreasing = TRUE)]

Heatmap(
  log10(heatdata),
  name = "Relative\nexpression",  # This sets the main heatmap legend title
  cluster_rows = FALSE,
  cluster_columns = FALSE
)

VlnPlot(dplyr::filter(aml_data, cell_state_harmony_rna != "DP T cell"), 
        group.by = "cell_state_harmony_rna", 
        assay = "ADT",
        features = high_ags,
        pt.size = 0)

```

# Mutation profiles
```{bash variant_annotation, eval = FALSE}

# the following bcltoosl command was used to filter ClinVar vcf file:
docker run -ti -v /data/project_workspace/tmp:/tmp staphb/bcftools:latest bcftools view -W -o /tmp/clinvar_20250115_HP0004808_aml.vcf.gz -O z -i 'INFO/CLNDISDB~"Human_Phenotype_Ontology:HP:0004808" || INFO/SCIDISDB~"Human_Phenotype_Ontology:HP:0004808" || INFO/ONCDN~"Human_Phenotype_Ontology:HP:0004808" || INFO/CLNDISDBINCL~"Human_Phenotype_Ontology:HP:0004808" || INFO/ONCDISDB~"Human_Phenotype_Ontology:HP:0004808"' /tmp/clinvar_20250115.vcf.gz && \
docker run -ti -v /data/project_workspace/tmp:/tmp staphb/bcftools:latest bcftools view -W -o /tmp/clinvar_20250115_HP0004808_AML_nonsynonymous.vcf.gz -O z -e 'INFO/MC~"SO:0001819|synonymous_variant"' /tmp/clinvar_20250115_HP0004808_aml.vcf.gz && \
docker run -ti -v /data/project_workspace/tmp:/tmp pegi3s/picard:latest SortVcf -I /tmp/clinvar_20250115_HP0004808_AML_nonsynonymous.vcf.gz -O /tmp/clinvar_20250115_HP0004808_AML_nonsynonymous_sorted.vcf.gz

# get genomic coordinates for mutations reported by UHN
docker run -v /data/project_workspace/tmp:/ref zhouwanding/transvar:latest \
   transvar panno -l /ref/uhn_ngs_variants.txt -g 2 -t 3 -m 5 --refseq --reference /ref/hg38.fa > uhn_ngs_variants_aa_anno.txt

docker run -v /data/project_workspace/tmp:/ref zhouwanding/transvar:latest \
   transvar canno -l /ref/uhn_ngs_variants.txt -g 2 -t 3 -m 4 --refseq --reference /ref/hg38.fa > uhn_ngs_variants_cdna_anno.txt
   
```

## Summarize with heatmaps
```{r mutation_profiles, fig.height = 5, fig.width = 12}

# load cbioportal and clinvar vcf files with known AML-associated variants
cbio_vcf <- aws.s3::s3read_using(FUN = read.delim,
                                 skip = 211,
                                 object = "s3://vorbio-bioinformatics/genomebrowser/cBioPortal/cBio_liftover_deduplicated_5926loci_6032mutations_manually_modified_header_sorted.vcf") %>%
  dplyr::rename(CHROM = X.CHROM, INFO_cbioportal = INFO) %>%
  dplyr::group_by(CHROM, POS) %>%
  dplyr::mutate(number_mutations_per_gene = n()) %>%
  dplyr::ungroup()

clinvar_vcf_full <- aws.s3::s3read_using(FUN = read.delim,
                                 skip = 41,
                                 object = "s3://vorbio-bioinformatics/genomebrowser/ClinVar/clinvar_20250115.vcf.gz") %>%
  dplyr::rename(CHROM = X.CHROM, INFO_clinvar = INFO)



clinvar_vcf <- aws.s3::s3read_using(FUN = read.delim,
                                 skip = 74,
                                 object = "s3://vorbio-bioinformatics/genomebrowser/ClinVar/clinvar_20250115_HP0004808_AML_nonsynonymous_sorted.vcf.gz") %>%
  dplyr::rename(CHROM = X.CHROM, INFO_clinvar = INFO) %>%
  dplyr::select(CHROM, POS, REF, ALT, ID, INFO_clinvar) %>%
  dplyr::mutate(CHROM = paste0("chr", CHROM)) %>%
  dplyr::filter(grepl("pathogenic", INFO_clinvar, ignore.case = TRUE)) 

stopifnot(all(grepl("Acute_myeloid_leukemia", clinvar_vcf$INFO)))
stopifnot(all(!grepl("SO:0001819|synonymous_variant", clinvar_vcf$INFO)))

# load Archer mutation vcf 
read_archer_vcf <- function(vcf_file_path) {
  
  if (grepl("^s3", vcf_file_path)) {
    myvcf <- aws.s3::s3read_using(FUN = vcfR::read.vcfR, object = vcf_file_path)
  } else {
    myvcf <- vcfR::read.vcfR(vcf_file_path)
  }
  
  myvcf <- myvcf %>% vcfR2tidy(single_frame = TRUE) 
  
  gt <- myvcf$dat %>%
    dplyr::filter(!is.na(gt_GT)) %>%
    dplyr::mutate(gt_FREQ = num(as.numeric(gsub("%", "", gt_FREQ)), label = "%"), gt_PVAL = as.numeric(gt_PVAL)) %>%
    dplyr::select(Indiv, CHROM, POS, gt_GT_alleles, REF, ALT, everything())
  
  return(gt)
  
}

vcf_s3_paths <- c("s3://vorbio-analysis/amplicon_sequencing/241015-JMNNX-00781/2.8.1/VA_CALL/combine.mpileup.varscan.vcf",
                  "s3://vorbio-analysis/amplicon_sequencing/240926-LWMNX-00777/2.8.1/VA_CALL/combine.mpileup.varscan.vcf")

vcf_list <- lapply(vcf_s3_paths, read_archer_vcf)
vcf_com <- do.call("bind_rows", vcf_list) %>%
  dplyr::mutate(Indiv = ifelse(Indiv %in% "AH-AH-Patient-65-6D-20241015_S23", "AH-Patient-65-6R-20241015_S22", Indiv)) %>%
  dplyr::filter(!Indiv %in% c("AH-archer-Patient-47-6D-20240925_S25", 
                              "AH-archer-Patient-47-6R-20240925_S26",
                              "AH-archer-Patient-48-6D-20240925_S27", 
                              "AH-archer-Patient-48-6R-20240925_S28"))


# use VEP to remove uninteresting mutations
vep1 <- aws.s3::s3read_using(FUN = read.delim, skip = 55,
                             object = "s3://vorbio-analysis/amplicon_sequencing/241015-JMNNX-00781/2.8.1/VA_VEP/Output.combine.Vep.raw.txt")
vep2 <- aws.s3::s3read_using(FUN = read.delim, skip = 55,
                             object = "s3://vorbio-analysis/amplicon_sequencing/240926-LWMNX-00777/2.8.1/VA_VEP/Output.combine.Vep.raw.txt")
vep <- dplyr::bind_rows(vep1, vep2) %>%
  janitor::clean_names() %>%
  dplyr::mutate(gt_GT_alleles = gsub(".*_", "", x_uploaded_variation),
                CHROM = gsub("\\:.*", "", location),
                POS = as.integer(gsub(".*\\:|-.*", "", location)),
                vep_insert_id = gsub(".*SYMBOL=|;SYMBOL_SOURCE.*", "", extra),
                vep_insert_id = gsub("-.*", "", vep_insert_id)) %>%
  dplyr::select(vep_insert_id, CHROM, POS, gt_GT_alleles, consequence) %>%
  dplyr::distinct() %>%
  dplyr::group_by(CHROM, POS, gt_GT_alleles) %>%
  dplyr::reframe(vep_insert_id = paste(unique(vep_insert_id), collapse = ", "), 
                 consequence = paste(unique(consequence), collapse = ",")) %>%
  dplyr::mutate(gt_GT_alleles = str_replace(gt_GT_alleles, "^([^/]*/[^/]*)/", "\\1,")) %>%
  tidyr::separate(gt_GT_alleles, into = c("REF", "ALT"), sep = "/", remove = TRUE)

vcf_com <- dplyr::left_join(vcf_com, vep, by = c("CHROM", "POS", "REF", "ALT"))

# load bed files for Archer vcf annotation
bed_file_hg19 <- aws.s3::s3read_using(
  FUN = read.delim,
  header = FALSE,
  object = "s3://vorbio-datascience/genomics/tmp/VariantPlex_MP00187_primers_target-v1.0 (1).bed") %>%
  dplyr::arrange(V1, V2, V3)

bed_file_hg38 <- aws.s3::s3read_using(
  FUN = read.delim, 
  header = FALSE, 
  object = "s3://vorbio-bioinformatics/efs/reference/annotations/hg38/VariantPlex_AML_Custom_Panel_MP00187/VariantPlex_AML_Custom_Panel_MP00187_primers_target-v1.0.sort.hg38.bed")

stopifnot(all(bed_file_hg19$V1 == bed_file_hg38$V1))

bed_file_hg38 <- bed_file_hg38 %>%
  dplyr::mutate(V5 = gsub("_.*", "", bed_file_hg19$V4))

# get gene symbols for SNP IDs
snp_table <- data.frame(id = unique(bed_file_hg38$V5)) %>% dplyr::filter(grepl("^rs",id))
snp_table$gene <- c("PALLD")

ensembl <- useMart("ensembl", host = "https://www.ensembl.org")
ensembl <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
gene_coord <- getBM(attributes = c('hgnc_symbol', 'chromosome_name', 'start_position', 'end_position'),
                    mart = ensembl) %>%
  dplyr::mutate(chromosome_name = paste0("chr", chromosome_name)) %>%
  dplyr::filter(hgnc_symbol != "")

# cross-reference mutations with UHN's NGS analysis
# annotated transcript level ngs file provided by UHN
uhn_vcf <- aws.s3::s3read_using(FUN = read.delim,
                                object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/uhn_ngs_variants.txt")

annot_list <- list()
for (i in 1:nrow(uhn_vcf)) {
  
  cat("\r", i)
  tmp <- uhn_vcf[i, ]
  gene_sym <- tmp$gene
  tx_id <- tmp$nm_transcript_id
  variant_cdna <- tmp$variant_cdna
  variant_aa <- tmp$variant_aa
  
  myinput <- paste(tx_id, variant_cdna, sep = ":")
  mycommand <- glue::glue("transvar canno -i {myinput} --refseq --gseq --reference /home/rstudio/data/project_workspace/tmp/hg38.fa")
  myout <- system(mycommand, intern = TRUE)
  
  if (length(myout) < 2) {
    myinput <- paste(tx_id, variant_aa, sep = ":")
    mycommand <- glue::glue("transvar panno -i {myinput} --refseq --gseq --reference /home/rstudio/data/project_workspace/tmp/hg38.fa")
    myout <- system(mycommand, intern = TRUE)
  }
  
  annot_list[[i]] <- myout[2]

}

uhn_vcf_annot <- lapply(annot_list, function(arg) unlist(strsplit(arg[1], split = "\t")))
uhn_vcf_annot <- do.call("rbind", uhn_vcf_annot) %>% 
  as_tibble %>%
  dplyr::select(V8, V9, V10, V11, V7)
colnames(uhn_vcf_annot) <- c("CHROM", "POS", "REF", "ALT", "INFO_UHN")

uhn_vcf_annot <- cbind(uhn_vcf, uhn_vcf_annot) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(CHROM = ifelse(INFO_UHN == "no_valid_transcript_found" | grepl("Unclassified", INFO_UHN), NA, as.character(CHROM)),
                POS = ifelse(INFO_UHN == "no_valid_transcript_found" | grepl("Unclassified", INFO_UHN), NA, POS),
                REF = ifelse(INFO_UHN == "no_valid_transcript_found" | grepl("Unclassified", INFO_UHN), NA, as.character(REF)),
                ALT = ifelse(INFO_UHN == "no_valid_transcript_found" | grepl("Unclassified", INFO_UHN), NA, as.character(ALT))) %>%
  dplyr::ungroup() %>%
  dplyr::rename(uhn_patient_id = patient_id)

id_map <- aws.s3::s3read_using(read.csv,
                                 object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/aml_atlas_sample_metadata_2024-04-03.csv") %>%
  dplyr::filter(!is.na(uhn_patient_id), grepl("Diagnosis", sample_id)) %>%
  dplyr::select(uhn_patient_id, sample_id) %>%
  dplyr::filter(grepl("Diagnosis", sample_id)) %>%
  dplyr::distinct() %>%
  dplyr::mutate(sample_id = gsub("_", "-", sample_id))

uhn_vcf_annot <- dplyr::right_join(uhn_vcf_annot, id_map, by = c("uhn_patient_id")) %>%
  dplyr::select(sample_id, everything()) %>%
  dplyr::mutate(af = as.numeric(gsub("%", "", af)),
                POS = as.integer(POS))

# extract PCR mutation validation data from metadata
metadata <- aws.s3::s3read_using(read.csv,
                                 object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/aml_atlas_sample_metadata_2024-04-03.csv") %>%
  dplyr::mutate(sample_id = gsub("_", "-", sample_id))

no_ngs_samples <- metadata$sample_id[is.na(metadata$other_mutations_ngs)]
no_ngs_samples <- gsub("_", "-", no_ngs_samples)

metadata <- metadata %>%
  dplyr::select(sample_id, npm1, flt3_itd, flt3_tkd, kit, runx1, tp53) %>%
  tidyr::pivot_longer(-c("sample_id"), names_to = "gene_pcr", values_to = "status") %>%
  dplyr::mutate(gene_pcr = toupper(gene_pcr),
                gene_pcr = gsub("_itd|_tkd", "", gene_pcr, ignore.case = TRUE))

pcr_mutations <- dplyr::select(metadata, sample_id, gene_pcr, status) %>%
  dplyr::filter(!grepl("undetectable|inconclusive", status, ignore.case = TRUE), !is.na(status)) %>%
  distinct() %>%
  dplyr::rename(gene = gene_pcr) %>%
  dplyr::select(-status) %>%
  dplyr::mutate(validation = "PCR") %>%
  dplyr::left_join(id_map, by = c("sample_id"))

uhn_vcf_annot <- dplyr::full_join(uhn_vcf_annot, pcr_mutations, by = c("sample_id", "gene", "uhn_patient_id"))

# annotate Archer mutation vcf file
blank_record <- tibble::tibble(CHROM = NA, 
                               POS = NA, REF = NA, 
                               ALT_cbioportal = NA, 
                               INFO_cbioportal = NA, 
                               ALT_clinvar = NA, 
                               INFO_clinvar = NA, 
                               POS_cbioportal = NA, 
                               REF_cbioportal = NA, 
                               POS_clinvar = NA, 
                               REF_clinvar = NA)

insert_id <- rep(NA, nrow(vcf_com))
in_cbio <- rep(NA, nrow(vcf_com))
db_records <- as.list(rep(NA, nrow(vcf_com)))
indel_sequence_flank_length <- 5
for (i in 1:nrow(vcf_com)) {
  
  cat("\r", i)
  chr <- vcf_com$CHROM[i]
  pos <- vcf_com$POS[i]
  insert_name <- bed_file_hg38$V5[bed_file_hg38$V1 == chr & bed_file_hg38$V2 <= pos & bed_file_hg38$V3 >= pos]
  gene_name <- gene_coord$hgnc_symbol[gene_coord$chromosome_name == chr & gene_coord$start_position <= pos & gene_coord$end_position >= pos] %>% unique()
  
  if (length(gene_name) == 1) {
    insert_id[i] <- gene_name
  } else {
    insert_id[i] <- insert_name
  }
  
  ref <- vcf_com$REF[i]
  alt <- vcf_com$ALT[i]
  alt <- unlist(strsplit(alt, split = ","))
  
  cbioportal_rec <- dplyr::filter(cbio_vcf, CHROM == chr, POS == pos, REF == ref, ALT %in% alt) %>%
    dplyr::select(-ID, -QUAL, -FILTER) %>%
    dplyr::group_by(CHROM, POS, REF) %>%
    dplyr::reframe(ALT = paste(ALT, collapse = ","), INFO_cbioportal = paste(INFO_cbioportal, collapse = " / "))
  
  clinvar_rec <- dplyr::filter(clinvar_vcf, CHROM == chr, POS == pos, REF == ref, ALT %in% alt) %>%
    dplyr::group_by(CHROM, POS, REF) %>%
    dplyr::reframe(ALT = paste(ALT, collapse = ","), INFO_clinvar = paste(INFO_clinvar, collapse = " / "))
  
  db_record <- dplyr::full_join(cbioportal_rec, clinvar_rec, by = c("CHROM", "POS", "REF"), suffix = c("_cbioportal", "_clinvar"))
  stopifnot(nrow(db_record) < 2)
  # check if allele is indel
  indel_present <- (!(all(lapply(alt, nchar) == 1))) | (nchar(ref) > 1)
  
  if (nrow(db_record) == 0 & indel_present) {
    
    cbioportal_rec <- dplyr::filter(cbio_vcf, 
                                    CHROM == chr, 
                                    (POS < (pos + indel_sequence_flank_length)) & (POS > (pos - indel_sequence_flank_length)), 
                                    grepl("TERM\\=INS|TERM\\=DEL", INFO_cbioportal)) %>%
      dplyr::select(-ID, -QUAL, -FILTER) %>%

      dplyr::reframe(CHROM = unique(CHROM), 
                     POS = paste(unique(POS), collapse = ","), 
                     REF = paste(unique(REF), collapse = ","), 
                     ALT = paste(unique(ALT), collapse = ","), 
                     INFO_cbioportal = paste(INFO_cbioportal, collapse = " / "))
    
    clinvar_rec <- dplyr::filter(clinvar_vcf, 
                                 CHROM == chr, 
                                 (POS < (pos + indel_sequence_flank_length)) & (POS > (pos - indel_sequence_flank_length)), 
                                 grepl("CLNVC=Deletion|CLNVC=Duplication|CLNVC=Indel|CLNVC=Insertion|CLNVC=Microsatellite", INFO_clinvar)) %>%
      dplyr::reframe(CHROM = unique(CHROM),
                     POS = paste(unique(POS), collapse = ","), 
                     REF = paste(unique(REF), collapse = ","), 
                     ALT = paste(unique(ALT), collapse = ","), 
                     INFO_clinvar = paste(INFO_clinvar, collapse = " / "))
    
    db_record_indel <- dplyr::full_join(cbioportal_rec, clinvar_rec, by = c("CHROM"), suffix = c("_cbioportal", "_clinvar"))
    stopifnot(nrow(db_record_indel) < 2)
    
    if (nrow(db_record_indel) > 0) {
      db_records[[i]] <- db_record_indel
    } else {
      db_records[[i]] <- blank_record
    }
    
  } else if (nrow(db_record) == 0) {
      db_records[[i]] <- blank_record 
  } else {
    db_records[[i]] <- db_record
  }

}

db_records_df <- do.call("bind_rows", db_records) %>%
  dplyr::mutate(insert_id = insert_id) %>%
  dplyr::select(insert_id, POS_cbioportal, REF_cbioportal, ALT_cbioportal, INFO_cbioportal,
                POS_clinvar, REF_clinvar, ALT_clinvar, INFO_clinvar)

vcf_com_annot <- vcf_com %>%
  dplyr::bind_cols(db_records_df) %>%
  dplyr::mutate(sample_id = gsub("AH-|AH-archer-", "", Indiv),
                sample_id = gsub("-2024.*", "", sample_id),
                sample_id = gsub("6D|1D", "Diagnosis", sample_id),
                sample_id = gsub("6R|1R", "Relapse", sample_id))

# some records may not have VEP annotation due to difference in indel
vcf_com_annot_na <- dplyr::filter(vcf_com_annot, is.na(consequence))






flagged_records <- vcf_com_annot %>%
  fuzzyjoin::fuzzy_right_join(uhn_vcf_annot,
                              by = c("sample_id", "insert_id" = "gene", "CHROM", "POS"),
                              match_fun = list(`==`, `==`, `==`, function(x, y) abs(x - y) <= indel_sequence_flank_length)) %>%
  dplyr::mutate(alt_match_db = case_when(!is.na(INFO_cbioportal) & !is.na(INFO_clinvar) ~ "all",
                                         !is.na(INFO_cbioportal) ~ "cbioportal",
                                         !is.na(INFO_clinvar) ~ "clinvar")) %>%
  dplyr::filter((is.na(sample_id.x) & !is.na(sample_id.y)),
                !is.na(variant_cdna))

# annotate Archer vcf with UHN NGS data after conversion with transvar
sample_mut <- vcf_com_annot %>%
  fuzzyjoin::fuzzy_left_join(uhn_vcf_annot,
                              by = c("sample_id", "insert_id" = "gene", "CHROM", "POS"),
                              match_fun = list(`==`, `==`, `==`, function(x, y) abs(x - y) <= indel_sequence_flank_length)) %>%
  dplyr::mutate(alt_match_db = case_when(!is.na(INFO_cbioportal) & !is.na(INFO_clinvar) ~ "all",
                                         !is.na(INFO_cbioportal) ~ "cbioportal",
                                         !is.na(INFO_clinvar) ~ "clinvar")) %>%
  dplyr::select(Indiv, insert_id, gt_FREQ, gt_PVAL, alt_match_db, consequence) %>%
  dplyr::filter(!is.na(alt_match_db)) %>%
  dplyr::filter(grepl("missense_variant|splice_region_variant|inframe_insertion|stop_gained|frameshift_variant|inframe_deletion|splice_polypyrimidine_tract_variant", consequence) | is.na(consequence)) %>%
  dplyr::group_by(Indiv, insert_id) %>% 
  dplyr::slice(which.min(gt_PVAL)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Indiv = gsub("AH-|-2024.*|archer-", "", Indiv),
                Indiv = gsub("6D|1D", "Diagnosis", Indiv),
                Indiv = gsub("6R|1R", "Relapse", Indiv),
                time = ifelse(grepl("Relapse", Indiv), "R", "D"),
                patient = gsub("-Relapse|-Diagnosis", "", Indiv),
                patient = gsub("Patient-", "P", patient)) %>%
  dplyr::left_join(patient_id_map, by = "patient") %>%
  dplyr::filter(!is.na(patient_manuscript_id)) %>%
  dplyr::arrange(patient_manuscript_id) %>%
  dplyr::mutate(sample_id = paste(patient_manuscript_id, time, sep = "")) %>%
  dplyr::select(sample_id, insert_id, gt_FREQ)


# plot VAF heatmap
sample_mut <- sample_mut %>%
  tidyr::pivot_wider(values_from = gt_FREQ, names_from = insert_id, values_fill = 0) %>%
  dplyr::select(sample_id, everything()) %>%
  #dplyr::bind_rows(data.frame(sample_id = c("P60D", "P60R"))) %>%
  tibble::column_to_rownames("sample_id") %>%
  t()


rsplit <- gsub("D$|R$", "", colnames(sample_mut))
rsplit <- factor(rsplit, levels = unique(rsplit))

sample_mut <- sample_mut[order(rowSums(sample_mut > 5, na.rm = TRUE), decreasing = TRUE), ]
sample_mut <- sample_mut[rowSums(sample_mut > 1, na.rm = TRUE) > 1, ]

heatdata <- sample_mut          
ComplexHeatmap::Heatmap(
  heatdata, 
  cluster_rows = FALSE, 
  cluster_columns = FALSE, 
  column_split = rsplit, 
  column_gap = unit(2, "mm"),
  col = c("white", "cornflowerblue", "red"),
  rect_gp = gpar(col = "black", lwd = 0.5),
  row_names_side = "left",
  na_col = "lightgray"
)

```

## Correlate mutations with cell state composition in blasts
```{r mutations_cell_state, fig.width=6, fig.height=5}

mut_df <- sample_mut %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sample_manuscript_id") %>%
  dplyr::mutate(time = ifelse(grepl("D$", sample_manuscript_id), "Dx", "Re")) %>%
  dplyr::mutate(patient_manuscript_id = gsub("D$|R$", "", sample_manuscript_id)) %>%
  dplyr::left_join(cluster_summary, by = c("patient_manuscript_id", "time")) %>%
  dplyr::group_by(sample_manuscript_id) %>%
  dplyr::mutate(shannon_index = vegan::diversity(count, index = "shannon"),
                simpson_index = vegan::diversity(count, index = "simpson")) %>%
  dplyr::ungroup()

# tumor suppressors and oncogenic signaling
plotdf <- mut_df %>%
  dplyr::select(sample_manuscript_id, SRSF2,  NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::mutate(mut_gp = case_when((TP53 > 1 | TET2 > 1 | WT1 > 1) & (KIT > 1 | FLT3 > 1) ~ "Tumor suppressor and\noncogenic signaling",
                                   TP53 > 1 | TET2 > 1 | WT1 > 1 ~ "Tumor suppressor\n(TP53, TET2, WT1)", 
                                   KIT > 1 | FLT3 > 1 ~ "Oncogenic signaling\n(FLT3, KIT)",
                                   TRUE ~ "Other mutations"),
                mut_gp = factor(mut_gp, levels = c("Tumor suppressor\n(TP53, TET2, WT1)",
                                                   "Oncogenic signaling\n(FLT3, KIT)",
                                                   "Tumor suppressor and\noncogenic signaling",
                                                   "Other mutations")),
                time = ifelse(grepl("D$", sample_manuscript_id), "Diagnosis", "Relapse")) %>%
  dplyr::filter(!is.na(mut_gp))

comparisons <- list(c("Tumor suppressor\n(TP53, TET2, WT1)", "Other mutations"))

xplot <- ggboxplot(dplyr::filter(plotdf, time == "Diagnosis"), 
          x = "mut_gp", 
          y = "shannon_index",
          ylim = c(0, 2.1),
          xlab = "",
          ylab = "Blast state diversity (Shannon index)",
          fill = "mut_gp",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(mapping = aes(label = paste0("p = ", ..p.format..)), comparisons = comparisons, method = "t.test") +  # Show p-values for significance
  theme_bw(base_size = 14) +  # Apply a clean minimal theme
  theme(legend.position = "none")

comparisons <- list(c("Tumor suppressor\n(TP53, TET2, WT1)", "Other mutations"), 
                    c("Oncogenic signaling\n(FLT3, KIT)", "Other mutations"))

yplot <- ggboxplot(dplyr::filter(plotdf, time == "Relapse"), 
          x = "mut_gp", 
          y = "shannon_index",
          ylim = c(0, 2.1),
          xlab = "",
          ylab = "",
          fill = "mut_gp",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(mapping = aes(label = paste0("p = ", ..p.format..)), comparisons = comparisons, method = "t.test") +  # Show p-values for significance
  theme_bw(base_size = 14) +  # Apply a clean minimal theme
    guides(fill = guide_legend(title = "")) +
theme(
    legend.position = "top",               # Place legend at the top
    legend.justification = "right",         # Center the legend
    legend.box = "horizontal",             # Arrange legend items horizontally
    legend.key.size = unit(1, "cm"),
  )
  
  
xplot + yplot

# Oncogenic signaling
plotdf <- mut_df %>%
  dplyr::select(sample_manuscript_id, SRSF2,  NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::mutate(mut_gp = factor(ifelse(KIT > 1 | FLT3 > 1, "Signal transduction\n(KIT, FLT3)", "Other"), levels = c("Signal transduction\n(KIT, FLT3)", "Other"))) %>%
  dplyr::filter(!is.na(mut_gp))

ggboxplot(plotdf, 
          x = "mut_gp", 
          y = "shannon_index",
          xlab = "",
          ylab = "Blast state diversity (Shannon index)",
          fill = "mut_gp",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          shape = 21,       # Hollow circles
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(label.x = 0.7, label.y = 0.1, mapping = aes(label = paste0("p = ", ..p.format..))) +  # Show p-values for significance
  theme_classic(base_size = 14) +  # Apply a clean minimal theme
  theme(legend.position = "none")

# NPM1
plotdf <- mut_df %>%
  dplyr::select(sample_manuscript_id, SRSF2,  NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::mutate(mut_gp = factor(ifelse(NPM1 > 1, "NPM1", "Other"), levels = c("NPM1", "Other"))) %>%
  dplyr::filter(!is.na(mut_gp))

ggboxplot(plotdf, 
          x = "mut_gp", 
          y = "shannon_index",
          xlab = "",
          ylab = "Clonality (Shannon index)",
          fill = "mut_gp",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          shape = 21,       # Hollow circles
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(label.x = 0.7, label.y = 0.1, mapping = aes(label = paste0("p = ", ..p.format..))) +  # Show p-values for significance
  theme_classic(base_size = 14) +  # Apply a clean minimal theme
  theme(legend.position = "none")

# Transcription factors
plotdf <- mut_df %>%
  dplyr::select(sample_manuscript_id, SRSF2,  NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::mutate(mut_gp = factor(ifelse(RUNX1 > 1 | CEBPA > 1, "TFs", "Other"), levels = c("TFs", "Other"))) %>%
  dplyr::filter(!is.na(mut_gp))

ggboxplot(plotdf, 
          x = "mut_gp", 
          y = "shannon_index",
          xlab = "",
          ylab = "Clonality (Shannon index)",
          fill = "mut_gp",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          shape = 21,       # Hollow circles
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(label.x = 0.7, label.y = 0.1, mapping = aes(label = paste0("p = ", ..p.format..))) +  # Show p-values for significance
  theme_classic(base_size = 14) +  # Apply a clean minimal theme
  theme(legend.position = "none")

# Splicing
plotdf <- mut_df %>%
  dplyr::select(sample_manuscript_id, SRSF2,  NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::mutate(mut_gp = factor(ifelse(SRSF2 > 1 | SF3B1 > 1, "Splicing", "Other"), levels = c("Splicing", "Other"))) %>%
  dplyr::filter(!is.na(mut_gp))

ggboxplot(plotdf, 
          x = "mut_gp", 
          y = "shannon_index",
          xlab = "",
          ylab = "Clonality (Shannon diversity index)",
          fill = "mut_gp",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          shape = 21,       # Hollow circles
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(label.x = 0.7, label.y = 0.1, mapping = aes(label = paste0("p = ", ..p.format..))) +  # Show p-values for significance
  theme_classic(base_size = 14) +  # Apply a clean minimal theme
  theme(legend.position = "none")

# Epigenetic modifiers
plotdf <- mut_df %>%
  dplyr::select(sample_manuscript_id, SRSF2,  NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::mutate(mut_gp = factor(ifelse(DNMT3A > 1 | IDH1 > 1 | IDH2 > 1, "Epigenetic\nmodifiers", "Other"), levels = c("Epigenetic\nmodifiers", "Other"))) %>%
  dplyr::filter(!is.na(mut_gp))

ggboxplot(plotdf, 
          x = "mut_gp", 
          y = "shannon_index",
          xlab = "",
          ylab = "Clonality (Shannon diversity index)",
          fill = "mut_gp",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          shape = 21,       # Hollow circles
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(label.x = 0.7, label.y = 0.1, mapping = aes(label = paste0("p = ", ..p.format..))) +  # Show p-values for significance
  theme_classic(base_size = 14) +  # Apply a clean minimal theme
  theme(legend.position = "none")

```

## Correlate mutations with surface antigens
```{r mutations_cell_state, fig.width=6, fig.height=5}

markers_remove <- c("ISOTYPE-IGG1K", "ISOTYPE-RAT", "ISOTYPE-ARMENIAN-HAMSTER", "CD4", "CD4B", "CD3", "CD8", "CD8A", "CD19", "CD5", "CD71", "CD235A", "CD22", "CD3")

plotdf <- mut_df %>%
  dplyr::select(sample_id, SRSF2,  NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA) %>%
  dplyr::distinct()

ref_samples <- plotdf$sample_id[plotdf$DNMT3A > 1 | plotdf$IDH1 > 1 | plotdf$IDH2 > 1]
ref_samples <- plotdf$sample_id[plotdf$TP53 > 1]
ref_samples <- ref_samples[!is.na(ref_samples)]

hsc_object <- blast_object %>%
  dplyr::mutate(de_group = ifelse(sample_manuscript_id %in% ref_samples, "mutant", "wt"))

mymarkers <- rownames(hsc_object@assays$ADT)
mymarkers <- mymarkers[!grepl("isotype", mymarkers, ignore.case = TRUE)]
hsc_pb <- AggregateExpression(hsc_object, features = mymarkers, group.by = "sample_manuscript_id", assay = "ADT", return.seurat = TRUE)
hsc_pb <- NormalizeData(hsc_pb, method = "CLR", margin = 2, assay = "ADT")
hsc_pb <- hsc_pb %>% dplyr::mutate(de_group = ifelse(sample_manuscript_id %in% ref_samples, "mutant", "wt")) 

DefaultAssay(hsc_pb) = "ADT"
Idents(hsc_pb) = "de_group"
toptab <- FindMarkers(hsc_pb,
                      ident.1 = "mutant",
                      ident.2 = "wt",
                      test.use = "wilcox_limma") %>%
  dplyr::arrange(p_val)

VlnPlot(hsc_pb, group.by = "de_group", features = rownames(toptab)[toptab$p_val_adj < 0.1])


# Correlation matrix
mat1 <- sample_mut %>%
  t()
mat2 <- FetchData(hsc_pb, assay = "ADT", layer = "data", vars = rownames(hsc_pb@assays$ADT))[rownames(mat1), ]
cor_mat <- cor(mat1, mat2, method = "spearman")










vaf_df <- plotdf %>% as.data.frame() %>% tibble::column_to_rownames("sample_id")
adt_expr <- as.data.frame(t(hsc_pb@assays$ADT$data)) 
samp_int <- intersect(rownames(vaf_df), rownames(adt_expr))
samp_int <- sort(samp_int)
adt_expr <- adt_expr[samp_int, !colnames(adt_expr) %in% markers_remove]
vaf_df <- vaf_df[samp_int, ]
test_res <- NULL
for (variant in colnames(vaf_df)) {
  for (ag in colnames(adt_expr)) {
    variant_profile <- vaf_df[, variant]
    variant_profile <- ifelse(variant_profile > 1, 1, 0)
    variant_profile <- factor(variant_profile, levels = c(1, 0))
    tres <- t.test(adt_expr[, ag] ~ variant_profile)
    test_res[[paste(variant, ag, sep = "_")]] <- data.frame(variant = variant, 
                                                            ag = ag, statistic = tres$statistic, 
                                                            p_value = tres$p.value)
  }
}

test_res <- dplyr::bind_rows(test_res) %>% 
  dplyr::arrange(p_value) %>%
  dplyr::mutate(adj_p = p.adjust(.$p_value, method = "BH")) %>%
  dplyr::filter(adj_p < 0.05)

res_mat <- test_res %>%
  dplyr::select(variant, ag, statistic) %>%
  tidyr::pivot_wider(names_from = ag, values_from = statistic, values_fill = 0) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("variant")

res_mat[res_mat > 0] <- 1
res_mat[res_mat < 0] <- -1
pheatmap(res_mat, color = c("blue", "white", "red"))



# Look into LSC fraction

ref_samples <- plotdf$sample_id[plotdf$DNMT3A > 1]
ref_samples <- ref_samples[!is.na(ref_samples)]
tmpdf <- blast_object@meta.data %>%
  dplyr::select(sample_id, barcode, lsc, predicted_id) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::summarize(lsc_fraction = sum(lsc == "LSC") / n()) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(mutant_group = ifelse(sample_id %in% gsub("-", "_", ref_samples), "mutant", "wt"))

ggboxplot(tmpdf, 
          x = "mutant_group", 
          y = "lsc_fraction",
          fill = "mutant_group",  # Color by group
          palette = "uchicago",  # Use a clean color palette
          add = "jitter",   # Add jitter points
          shape = 21,       # Hollow circles
          add.params = list(alpha = 0.5, size = 1.8, width = 0.2)) +  # Adjust jitter aesthetics
  stat_compare_means(label.x = 0.7, label.y = 0.1, mapping = aes(label = paste0("p = ", ..p.format..))) +  # Show p-values for significance
  theme_classic(base_size = 14) +  # Apply a clean minimal theme
  theme(legend.position = "none")


```




# UMAP plots of selected patients with compositional changes
```{r global_umap}

blast_object <- dplyr::left_join(blast_object, 
                                 dplyr::mutate(patient_id_map,
                                               patient = gsub("^P", "Patient_", patient), 
                                               by = "patient"))
DefaultAssay(blast_object) <- "RNA"
blast_object <- FindVariableFeatures(blast_object, nfeatures = 1000)
blast_object <- ScaleData(blast_object, vars.to.regress = "patient_manuscript_id")
blast_object <- RunPCA(blast_object, seed.use = 3142)
blast_object <- RunUMAP(blast_object, dims = 1:30, seed.use = 45751, repulsion = 3, n.neighbors = 40, spread = 3)


DimPlot(blast_object, reduction = "umap", group.by = "patient_manuscript_id", label = TRUE, repel = TRUE, alpha = 0.2) + NoLegend() + NoAxes() + ggtitle(NULL)

DimPlot(blast_object, reduction = "umap", group.by = "time", alpha = 0.2, cols = c("#0073C2FF", "#EFC000FF"), shuffle = TRUE) + NoLegend() + NoAxes() + ggtitle(NULL)




blast_object <- blast_object %>%
  dplyr::mutate(broad_cell_state = dplyr::case_when(predicted_id %in% c("NK", 
                                                                        "NKT", 
                                                                        "B cell", 
                                                                        "CD4+ T", 
                                                                        "CD4+ memory T", 
                                                                        "CD8+ T", 
                                                                        "CD8+ memory T", 
                                                                        "CD4+ naive T", 
                                                                        "CD8+ naive T") ~ "Lymphoid", 
                                                    TRUE ~ as.character(predicted_id)),
                broad_cell_state = factor(broad_cell_state, levels = c("HSC", 
                                                                       "GMP", 
                                                                       "MEP", 
                                                                       "Monocyte", 
                                                                       "cDC", 
                                                                       "pDC", 
                                                                       "preMegakaryocyte", 
                                                                       "preB", 
                                                                       "Erythroblast", 
                                                                       "Erythroid", 
                                                                       "Lymphoid")))

DimPlot(filter(blast_object, patient_manuscript_id == "P01"), reduction = "umap", group.by = "broad_cell_state", alpha = 0.2, cols = mycols) + 
  NoLegend() + NoAxes() + ggtitle(NULL)

```

```{r feature_plot_selected_patients}

color_map <- data.frame(cell_states, colors = mycols[1:length(cell_states)])

pat58 <- dplyr::filter(blast_object, patient == "Patient_58") 
DefaultAssay(pat58) <- "RNA"

pat58 <- pat58 %>%
  FindVariableFeatures() %>%
  NormalizeData() %>%
  ScaleData()

pat58 <- RunPCA(pat58)
pat58 <- RunUMAP(pat58, dims = 1:50)
DimPlot(dplyr::filter(pat58, predicted_id %in% c("HSC", "MEP", "GMP", "Monocyte", "cDC", "pDC")) %>% 
          dplyr::mutate(predicted_id = factor(predicted_id, levels = c("HSC", "GMP", "MEP", "Monocyte", "cDC", "pDC"))), 
        split.by = "time", 
        group.by = "predicted_id", 
        reduction = "umap", 
        cols = mycols,
        ncol = 1) + NoAxes() + 
  theme(strip.text.x = element_blank())


pat44 <- dplyr::filter(blast_object, patient == "Patient_44") 
DefaultAssay(pat44) <- "RNA"

pat44 <- pat44 %>%
  FindVariableFeatures() %>%
  NormalizeData() %>%
  ScaleData()

pat44 <- RunPCA(pat44)
pat44 <- RunUMAP(pat44, dims = 1:50)

DimPlot(dplyr::filter(pat44, predicted_id %in% c("HSC", "MEP", "GMP", "Monocyte", "cDC", "pDC")) %>% 
          dplyr::mutate(predicted_id = factor(predicted_id, levels = c("HSC", "GMP", "MEP", "Monocyte", "cDC", "pDC"))), 
        split.by = "time", 
        group.by = "predicted_id", 
        reduction = "umap", 
        cols = mycols,
        ncol = 1) + NoAxes() + 
  theme(strip.text.x = element_blank())

pat70 <- dplyr::filter(blast_object, patient == "Patient_70") 
DefaultAssay(pat70) <- "RNA"

pat70 <- pat70 %>%
  FindVariableFeatures() %>%
  NormalizeData() %>%
  ScaleData()

pat70 <- RunPCA(pat70)
pat70 <- RunUMAP(pat70, dims = 1:50)

DimPlot(dplyr::filter(pat70, predicted_id %in% c("HSC", "MEP", "GMP", "Monocyte", "cDC", "pDC")) %>% 
          dplyr::mutate(predicted_id = factor(predicted_id, levels = c("HSC", "GMP", "MEP", "Monocyte", "cDC", "pDC"))), 
        split.by = "time", 
        group.by = "predicted_id", 
        reduction = "umap", 
        cols = mycols,
        ncol = 1) + NoAxes() + 
  theme(strip.text.x = element_blank())



blast_object_filt <- dplyr::filter(blast_object, patient %in% c("Patient_44", "Patient_58", "Patient_61", "Patient_70", "Patient_43"),
                                   predicted_id %in% c("HSC", "MEP", "GMP", "PreB", "Erythroblast", "Monocyte", "cDC", "pDC"))
blast_object_filt <- SplitObject(blast_object_filt, split.by = "patient")
mycols <- c("firebrick1", "seagreen3", "lightcyan3", "cyan", "purple", "hotpink", "gold1", "mediumturquoise", "yellow2", "sienna3", "springgreen", "olivedrab", "orange1",  "cornflowerblue", "rosybrown1", "maroon2", "slateblue2", "tan2", "peachpuff1", "orchid1", "yellowgreen", "burlywood2")

object_list <- lapply(blast_object_filt, function(arg) {
  
  arg <- arg %>%
    FindVariableFeatures() %>%
    NormalizeData() %>%
    ScaleData() %>%
    RunPCA() %>%
    RunUMAP(dims = 1:50) %>%
    dplyr::mutate(predicted_id = factor(predicted_id, levels = c("HSC", "GMP", "MEP", "PreB", "Erythroblast", "Monocyte", "cDC", "pDC")))
  
})
  
DimPlot(object_list[["Patient_44"]], 
        group.by = "predicted_id", 
        cols = mycols,
        reduction = "umap") + NoAxes() + NoLegend() + ggtitle(NULL)

DimPlot(object_list[["Patient_58"]], 
        group.by = "predicted_id", 
        cols = mycols,
        reduction = "umap") + NoAxes() + NoLegend() + ggtitle(NULL)

DimPlot(object_list[["Patient_61"]], 
        group.by = "predicted_id", 
        cols = mycols,
        reduction = "umap") + NoAxes() + NoLegend() + ggtitle(NULL)

DimPlot(object_list[["Patient_70"]], 
        group.by = "predicted_id", 
        cols = mycols,
        reduction = "umap") + NoAxes() + NoLegend() + ggtitle(NULL)

DimPlot(object_list[["Patient_43"]], 
        group.by = "predicted_id", 
        cols = mycols,
        reduction = "umap") + NoAxes() + NoLegend() + ggtitle(NULL)



DimPlot(arg, 
        group.by = "time", 
        cols = mycols,
        reduction = "umap") + NoAxes()
DimPlot(arg, 
        group.by = "predicted_id", 
        cols = mycols,
        reduction = "umap") + NoAxes()


```

# Survival analysis
## Compositional shift using Atchison's distance
```{r survival_analysis_composition_shift}

patient_age <- aws.s3::s3read_using(FUN = readr::read_csv,
                                    object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/patient_age_at_dx.csv")

comp_data <- dplyr::filter(cluster_summary) %>%
  dplyr::select(-pct) %>%
  dplyr::distinct() %>%
  dplyr::select(patient, time, broad_cell_state, count, sex)

comp_data_dx <- dplyr::filter(comp_data, time == "Dx") %>%
  dplyr::select(-time)
comp_data_re <- dplyr::filter(comp_data, time == "Re") %>%
  dplyr::select(-time)

# calculate atchison's distance between timepoints; if a particular cell type is absent from group, add 1 pseudocell to enable calculation
pseudocount <- 10
comp_df <- dplyr::full_join(comp_data_dx, comp_data_re, by = c("patient", "broad_cell_state", "sex"), suffix = c("_dx", "_re")) %>%
  dplyr::mutate(count_dx = ifelse(is.na(count_dx), 0, count_dx) + pseudocount,
                count_re = ifelse(is.na(count_re), 0, count_re) + pseudocount) %>%
  dplyr::group_by(patient) %>%
  dplyr::mutate(adist = as.numeric(vegan::vegdist(t(as.matrix(cbind(count_dx, count_re))), method = "aitchison"))) %>%
  dplyr::ungroup() 

surv_data <- dplyr::select(comp_df, patient, sex, adist) %>%
  dplyr::distinct() %>%
  dplyr::left_join(dplyr::select(cluster_summary, patient, patient_manuscript_id, time_dx_to_re, time_dx_to_last_follow_up, time_re_to_last_follow_up, vital_status, relapse_status) %>% distinct(), 
                   by = "patient") %>%
  dplyr::left_join(patient_age, by = "patient_manuscript_id") %>%
  dplyr::mutate(vital_status = ifelse(vital_status == "Dead", 1, 0),
                relapse_status = ifelse(relapse_status == "Relapsed", 1, 0),
                adist_binary = ifelse(adist > mean(adist), "Hi", "Low"),
                patient = as.character(patient),
                sex = factor(sex, levels = c("male", "female")))

clone_index <- mut_df %>%
  dplyr::select(sample_id, SRSF2, NPM1, IDH1, ASXL1, FLT3, IDH2, KIT, NRAS, DNMT3A, TP53, TET2, SF3B1, RUNX1, WT1, STAG2, JAK2, CEBPA, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::filter(grepl("D$", sample_id)) %>%
  dplyr::mutate(patient = gsub("-Diagnosis|-Relapse", "", sample_id),
                patient = gsub("Patient-", "P", patient),
                shannon_index_binary = ifelse(shannon_index > 0.1, "high", "low")) %>%
  dplyr::select(-sample_id)

surv_data <- surv_data %>%
  dplyr::left_join(clone_index, by = "patient") %>%
  as.data.frame() %>%
  dplyr::mutate(ts = ifelse(TP53 > 1 | TET2 > 1 | WT1 > 1, "yes", "no"),
                signaling = ifelse(FLT3 > 1 | KIT > 1, "yes", "no"),
                epigenetic = ifelse(DNMT3A > 1 | IDH2 > 1, "yes", "no"))

mysurv <- survival::Surv(time = surv_data$time_dx_to_last_follow_up, event = surv_data$vital_status)
mysurv_re_to_event <- survival::Surv(time = surv_data$time_re_to_last_follow_up, event = surv_data$vital_status)
mysurv_dx_to_re <- survival::Surv(time = surv_data$time_dx_to_re, event = surv_data$relapse_status)

mycox <- coxph(mysurv ~ adist + sex + age_at_aml_dx, data = surv_data)
summary(mycox)
cox.zph(mycox)
ggforest(model = mycox, data = surv_data)

mycox <- coxph(mysurv_re_to_event ~ adist + sex + age_at_aml_dx, data = surv_data)
summary(mycox)
cox.zph(mycox)
ggforest(model = mycox, data = surv_data)

mycox <- coxph(mysurv ~ shannon_index + sex, data = surv_data)
ggforest(model = mycox, data = surv_data)

mycox <- coxph(mysurv ~ simpson_index + sex, data = surv_data)
ggforest(model = mycox, data = surv_data)

mycox <- coxph(mysurv ~ epigenetic + sex, data = surv_data)
ggforest(model = mycox, data = surv_data)



# generate kaplan-meier plot
myfit <- survfit(mysurv ~ adist_binary, data = surv_data)
myplot <- survminer::ggsurvplot(fit = myfit, 
                                data = surv_data, 
                                #surv.median.line = "h", 
                                palette = "npg", 
                                pval = TRUE, 
                                pval.coord = c(3700, 1)) 
myplot$plot <- myplot$plot +
  xlab("Days after relapse") +
  ylab("Overall survival") +
  scale_x_continuous(breaks = c(0, 500, 1000, 1500, 2000, 5000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
myplot

# plot Aitchison's distnace
ggbarplot(data = surv_data,
          x = "patient_manuscript_id",
          y = "adist",
          xlab = "",
          ylab = "Aitchison D",
          fill = "gray",
          color = "white",
          width = 0.5) +
    theme(axis.text.y = element_text(size = 15),
        axis.title.y = element_text(size = 15))

# plot Shannon index
diversity_df <- mut_df %>%
  dplyr::select(sample_id, shannon_index, simpson_index) %>%
  dplyr::distinct() %>%
  dplyr::mutate(time = ifelse(grepl("Diagnosis", sample_id), "D", "R"),
                patient = gsub("Patient-", "P", sample_id),
                patient = gsub("-Diagnosis|-Relapse", "", patient)) %>%
  dplyr::select(-sample_id) %>%
  dplyr::left_join(patient_id_map, by = "patient")

ggbarplot(data = diversity_df,
          x = "time",
          y = "shannon_index",
          xlab = "",
          ylab = "Shannon index",
          fill = "time",
          palette = c("gray", "gray"),
          color = "white",
          width = 0.7) +
  facet_grid(~patient_manuscript_id) +
  theme(axis.text.y = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        strip.text = element_blank())

```

# Compare compositional shift with clonal stability
```{r mut_vs_comp, eval = FALSE}

mutation_shift <- sample_mut[, !grepl("-65-|-60-", colnames(sample_mut_cat))]
mutation_shift_dx <- mutation_shift[, grepl("Diagnosis", colnames(mutation_shift))]
mutation_shift_re <- mutation_shift[, grepl("Relapse", colnames(mutation_shift))]
stopifnot(all(gsub("-Diagnosis", "", colnames(mutation_shift_dx)) == gsub("-Relapse", "", colnames(mutation_shift_re))))

mutation_ed <- NULL
mypats <- gsub("-Diagnosis", "", colnames(mutation_shift_dx))
for (pat in mypats) {

  x <- mutation_shift_dx[, grepl(paste0(pat, "-"), colnames(mutation_shift_dx))]
  y <- mutation_shift_re[, grepl(paste0(pat, "-"), colnames(mutation_shift_re))]
  num_change <- ifelse(all(x == y), "stable", "unstable")
  mutation_ed[[pat]] <- data.frame(patient = gsub("Patient-", "P", pat), mutation_distance = num_change)
  
}
mutation_ed <- do.call("rbind", mutation_ed)

surv_data_mut <- dplyr::inner_join(surv_data, mutation_ed, by = c("patient")) %>%
  dplyr::mutate(chr_stability = ifelse(patient %in% c("P61", "P29", "P30", "P50", "P28", "P49", "P52", "P43"), "unstable", "stable")) %>%
  dplyr::mutate(overall_stability = ifelse(mutation_distance == "stable" & chr_stability == "stable", "stable", "unstable"))

ggscatter(data = surv_data_mut,
          x = "adist",
          y = "mutation_distance",
          label = "patient",
          cor.coef = TRUE)

ggboxplot(data = surv_data_mut,
          x = "mutation_distance",
          y = "adist",
          label = "patient") +
  stat_compare_means()

ggboxplot(data = surv_data_mut,
          x = "mutation_distance",
          y = "adist",
          add = "jitter") +
  stat_compare_means(method.args = list(alternative = "less"))






mutation_shift <- sample_mut[, !grepl("-65-|-60-", colnames(sample_mut))]
mutation_shift_dx <- mutation_shift[, grepl("Diagnosis", colnames(mutation_shift))]
mutation_shift_re <- mutation_shift[, grepl("Relapse", colnames(mutation_shift))]
stopifnot(all(gsub("-Diagnosis", "", colnames(mutation_shift_dx)) == gsub("-Relapse", "", colnames(mutation_shift_re))))

mutation_ed <- NULL
mypats <- gsub("-Diagnosis", "", colnames(mutation_shift_dx))
for (pat in mypats) {

  x <- mutation_shift_dx[, grepl(paste0(pat, "-"), colnames(mutation_shift_dx))]
  y <- mutation_shift_re[, grepl(paste0(pat, "-"), colnames(mutation_shift_re))]
  ed <- 1 - cor(x, y, method = "pearson")
  mutation_ed[[pat]] <- data.frame(patient = gsub("Patient-", "P", pat), mutation_distance = ed)
  
}
mutation_ed <- do.call("rbind", mutation_ed)

surv_data_mut <- dplyr::inner_join(surv_data, mutation_ed, by = c("patient")) %>%
  dplyr::mutate(chr_stability = ifelse(patient %in% c("P61", "P29", "P30", "P50", "P28", "P49", "P52", "P43"), "unstable", "stable")) %>%
  dplyr::mutate(hsct = ifelse(patient %in% c("P26", "P27", "P29", "P30", "P28", "P4", "P31"), "yes", "no")) %>%
  dplyr::slice(match(gsub("atient-|-Diagnosis|-Relapse", "", myorder), patient))

ggscatter(data = surv_data_mut,
          x = "adist",
          y = "mutation_distance",
          xlab = "Aitchison distance",
          ylab = "Pearson distance",
          label = "patient",
          color = "hsct",
          repel = TRUE,
          cor.coef = TRUE,
          #facet.by = "hsct",
          size = 2) +
  geom_smooth(method = "lm", lty = 2, lwd = 0.45)

bardata <- data.frame(patient = surv_data_mut$patient, value = surv_data_mut$adist, distance = "Aitchison distance")
bardata <- rbind(bardata, data.frame(patient = surv_data_mut$patient, value = surv_data_mut$mutation_distance, distance = "Pearson distance"))
bardata <- dplyr::mutate(bardata, patient = factor(patient, levels = patient_order))

ggplot(dplyr::filter(bardata, distance == "Aitchison distance"), aes(x = patient, y = value, fill = distance)) + 
  geom_bar(stat = "identity", width = 0.5, fill = "azure4") +
  xlab("") +
  ylab("Aitchison D") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(dplyr::filter(bardata, distance == "Pearson distance"), aes(x = patient, y = value, fill = distance)) + 
  geom_bar(stat = "identity", position = "identity", width = 0.45) +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 

```


## Broad cell states
```{r broad_cell_states}

surv_data_ref <- comp_data %>%
  janitor::clean_names() %>%
  dplyr::group_by(patient, sex, time) %>%
  dplyr::mutate(percent = count / sum(count) * 100) %>%
  dplyr::ungroup() %>%
  dplyr::select(-count) %>%
  tidyr::pivot_wider(values_from = percent, names_from = time, values_fill = 0) %>%
  dplyr::mutate(blast_cell_state_l2 = case_when(blast_cell_state %in% c("HSC", "MEP", "GMP") ~ "HSPC",
                                                grepl("T$", blast_cell_state) ~ "T cell",
                                                blast_cell_state %in% c("preB", "B cell") ~ "B cell",
                                                blast_cell_state %in% c("Monocyte", "DC") ~ "Differentiated myeloid"))

surv_data <- surv_data_ref %>%
  dplyr::select(-blast_cell_state) %>%
  dplyr::group_by(patient, sex, blast_cell_state_l2) %>%
  dplyr::summarize(Dx = sum(Dx), Re = sum(Re)) %>%
  dplyr::ungroup()

surv_data <- surv_data %>%
  dplyr::mutate(percent_change_from_dx = Dx - Re) %>%
  dplyr::left_join(dplyr::select(cluster_summary, patient, time_dx_to_last_follow_up, time_dx_to_re, vital_status) %>% distinct(), by = "patient") %>%
  dplyr::mutate(vital_status = ifelse(vital_status == "Dead", 1, 0)) %>%
  as.data.frame() 

# HSPC
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state_l2 %in% c("HSPC"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)
ggforest(mycox)

# T cell
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state_l2 %in% c("T cell"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)
ggforest(mycox)

# B cell
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state_l2 %in% c("B cell"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)
ggforest(mycox)

# Differentiated myeloid
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state_l2 %in% c("Differentiated myeloid"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)
ggforest(mycox)

```

## Specific cell states
```{r specific_cell_states}

surv_data <- surv_data_ref %>%
  dplyr::select(-blast_cell_state_l2) %>%
  dplyr::group_by(patient, sex, blast_cell_state) %>%
  dplyr::summarize(Dx = sum(Dx), Re = sum(Re)) %>%
  dplyr::ungroup()

surv_data <- surv_data %>%
  dplyr::mutate(percent_change_from_dx = Dx - Re) %>%
  dplyr::left_join(dplyr::select(cluster_summary, patient, time_dx_to_last_follow_up, time_dx_to_re, vital_status) %>% distinct(), by = "patient") %>%
  dplyr::mutate(vital_status = ifelse(vital_status == "Dead", 1, 0)) %>%
  as.data.frame() 

# CD4+ naive T
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state %in% c("CD4+ naive T"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)

# CD4+ memory T
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state %in% c("CD4+ memory T"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)
ggforest(mycox)

# CD8+ naive T 
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state %in% c("CD8+ naive T"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)
ggforest(mycox)

# CD8+ memory T
tmp_surv_data <- dplyr::filter(surv_data, blast_cell_state %in% c("CD8+ memory T"))
mysurv <- survival::Surv(time = tmp_surv_data$time_dx_to_last_follow_up, event = tmp_surv_data$vital_status)
mycox <- coxph(mysurv ~ Dx + Re + sex, data = tmp_surv_data)
summary(mycox)

```

# Cluster all modalities together
```{r cluster_all_modalities}

clust_data <- sample_mut
tmp <- comp_data %>% 
  dplyr::select(patient, time, broad_cell_state, count) %>%
  dplyr::mutate(patient = gsub("P", "Patient-", patient),
                time = ifelse(time == "Dx", "Diagnosis", "Relapse")) %>%
  tidyr::unite(sample_id, patient, time, sep = "-") %>%
  tidyr::pivot_wider(names_from = sample_id, values_from = count, values_fill = 0) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("broad_cell_state")
tmp <- tmp[, colnames(clust_data)]

tmp2 <- do.call("rbind", heatdata_list)
tmp2[is.na(tmp2)] <- 0
tmp2 <- t(tmp2)
tmp2 <- tmp2[, colnames(clust_data)]
rownames(tmp2) <- paste0("ag-", rownames(tmp2))
clust_data <- rbind(clust_data, tmp)
clust_data <- rbind(clust_data, tmp2)

clust_data <- scale(t(clust_data))
pheatmap(cor(t(clust_data), use = "pairwise.complete.obs"))

```

# Pseudobulk differential expression analysis and enrichment between relapse and relapse groups
## Load gene sets from MSigDB
```{r gene_sets}

all_gene_sets = msigdbr(species = "Homo sapiens") %>%
  dplyr::filter(gs_cat %in% "H")

pgs_list <- all_gene_sets %>%
  dplyr::group_by(gs_name) %>%
  dplyr::group_split()

list_names <- lapply(pgs_list, function(arg) unique(arg$gs_name)) %>% unlist()
pgs_list <- lapply(pgs_list, function(arg) arg$human_gene_symbol)
names(pgs_list) <- gsub("HALLMARK_|GOBP_", "", list_names)


all_gene_sets = msigdbr(species = "Homo sapiens") %>%
  dplyr::filter(gs_cat %in% "C1")

pos_gs_list <- all_gene_sets %>%
  dplyr::group_by(gs_name) %>%
  dplyr::group_split()

list_names <- lapply(pos_gs_list, function(arg) unique(arg$gs_name)) %>% unlist()
pos_gs_list <- lapply(pos_gs_list, function(arg) arg$human_gene_symbol)
names(pos_gs_list) <- list_names

all_gene_sets <- msigdbr(species = "Homo sapiens") %>%
  dplyr::filter(gs_cat %in% "C7", 
                gs_subcat %in% "IMMUNESIGDB", 
                grepl("IL1|IL6|IL8|IL12|IFNG|TNFA", gs_description) | grepl("GSE5099_CLASSICAL_M1_VS_ALTERNATIVE_M2_MACROPHAGE", gs_name),
                !grepl("MICROGLIA", gs_description))

immune_gs_list <- all_gene_sets %>%
  dplyr::group_by(gs_name) %>%
  dplyr::group_split()

list_names <- lapply(immune_gs_list, function(arg) unique(arg$gs_name)) %>% unlist()
immune_gs_list <- lapply(immune_gs_list, function(arg) arg$human_gene_symbol)
names(immune_gs_list) <- list_names



# senmayo <- msigdbr(species = "Homo sapiens") %>%
#   dplyr::filter(gs_cat %in% "C2", gs_subcat %in% "CGP")
#   dplyr::filter(gs_name == "SAUL_SEN_MAYO")
# 
#   
# senmayo <- list(senmayo = c("ACVR1B","ANG","ANGPT1","ANGPTL4","AREG","AXL","BEX3","BMP2","BMP6","C3","CCL1","CCL13","CCL16","CCL2","CCL20","CCL24","CCL26","CCL3","","CCL4","CCL5","CCL7","CCL8","CD55","CD9","CSF1","CSF2","CSF2RB","CST4","CTNNB1","CTSB","CXCL1","CXCL10","CXCL12","CXCL16","CXCL2","CXCL3","CXCL8","CXCR2","DKK1","EDN1","EGF","EGFR","EREG","ESM1","ETS2","FAS","FGF1","FGF2","FGF7","GDF15","GEM","GMFG","HGF","HMGB1","ICAM1","ICAM3","IGF1","IGFBP1","IGFBP2","IGFBP3","IGFBP4","IGFBP5","IGFBP6","IGFBP7","IL10","IL13","IL15","IL18","IL1A","IL1B","IL2","IL32","IL6","IL6ST","IL7","INHA","IQGAP2","ITGA2","ITPKA","JUN","KITLG","LCP1","MIF","MMP1","MMP10","MMP12","MMP13","MMP14","MMP2","MMP3","MMP9","NAP1L4","NRG1","PAPPA","PECAM1","PGF","PIGF","PLAT","PLAU","PLAUR","PTBP1","PTGER2","PTGES","RPS6KA5","SCAMP4","SELPLG","SEMA3F","SERPINB4","SERPINE1","SERPINE2","SPP1","SPX","TIMP2","TNF","TNFRSF10C","TNFRSF11B","TNFRSF1A","TNFRSF1B","TUBGCP2","VEGFA","VEGFC","VGF","WNT16","WNT2"))

```

## All blasts
```{r blasts_whole}

pseudobulk <- edgeR::Seurat2PB(dplyr::filter(blast_object, cell_state_harmony_rna != "Unknown"), cluster = "time", sample = "sample_id")
pseudobulk$samples$patient <- gsub("_relapse|_Relapse", "", pseudobulk$samples$sample)
keep.genes <- filterByExpr(pseudobulk, min.count = 10, min.total.count = 100)
pseudobulk <- pseudobulk[keep.genes, , keep = FALSE]
pseudobulk <- normLibSizes(pseudobulk)

cluster <- as.factor(pseudobulk$samples$cluster)
patient <- as.factor(pseudobulk$samples$patient)

design <- model.matrix(~ 0 + cluster + patient)
con <- makeContrasts(clusterRelapse - clusterrelapse, levels = design)
pseudobulk <- estimateDisp(pseudobulk, design, robust = TRUE)
fit <- glmQLFit(pseudobulk, design, robust = TRUE)
qlf <- glmQLFTest(fit, contrast = con)

top_table_edger <- qlf$table %>%
  dplyr::mutate(padj = p.adjust(PValue, method = "BH")) %>%
  dplyr::arrange(PValue)

EnhancedVolcano(top_table_edger,
                lab = top_table_edger %>% rownames,
                x = 'logFC',
                y = 'PValue',
                FCcutoff = 0.5,
                pCutoff = 0.05)


mystats <- top_table_edger$logFC
names(mystats) <- rownames(top_table_edger)
mystats <- mystats[is.finite(mystats)]
es_res <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 10000) %>%
  dplyr::arrange(pval)

es_res


```


## Blasts subset on different cell states
```{r de_and_enrichment}


pseudobulk <- AggregateExpression(dplyr::filter(blast_object, predicted_id %in% c("HSC", "GMP", "MEP", "Monocyte", "cDC", "preB")), 
                                  features = rownames(blast_object@assays$RNA), 
                                  assays = "RNA", 
                                  return.seurat = TRUE, 
                                  group.by = c("sample_manuscript_id", "predicted_id"))

pseudobulk <- pseudobulk %>%
  dplyr::mutate(time = substr(sample_manuscript_id, 4, 4),
                patient = substr(sample_manuscript_id, 1, 3))
                                  
pseudobulk <- SplitObject(pseudobulk, split.by = "predicted_id")

# remove lowly expressed genes - features must have at least 10 counts in at last 50% of samples
pseudobulk_filt <- NULL
for(tmp_object in pseudobulk) {
  
  features <- rownames(tmp_object@assays$RNA)[rowSums(tmp_object@assays$RNA@layers$counts > 10) > (0.5 * ncol(tmp_object))]
  if (length(features) < 10) {
    next
  } else {
    tmp_object <- subset(tmp_object, features = features)
    stopifnot(nrow(tmp_object) == length(features))
    pseudobulk_filt[[unique(tmp_object$predicted_id)]] <- tmp_object
  }
  
}

limma_res <- NULL
for (tmp_object in pseudobulk_filt) {

   # paired analysis requires addition of patient ID to model formula
   # there are some patients where a cell state is in one time point but not the other
   design_matrix <- model.matrix(~time + patient, data = tmp_object@meta.data)

   v <- limma::voom(counts = as.matrix(LayerData(tmp_object, layer = "count", assay = "RNA")),
                    design = design_matrix,
                    plot = TRUE,
                    normalize = "quantile")
    
   fit <- limma::lmFit(v, design = design_matrix)
   fit <- limma::eBayes(fit)
   top_table <- limma::topTable(fit, number = Inf, coef = "timeR", sort.by = "P")
   top_table <- dplyr::mutate(top_table, blast_lineage = unique(tmp_object$predicted_id))

  limma_res[[unique(tmp_object$predicted_id)]] <- top_table

}

```

## T cells
```{r de_and_enrichment_tcells}

tcell_object <- dplyr::filter(aml_data, 
                              cell_state_harmony_rna %in% c("CD8 T cell"))
                                
pseudobulk <- edgeR::Seurat2PB(tcell_object, cluster = "time", sample = "sample_id")
pseudobulk$samples$patient <- gsub("_relapse|_Relapse", "", pseudobulk$samples$sample)
keep.genes <- filterByExpr(pseudobulk, min.count = 10, min.total.count = 100)
pseudobulk <- pseudobulk[keep.genes, , keep = FALSE]
pseudobulk <- normLibSizes(pseudobulk)

cluster <- as.factor(pseudobulk$samples$cluster)
patient <- as.factor(pseudobulk$samples$patient)

design <- model.matrix(~ 0 + cluster + patient)
con <- makeContrasts(clusterRelapse - clusterrelapse, levels = design)
pseudobulk <- estimateDisp(pseudobulk, design, robust = TRUE)
fit <- glmQLFit(pseudobulk, design, robust = TRUE)
qlf <- glmQLFTest(fit, contrast = con)

top_table_edger <- qlf$table %>%
  dplyr::mutate(padj = p.adjust(PValue, method = "BH")) %>%
  dplyr::arrange(PValue)

EnhancedVolcano(top_table_edger,
                lab = top_table_edger %>% rownames,
                x = 'logFC',
                y = 'PValue',
                FCcutoff = 0.5,
                pCutoff = 0.05)


mystats <- top_table_edger$logFC
names(mystats) <- rownames(top_table_edger)
mystats <- mystats[is.finite(mystats)]
es_res <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 10000) %>%
  dplyr::arrange(pval)

es_res


```

# GSVA on sample peudobulk profiles
## Blasts
```{r ssgsea_blast_pseudobulk}

pseudobulk <- AggregateExpression(dplyr::filter(blast_object, !grepl("T$|NK|B cell|Erythroid|Unknown", predicted_id)), 
                                  assays = c("RNA"), 
                                  group.by = c("sample_manuscript_id", "predicted_id"))

pseudobulk_rna <- pseudobulk$RNA

myparam <- GSVA::gsvaParam(exprData = pseudobulk_rna, 
                           geneSets = pgs_list, 
                           kcdf = "Poisson", 
                           minSize = 5, 
                           sparse = TRUE)

hallmark_es <- GSVA::gsva(myparam)

# pathways_to_include <- c("INFLAMMATORY_RESPONSE",
#                          "TNFA_SIGNALING_VIA_NFKB",
#                          "INTERFERON_ALPHA_RESPONSE",
#                          "IL6_JAK_STAT3_SIGNALING",
#                          #"INTERFERON_GAMMA_REPONSE",
#                          "IL2_STAT5_SIGNALING",
#                          "COMPLEMENT",
#                          "ALLOGRAFT_REJECTION",
#                          "E2F_TARGETS",
#                          "MYC_TARGETS_V1",
#                          "MYC_TARGETS_V2",
#                          "G2M_CHECKPOINT",
#                          "OXIDATIVE_PHOSPHORYLATION",
#                          "MYC_TARGETS_V2",
#                          "DNA_REPAIR")
# 
# hallmark_es_filt <- hallmark_es_filt[pathways_to_include, ]
hallmark_es_relapse <- hallmark_es[, grepl("D_", colnames(hallmark_es))]
hallmark_es_relapse <- hallmark_es[, grepl("R_", colnames(hallmark_es))]

pheatmap(hallmark_es_relapse)
pheatmap(hallmark_es_relapse)
pheatmap(hallmark_es)

# relapse
patient <- gsub("D_.*", "", colnames(hallmark_es_relapse))
cell_state <- gsub(".*_", "", colnames(hallmark_es_relapse))

patient_colors <- pals::glasbey()
names(patient_colors) <- unique(aml_data$patient_manuscript_id)
patient_colors <- patient_colors[!is.na(names(patient_colors))]

lineage_colors <- c("firebrick1", "lightcyan3", "seagreen3", "cyan", "purple", "hotpink", "gold1", "mediumturquoise", "yellow2", "sienna3", "springgreen", "olivedrab", "orange1",  "cornflowerblue", "rosybrown1", "maroon2", "slateblue2", "tan2")
names(lineage_colors) <- levels(aml_data$predicted_id)
lineage_colors <- lineage_colors[names(lineage_colors) %in% cell_state]
ha <- HeatmapAnnotation(
    "Patient" = patient,
    "Lineage state" = factor(cell_state, levels = unique(cell_state)),
    col = list(`Patient` = patient_colors, `Lineage state` = lineage_colors),
    na_col = "lightgray",
    annotation_name_side = "left"
)

hallmark_es_relapse <- hallmark_es_relapse[order(rowMeans(hallmark_es_relapse), decreasing = TRUE), ]
mysplit <- kmeans(hallmark_es_relapse, 3)$cluster

myheatmap <- ComplexHeatmap::Heatmap(
  hallmark_es_relapse,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  clustering_method_rows = "ward.D2", 
  clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  #row_split = mysplit,
  column_split = patient,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1.5, "mm"),
  row_gap = unit(1.5, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es), 
    gp = gpar(fontsize = 12)
  )
)

draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")


# relapse
patient <- gsub("R_.*", "", colnames(hallmark_es_relapse))
cell_state <- gsub(".*_", "", colnames(hallmark_es_relapse))

ha <- HeatmapAnnotation(
    "Patient" = patient,
    "Lineage state" = factor(cell_state, levels = unique(cell_state)),
    col = list(`Patient` = patient_colors, `Lineage state` = lineage_colors),
    na_col = "lightgray",
    annotation_name_side = "left"
)

hallmark_es_relapse <- hallmark_es_relapse[order(rowMeans(hallmark_es_relapse), decreasing = TRUE), ]
#mysplit <- kmeans(hallmark_es_relapse, 3)$cluster

myheatmap <- ComplexHeatmap::Heatmap(
  hallmark_es_relapse,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  clustering_method_rows = "ward.D2", 
  clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  #row_split = mysplit,
  column_split = patient,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1.5, "mm"),
  row_gap = unit(1.5, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es), 
    gp = gpar(fontsize = 12)
  )
)

draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")

```

## Blasts small molecule drug signatures
```{r ssgsea_blast_pseudobulk_small_molecule_signatures}

pseudobulk <- AggregateExpression(dplyr::filter(blast_object, !grepl("T$|NK|B cell|Erythroid|Unknown", predicted_id)), 
                                  assays = c("RNA"), 
                                  group.by = c("sample_manuscript_id", "predicted_id"))

pseudobulk_rna <- pseudobulk$RNA

pseudobulk_adt <- AggregateExpression(dplyr::filter(blast_object, !grepl("T$|NK|B cell|Erythroid|Unknown", predicted_id)), 
                                      assays = c("ADT"), 
                                      group.by = c("sample_manuscript_id", "predicted_id"),
                                      return.seurat = TRUE)

pseudobulk_adt <- NormalizeData(pseudobulk_adt$ADT, normalization.method = "CLR", margin = 2)
pseudobulk_adt <- FetchData(pseudobulk_adt, vars = c("CD33", "CLL1", "CD305", "CD49D", "CD205", "CD244"), assay = "ADT", layer = "data")
pseudobulk_adt <- t(pseudobulk_adt)
pseudobulk_adt_relapse <- pseudobulk_adt[, grepl("D_", colnames(pseudobulk_adt))]
pseudobulk_adt_relapse <- pseudobulk_adt[, grepl("R_", colnames(pseudobulk_adt))]

# Run GSVA on RNA pseudobulk using MERGE drug signatures
drug_sigs <- aws.s3::s3read_using(FUN = readr::read_csv,
                                  col_names = FALSE,
                                  object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/gene_signatures/drug_signatures.csv")

drug_sigs <- drug_sigs %>%
  dplyr::group_split(X2)
drug_sigs_names <- lapply(drug_sigs, function(arg) unique(dplyr::pull(arg, X2)))
drug_sigs <- lapply(drug_sigs, function(arg) dplyr::pull(arg, X1))
names(drug_sigs) <- drug_sigs_names
                  
myparam <- GSVA::gsvaParam(exprData = pseudobulk_rna, 
                           geneSets = drug_sigs, 
                           kcdf = "Poisson", 
                           minSize = 5, 
                           sparse = TRUE)

hallmark_es <- GSVA::gsva(myparam)
hallmark_es_relapse <- hallmark_es[, grepl("D_", colnames(hallmark_es))]
hallmark_es_relapse <- hallmark_es[, grepl("R_", colnames(hallmark_es))]

# all samples
patient <- gsub("D_.*|R_.*", "", colnames(hallmark_es))
cell_state <- gsub(".*_", "", colnames(hallmark_es))
timepoint <- unlist(lapply(colnames(hallmark_es), function(arg) substr(arg, 4, 4)))

ha <- HeatmapAnnotation(
  "Patient" = patient,
  "Time" = timepoint,
  "Lineage state" = cell_state,
    na_col = "lightgray",
    annotation_name_side = "left"
)

#mysplit <- kmeans(hallmark_es_relapse, 3)$cluster

myheatmap <- ComplexHeatmap::Heatmap(
  hallmark_es,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  rect_gp = gpar(col = "white", lwd = 2),
  clustering_method_rows = "ward.D2", 
  clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  #row_split = mysplit,
  column_split = patient,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1.5, "mm"),
  row_gap = unit(1.5, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es), 
    gp = gpar(fontsize = 12)
  )
)


draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")

# relapse
patient <- gsub("D_.*|R_.*", "", colnames(hallmark_es_relapse))
cell_state <- gsub(".*_", "", colnames(hallmark_es_relapse))

patient_colors <- pals::glasbey()
names(patient_colors) <- unique(aml_data$patient_manuscript_id)
patient_colors <- patient_colors[!is.na(names(patient_colors))]

lineage_colors <- c("firebrick1", "lightcyan3", "seagreen3", "cyan", "purple", "hotpink", "gold1", "mediumturquoise", "yellow2", "sienna3", "springgreen", "olivedrab", "orange1",  "cornflowerblue", "rosybrown1", "maroon2", "slateblue2", "tan2")
names(lineage_colors) <- levels(aml_data$predicted_id)
lineage_colors <- lineage_colors[names(lineage_colors) %in% cell_state]
ha <- HeatmapAnnotation(
    "Patient" = patient,
    "Lineage state" = factor(cell_state, levels = unique(cell_state)),
    col = list(`Patient` = patient_colors, `Lineage state` = lineage_colors),
    na_col = "lightgray",
    annotation_name_side = "left"
)
set.seed(4602)
mysplit <- kmeans(hallmark_es_relapse, 3)$cluster
myheatmap <- ComplexHeatmap::Heatmap(
  hallmark_es_relapse,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  clustering_method_rows = "ward.D2", 
  clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  row_split = mysplit,
  column_split = patient,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1, "mm"),
  row_gap = unit(1, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es), 
    gp = gpar(fontsize = 12)
  )
)
draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")


pseudobulk_adt_relapse <- pseudobulk_adt[, grepl("D_", colnames(pseudobulk_adt))]
ag_heatmap <- ComplexHeatmap::Heatmap(
  pseudobulk_adt_relapse,
  col = c("blue", "white", "yellow"),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  column_split = patient,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1, "mm"),
  row_gap = unit(1, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es), 
    gp = gpar(fontsize = 12)
  )
)

draw(ag_heatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")



# relapse
patient <- gsub("D_.*|R_.*", "", colnames(hallmark_es_relapse))
cell_state <- gsub(".*_", "", colnames(hallmark_es_relapse))

ha <- HeatmapAnnotation(
    "Patient" = patient,
    "Lineage state" = factor(cell_state, levels = unique(cell_state)),
    col = list(`Patient` = patient_colors, `Lineage state` = lineage_colors),
    na_col = "lightgray",
    annotation_name_side = "left"
)

mysplit <- kmeans(hallmark_es_relapse, 3)$cluster

set.seed(4602)
myheatmap <- ComplexHeatmap::Heatmap(
  hallmark_es_relapse,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  row_split = mysplit,
  column_split = patient,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1., "mm"),
  row_gap = unit(1, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es), 
    gp = gpar(fontsize = 12)
  )
)

draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")

pseudobulk_adt_relapse <- pseudobulk_adt[, grepl("R_", colnames(pseudobulk_adt))]
ag_heatmap <- ComplexHeatmap::Heatmap(
  pseudobulk_adt_relapse,
  col = c("blue", "white", "yellow"),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  column_split = patient,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1, "mm"),
  row_gap = unit(1, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es), 
    gp = gpar(fontsize = 12)
  )
)

draw(ag_heatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")


```


## Microenvironment
```{r microenvironment}

me_aml <- dplyr::filter(aml_data, grepl("T$|B cell", predicted_id))

pseudobulk <- AggregateExpression(me_aml,
                                  assays = c("RNA"), 
                                  group.by = c("sample_id", "predicted_id"))

pseudobulk_rna <- pseudobulk$RNA

myparam <- GSVA::gsvaParam(exprData = pseudobulk_rna, 
                           geneSets = pgs_list, 
                           kcdf = "Poisson", 
                           minSize = 5, 
                           sparse = TRUE)

hallmark_es_me <- GSVA::gsva(myparam)
hallmark_es_me <- hallmark_es_me %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("pb_id") %>%
  tidyr::separate(col = pb_id, into = c("sample_id", "cell_state"), sep = "_", remove = FALSE) %>%
  dplyr::mutate(time = ifelse(grepl("Diagnosis", sample_id), "Diagnosis", "Relapse"),
                patient = gsub("-Diagnosis|-Relapse", "", sample_id)) %>%
  dplyr::arrange(cell_state, time, patient) %>%
  dplyr::select(-sample_id, -cell_state, -time, -patient) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("pb_id") %>%
  t() %>%
  as.data.frame()



# all samples
patient <- gsub("_.*", "", colnames(hallmark_es_me))
cell_state <- gsub(".*_", "", colnames(hallmark_es_me))
timepoint <- gsub(".*-|_.*", "", colnames(hallmark_es_me))
cell_state_timepoint <- paste(cell_state, timepoint, sep = "_")

hsct_samples <- unique(aml_data$sample_id[aml_data$hsct == "HSCT"])
hsct_samples <- gsub("_", "-", hsct_samples)
hsct <- ifelse(gsub("_.*", "", colnames(hallmark_es_me)) %in% hsct_samples, "hsct", "none")

ha <- HeatmapAnnotation(
  "Lineage state" = cell_state,
  "Time" = timepoint,
  "Patient" = patient,
  "HSCT" = hsct,
    # col = list(`Response to chemotherapy` = c("Complete response" = "plum1", "Partial response" = "brown", "No response" = "darkorange"),
    #            `Initial treatment` = c("3+7" = "aquamarine4", "Other" = "olivedrab2"), 
    #            HSCT = c("HSCT" = "hotpink", "none" = "bisque")
    #        ),
    na_col = "lightgray",
    annotation_name_side = "left"
)

#mysplit <- kmeans(hallmark_es_me_relapse, 3)$cluster

hallmark_es_me_filt <- hallmark_es_me[order(apply(hallmark_es_me, 1, sd), decreasing = TRUE), ]
hallmark_es_me_filt <- hallmark_es_me_filt[1:25, ]
myheatmap <- ComplexHeatmap::Heatmap(
  hallmark_es_me_filt,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  rect_gp = gpar(col = "white", lwd = 2),
  clustering_method_rows = "ward.D2", 
  clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  #row_split = mysplit,
  column_split = cell_state,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(1.5, "mm"),
  row_gap = unit(1.5, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(hallmark_es_me), 
    gp = gpar(fontsize = 12)
  )
)


draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")

```



# Single cell DE analysis across timepoints
```{r de_and_enrichment}

# blast_split <- SplitObject(dplyr::filter(blast_object, !predicted_id %in% "Unknown"), split.by = "predicted_id")  
# blast_split <- lapply(blast_split, function(arg) {
#   arg <- SplitObject(arg, split.by = "patient")
#   return(arg)
# })

scde <- list()
for (cell_state in unique(blast_object$predicted_id)) {
  cell_state <- dplyr::filter(blast_object, predicted_id %in% cell_state)
  for (pat in unique(cell_state$patient_manuscript_id)) {
    pat <- dplyr::filter(cell_state, patient_manuscript_id %in% pat)
    
    if (ncol(pat) < 50) next
    set.seed(52949)
    # cell_sample <- sample(colnames(pat), min(ncol(pat), 2000))
    # pat <- subset(x = pat, cells = cell_sample)
    pat <- CreateSeuratObject(counts = as(LayerData(pat, layer = "counts", assay = "RNA"), "dgCMatrix"), meta.data = pat@meta.data)
    pat <- NormalizeData(pat)
    
    Idents(pat) <- "time"
    
    if (min(table(pat$time)) < 10 | length(unique(pat$time)) < 2) next
    
    top_table <- FindMarkers(pat, 
                             test.use = "wilcox", 
                             #latent.vars = c("percent.mt"),
                             ident.1 = "Relapse",
                             ident.2 = "Diagnosis",
                             random.seed = 56739, 
                             max.cells.per.ident = min(table(pat$time)),
                             logfc.threshold = log2(1.20),
                             min.cells.feature = 10,
                             min.cells.group = 10)
    
    if (nrow(top_table) < 1) {
      next
    } else {
      
      top_table <- top_table %>% 
        dplyr::arrange(desc(avg_log2FC)) %>%
        as.data.frame() %>%
        tibble::rownames_to_column("gene")
      
      pid <- as.character(unique(pat$predicted_id)) 
      patient_id <- as.character(unique(pat$patient))
      scde[[pid]][[patient_id]] <- top_table
      
    }
  } 
}


```



## Run FGSEA for each cell type
```{r enrichment_analysis}

gs_list <- pos_gs_list

es_list <- list()
for (cell_state in names(scde)) {
  
  tmp <- lapply(names(scde[[cell_state]]), function(arg) {
    
    patient <- arg
    arg <- scde[[cell_state]][[arg]]
    mystats <- arg$avg_log2FC
    names(mystats) <- arg$gene
    mystats <- mystats[is.finite(mystats)]
    
    if ((sum(duplicated(mystats)) / length(mystats)) > 0.2) {
      
      return(NULL)
      
    } else {
    
      es_res <- fgsea(pathways = gs_list, stats = mystats, nPermSimple = 100000, nproc = 8, gseaParam = 1) 
  
      if (nrow(es_res) > 0) {
        
        es_res <- es_res %>%
          dplyr::arrange(pval) %>%
          dplyr::mutate(predicted_id = cell_state, patient = patient)
        
        return(es_res)
      }
    }
    
  })
  
  es_list[[cell_state]] <- do.call("rbind", tmp)
  
}

es_df <-  do.call("rbind", es_list)


```

## Heatmap and clustering of ES
```{r heatmap_and_clustering_es, fig.width = 12}

es_df_hsc <- es_df %>% dplyr::filter(!is.na(NES))

mysurv <- surv_data %>%
  dplyr::distinct() %>%
  dplyr::mutate(patient = gsub("^P", "Patient_", patient))

pathways_remove <- c("ADIPOGENESIS", "UV_RESPONSE_UP", "UV_RESPONSE_DN", "PANCREAS_BETA_CELLS", "SPERMATOGENESIS", "ANGIOGENESIS", "COAGULATION", "APICAL_JUNCTION", "APICAL_SURFACE", "MYOGENESIS")

pathways_to_include <- es_df_hsc %>%
  dplyr::filter(padj < 0.1) %>%
  dplyr::pull(pathway) %>%
  unique()

plotdata <- es_df_hsc %>%
  dplyr::filter(pathway %in% pathways_to_include) %>%
  dplyr::filter(padj < 0.1) %>%
  dplyr::select(patient, pathway, NES) %>%
  tidyr::pivot_wider(names_from = pathway, values_from = NES, values_fill = 0) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("patient") %>%
  as.matrix() %>% 
  t()

mutation_data <- aml_data@meta.data %>%
  dplyr::filter(time %in% "Diagnosis") %>%
  dplyr::select(patient_manuscript_id, npm1, flt3_itd, flt3_tkd, kit, runx1, tp53, asxl1, other_mutations_ngs, initial_tx, response_to_initial_tx) %>%
  dplyr::distinct() %>%
  dplyr::mutate(npm1 = ifelse(npm1 %in% "Positive" | grepl("npm1", other_mutations_ngs, ignore.case = TRUE), "Positive", npm1),
                tet2 = case_when(grepl("tet2", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                dnmt3a = case_when(grepl("dnmt3a", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                   is.na(other_mutations_ngs) ~ NA,
                                   TRUE ~ "Undetectable"),
                idh1 = case_when(grepl("idh1", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                idh2 = case_when(grepl("idh2", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                flt3 = case_when(flt3_itd %in% "Positive" | flt3_tkd %in% "Positive" | grepl("FLT3", other_mutations_ngs) ~ "Positive",
                                 flt3_itd %in% "Undetectable" | flt3_tkd %in% "Undetectable" ~ "Undetectable")) %>%
  dplyr::left_join(mysurv, by = "patient_manuscript_id") %>%
  dplyr::filter(patient_manuscript_id %in% colnames(plotdata)) %>%
  dplyr::mutate(response_to_initial_tx = ifelse(response_to_initial_tx == "Complete response (CR)", "Complete response", response_to_initial_tx),
                response_to_initial_tx = factor(response_to_initial_tx, levels = c("Complete response", "Partial response", "No response"))) %>%
  dplyr::arrange(response_to_initial_tx)


npm1 <- mutation_data$npm1
names(npm1) <- mutation_data$patient
dnmt3a <- mutation_data$dnmt3a
names(dnmt3a) <- mutation_data$patient
idh1 = mutation_data$idh1
names(idh1) <- mutation_data$patient
idh2 <- mutation_data$idh2
names(idh2) <- mutation_data$patient
flt3 <- mutation_data$flt3
names(flt3) <- mutation_data$patient

overall_survival <- as.numeric(mutation_data$time_dx_to_last_follow_up)
names(overall_survival) <- mutation_data$patient
time_to_relapse <- as.numeric(mutation_data$time_dx_to_re)
names(time_to_relapse) <- mutation_data$patient
response_to_initial_tx <- mutation_data$response_to_initial_tx
names(response_to_initial_tx) <- mutation_data$patient
initial_tx <- mutation_data$initial_tx
names(initial_tx) <- mutation_data$patient
  
# Re-order ES data based on CITEseq metadata
plotdata <- plotdata[, match(mutation_data$patient_manuscript_id, colnames(plotdata))]

ha <- HeatmapAnnotation(
    # "Overall survival" = overall_survival,
    # "Time to relapse" = time_to_relapse,
    "Initial treatment" = initial_tx,
    "Response to chemotherapy" = response_to_initial_tx,
    # NPM1 = npm1,
    # FLT3 = flt3,
    # DNMT3A = dnmt3a,
    # IDH1 = idh1,
    # IDH2 = idh2,
    col = list(NPM1 = c("Undetectable" = "lightgreen", "Positive" = "magenta"),
               DNMT3A = c("Undetectable" = "lightgreen", "Positive" = "magenta"),
               IDH1 = c("Undetectable" = "lightgreen", "Positive" = "magenta"),
               IDH2 = c("Undetectable" = "lightgreen", "Positive" = "magenta"),
               FLT3 = c("Undetectable" = "lightgreen", "Positive" = "magenta"),
               `Response to chemotherapy` = c("Complete response" = "cornflowerblue", "Partial response" = "brown", "No response" = "yellow"),
               `Initial treatment` = c("3+7" = "aquamarine4", 
                                       "Azacitidine" = "purple", 
                                       "Clinical Trial with SGI" = "tomato1", 
                                       "Clinical Trial with Temozolomide" = "olivedrab2", 
                                       "CPX-351" = "dodgerblue", 
                                       "FLAG-IDA" = "tan2")
               ),
    na_col = "lightgray",
    annotation_name_side = "left"
)


set.seed(194914)
mysplit <- kmeans(plotdata, 3)$cluster
mysplit <- factor(mysplit, levels = c(1, 2, 3))

myheatmap <- ComplexHeatmap::Heatmap(
  plotdata,
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  rect_gp = gpar(col = "white", lwd = 2),
  clustering_method_rows = "ward.D2", 
  #clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  column_km = 3,
  #row_km = 3,
  split = mysplit,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(2, "mm"),
  row_gap = unit(2, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(plotdata), 
    gp = gpar(fontsize = 12)
  )
)

draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")



es_df_hsc <- es_df_hsc %>%
  dplyr::mutate(cluster = case_when(grepl("_41|_50|_52|_54|_61|_67", patient) ~ "cluster_1",
                                    grepl("26|_4|_43|_51|_64|_65|_70|_44", patient) ~ "cluster_3",
                                    TRUE ~ "cluster_2"))

ggboxplot(es_df_hsc, x = "cluster", y = "NES", facet.by = "pathway", fill = "cluster", palette = "npg", ylim = c(-4, 10)) + 
  stat_compare_means()



```


# Single cell DE analysis across timepoints all blasts
```{r de_and_enrichment}

#blast_split <- SplitObject(dplyr::filter(blast_object, !predicted_id %in% "Unknown"), split.by = "patient")  

scde <- list()
for (patient_id in unique(blast_object$patient_manuscript_id)) {
  
  #set.seed(52949)
  # cell_sample <- sample(colnames(pat), min(ncol(pat), 2000))
  # pat <- subset(x = pat, cells = cell_sample)
  pat <- blast_object %>% dplyr::filter(patient_manuscript_id %in% patient_id)
  if (ncol(pat) < 50) next
  pat <- CreateSeuratObject(counts = as(LayerData(pat, layer = "counts", assay = "RNA"), "dgCMatrix"), 
                            meta.data = pat@meta.data)
  pat <- NormalizeData(pat)
  
  Idents(pat) <- "time"
  
  if (min(table(pat$time)) < 10 | length(unique(pat$time)) < 2) next
  
  top_table <- FindMarkers(pat, 
                           test.use = "wilcox", 
                           #latent.vars = c("percent.mt"),
                           ident.1 = "Relapse",
                           ident.2 = "Diagnosis",
                           random.seed = 56739, 
                           max.cells.per.ident = min(table(pat$time)),
                           logfc.threshold = log2(1.20),
                           min.cells.feature = 10,
                           min.cells.group = 10)
  
  if (nrow(top_table) < 1) {
    next
  } else {
    
    top_table <- top_table %>% 
      dplyr::arrange(desc(avg_log2FC)) %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene")
    
    scde[[patient_id]] <- top_table
    
  }
} 


```



## Run FGSEA
```{r enrichment_analysis}

gs_list <- pgs_list

es_list <- lapply(names(scde), function(patient) {
  
  arg <- scde[[patient]]
  mystats <- arg$avg_log2FC
  names(mystats) <- arg$gene
  mystats <- mystats[is.finite(mystats)]
  
  if ((sum(duplicated(mystats)) / length(mystats)) > 0.2) {
    
    return(NULL)
    
  } else {
  
    es_res <- fgsea(pathways = gs_list, stats = mystats, nPermSimple = 100000, nproc = 1, gseaParam = 1) 

    if (nrow(es_res) > 0) {
      
      es_res <- es_res %>%
        dplyr::arrange(pval) %>%
        dplyr::mutate(patient = patient)
      
      return(es_res)
    }
  }
  
})
  

es_df <-  do.call("rbind", es_list)


```

## Heatmap and clustering of ES
```{r heatmap_and_clustering_es, fig.width = 12, fig.height = 4.5}

es_df_hsc <- es_df %>% dplyr::filter(!is.na(NES))
tmp_surv_data <- surv_data %>%
  dplyr::mutate(chr_stability = ifelse(patient %in% c("P61", "P29", "P30", "P50", "P28", "P49", "P52", "P43"), "unstable", "stable"),
                hsct = ifelse(patient %in% c("P26", "P27", "P29", "P30", "P28", "P4", "P31"), "yes", "no"))

pathways_to_include <- c("INFLAMMATORY_RESPONSE",
                         "TNFA_SIGNALING_VIA_NFKB",
                         "INTERFERON_ALPHA_RESPONSE",
                         "IL6_JAK_STAT3_SIGNALING",
                         "INTERFERON_GAMMA_REPONSE",
                         "IL2_STAT5_SIGNALING",
                         "COMPLEMENT",
                         "TGF_BETA_SIGNALING",
                         "MITOTIC_SPINDLE",
                         "E2F_TARGETS",
                         "MYC_TARGETS_V1",
                         "MYC_TARGETS_V2",
                         "G2M_CHECKPOINT",
                         "OXIDATIVE_PHOSPHORYLATION"
                         )
                         
plotdata <- es_df_hsc %>%
  dplyr::filter(pathway %in% pathways_to_include) %>%
  #dplyr::filter(pval < 0.2) %>%
  dplyr::select(patient, pathway, NES) %>%
  tidyr::pivot_wider(names_from = pathway, values_from = NES, values_fill = 0) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("patient") %>%
  as.matrix() %>% 
  t()

pat_order <- plotdata[c("INFLAMMATORY_RESPONSE",
                         "TNFA_SIGNALING_VIA_NFKB",
                         "INTERFERON_ALPHA_RESPONSE",
                         "IL6_JAK_STAT3_SIGNALING",
                         "IL2_STAT5_SIGNALING",
                         "COMPLEMENT",
                         "TGF_BETA_SIGNALING"
                        ), ]
pat_order <- colMeans(pat_order)
pat_order <- sort(pat_order, decreasing = TRUE)
pat_order <- names(pat_order)
plotdata <- plotdata[, pat_order]

# Prepare mutation data
mutation_data <- aml_data@meta.data %>%
  dplyr::filter(time %in% "Diagnosis") %>%
  dplyr::select(patient_manuscript_id, hsct, npm1, flt3_itd, flt3_tkd, kit, runx1, tp53, asxl1, other_mutations_ngs, initial_tx, response_to_initial_tx) %>%
  dplyr::distinct() %>%
  dplyr::mutate(npm1 = ifelse(npm1 %in% "Positive" | grepl("npm1", other_mutations_ngs, ignore.case = TRUE), "Positive", npm1),
                tet2 = case_when(grepl("tet2", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                dnmt3a = case_when(grepl("dnmt3a", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                   is.na(other_mutations_ngs) ~ NA,
                                   TRUE ~ "Undetectable"),
                idh1 = case_when(grepl("idh1", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                idh2 = case_when(grepl("idh2", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                flt3 = case_when(flt3_itd %in% "Positive" | flt3_tkd %in% "Positive" | grepl("FLT3", other_mutations_ngs) ~ "Positive",
                                 flt3_itd %in% "Undetectable" | flt3_tkd %in% "Undetectable" ~ "Undetectable")) %>%
  dplyr::left_join(mysurv, by = "patient_manuscript_id") %>%
  dplyr::filter(patient_manuscript_id %in% colnames(plotdata)) %>%
  dplyr::mutate(response_to_initial_tx = case_when(response_to_initial_tx == "Complete response (CR)" ~ "CR", 
                                                   response_to_initial_tx == "Partial response" ~ "PR",
                                                   response_to_initial_tx == "No response" ~ "PRD"),
                response_to_initial_tx = factor(response_to_initial_tx, levels = c("CR", "PR", "PRD")),
                hsct = ifelse(hsct == "HSCT", "HCT", hsct)) %>%
  dplyr::arrange(response_to_initial_tx) %>%
  dplyr::slice(match(pat_order, patient_manuscript_id))

npm1 <- mutation_data$npm1
names(npm1) <- mutation_data$patient_manuscript_id
dnmt3a <- mutation_data$dnmt3a
names(dnmt3a) <- mutation_data$patient_manuscript_id
idh1 = mutation_data$idh1
names(idh1) <- mutation_data$patient_manuscript_id
idh2 <- mutation_data$idh2
names(idh2) <- mutation_data$patient_manuscript_id
flt3 <- mutation_data$flt3
names(flt3) <- mutation_data$patient_manuscript_id

overall_survival <- as.numeric(mutation_data$time_dx_to_last_follow_up)
names(overall_survival) <- mutation_data$patient_manuscript_id
time_to_relapse <- as.numeric(mutation_data$time_dx_to_re)
names(time_to_relapse) <- mutation_data$patient_manuscript_id
response_to_initial_tx <- mutation_data$response_to_initial_tx
names(response_to_initial_tx) <- mutation_data$patient_manuscript_id
initial_tx <- mutation_data$initial_tx
names(initial_tx) <- mutation_data$patient_manuscript_id
hsct <- mutation_data$hsct
names(hsct) <- mutation_data$patient_manuscript_id
# Re-order ES data based on CITEseq metadata
adist <- mutation_data$adist
mdist <- mutation_data$mutation_distance

# Re-order ES data based on CITEseq metadata
set.seed(34532)
mysplit <- factor(kmeans(plotdata, 2)$cluster, levels = c(2, 1))
set.seed(187134)
colsplit <- factor(kmeans(t(plotdata), 3)$cluster, levels = c(3, 2, 1))

# col_fun <- colorRamp2(quantile(as.numeric(as.matrix(sample_mut_filt[, -1])), 
#                                probs = c(0.05, 0.5, 0.95), na.rm = TRUE), 
#                       c("blue", "white", "red"))

ha <- HeatmapAnnotation(
    #`Aitchison D` = anno_barplot(adist, bar_width = 0.4, height = unit(3, "cm"), gp = gpar(fill = "azure4")),
    #"Overall survival" = anno_barplot(overall_survival, height = unit(2, "cm")),
    #"Time to relapse" = anno_barplot(time_to_relapse, height = unit(4, "cm")),
    "Initial treatment" = ifelse(initial_tx == "3+7", "3+7", "Other"),
    "Response to chemotherapy" = response_to_initial_tx,
    HCT = hsct,
    col = list(`Response to chemotherapy` = c("CR" = "plum1", "PR" = "brown", "PRD" = "darkorange"),
               `Initial treatment` = c("3+7" = "aquamarine4", "Other" = "olivedrab2"), 
               HSCT = c("HSCT" = "hotpink", "none" = "bisque")
           ),
    na_col = "lightgray",
    annotation_name_side = "left"
)

myheatmap <- ComplexHeatmap::Heatmap(
  plotdata,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  #clustering_method_rows = "ward.D2", 
  #clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  na_col = "whitesmoke",
  row_split = mysplit,
  #column_split = colsplit,
  cluster_row_slices = FALSE,
  row_names_side = "left",
  column_gap = unit(2, "mm"),
  row_gap = unit(2, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11, angle = 45),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(plotdata), 
    gp = gpar(fontsize = 12)
  )
)

draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")

```

## Pathway boxplots
```{r pathway_boxplots}

#Generate boxplots
plotdata <- es_df_hsc %>%
  dplyr::select(patient, pathway, NES) %>%
  dplyr::rename(patient_manuscript_id = patient) %>%
  tidyr::pivot_wider(names_from = pathway, values_from = NES, values_fill = NA) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("patient_manuscript_id") %>%
  t()

plotdata <- plotdata[, match(as.character(mutation_data$patient_manuscript_id), colnames(plotdata))]

select_pathway <- c("INFLAMMATORY_RESPONSE",
                     "TNFA_SIGNALING_VIA_NFKB",
                     "INTERFERON_ALPHA_RESPONSE",
                     "IL6_JAK_STAT3_SIGNALING",
                     "IL2_STAT5_SIGNALING",
                     "COMPLEMENT",
                     "TGF_BETA_SIGNALING")

response_plot_list <- lapply(select_pathway, function(arg) {
  
  boxdata <- data.frame(pathway = as.numeric(plotdata[arg, ]),
                        response = mutation_data$response_to_initial_tx) %>%
    dplyr::mutate(response = factor(ifelse(response %in% "CR", "CR", "PR/PRD"), levels = c("CR", "PR/PRD")))
  
  ggboxplot(boxdata,
            x = "response",
            y = "pathway",
            xlab = "",
            ylab = glue::glue("{arg} NES"),
            fill = "wheat2",
            legend = "none",
            add = "jitter") +
    stat_compare_means(comparisons = list(c(1, 2)), digits = 2, method = "t.test") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
})

response_plot_list[[1]] + response_plot_list[[2]] + response_plot_list[[3]] + response_plot_list[[4]] + response_plot_list[[5]] + response_plot_list[[6]] + response_plot_list[[7]] + plot_layout(ncol = 4)

hsct_plot_list <- lapply(select_pathway, function(arg) {
  
  boxdata <- data.frame(pathway = as.numeric(plotdata[arg, ]),
                        hct = factor(hsct, levels = c("HCT", "none"))) 
  
  ggboxplot(boxdata,
            x = "hct",
            y = "pathway",
            xlab = "",
            ylab = glue::glue("{arg} NES"),
            fill = "wheat2",
            legend = "none",
            add = "jitter") +
    stat_compare_means(comparisons = list(c(1, 2)), digits = 2, method = "t.test") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
})

hsct_plot_list[[1]] + hsct_plot_list[[2]] + hsct_plot_list[[3]] + hsct_plot_list[[4]] + hsct_plot_list[[5]] + hsct_plot_list[[6]] + hsct_plot_list[[7]] + plot_layout(ncol = 4)

```

## Feature plots of leading edge genes
```{r leading_edge_genes}

leg_df <- es_df_hsc %>%
  dplyr::filter(pathway %in% c("E2F_TARGETS", "MYC_TARGETS_V1", "TNFA_SIGNALING_VIA_NFKB"))

legs <- NULL
for (mypath in unique(leg_df$pathway)) {
  
  legs[[mypath]] <- leg_df %>% dplyr::filter(pathway %in% mypath) %>% dplyr::pull(leadingEdge) %>% unlist() %>% c() %>% unique()

}

blast_object <- AddModuleScore(blast_object, legs, name = names(legs)) %>%
  dplyr::rename(E2F_TARGETS = E2F_TARGETS1, 
                MYC_TARGETS_V1 = MYC_TARGETS_V12, 
                TNFA_SIGNALING_VIA_NFKB = TNFA_SIGNALING_VIA_NFKB3)

test <- dplyr::filter(blast_object, patient == "Patient_26")
DefaultAssay(test) <- "RNA"
test <- ScaleData(test, features = rownames(test@assays$RNA))
test <- AddModuleScore(test, legs, name = names(legs))

VlnPlot(dplyr::filter(test, predicted_id %in% c("HSC", "MEP", "Monocyte", "DC", "preB")), 
            features = c('E2F_TARGETS1', 'MYC_TARGETS_V12', 'TNFA_SIGNALING_VIA_NFKB3'), 
            group.by = "predicted_id", 
            split.by = "time", 
            cols = c("blue", "red"),
        pt.size = 0.1, alpha = 0.01)

```

# Compare relapse samples between responders and non-responders
```{r respond_vs_non_responde_chemo}

blast_object_dx <- dplyr::filter(blast_object, time == "Diagnosis") 
blast_object_dx <- blast_object_dx %>% dplyr::mutate(response = ifelse(response_to_initial_tx == "Complete response (CR)", "CR", "NR"))
Idents(blast_object_dx) <- "response"

DefaultAssay(blast_object_dx) <- "ADT"
response_de <- FindMarkers(blast_object_dx, ident.1 = "CR", ident.2 = "NR") %>%
  dplyr::arrange(desc(avg_log2FC))



heatmap_sample <- subset(blast_object_dx, cells = sample(colnames(blast_object_dx), 1000))
myfeats <- rownames(response_de)
myfeats <- myfeats[!grepl("isotype", myfeats, ignore.case = TRUE)]
heatmap_sample <- ScaleData(heatmap_sample, features = myfeats)

DoHeatmap(heatmap_sample,
          features = myfeats,
          group.by = "response")


DefaultAssay(heatmap_sample) <- "RNA"
heatmap_sample <- FindVariableFeatures(heatmap_sample, assay = "RNA")
heatmap_sample <- ScaleData(heatmap_sample, assay = "RNA")
heatmap_sample <- RunPCA(heatmap_sample, assay = "RNA")
heatmap_sample <- FindNeighbors(heatmap_sample, assay = "RNA")
heatmap_sample <- FindClusters(heatmap_sample)
heatmap_sample <- RunUMAP(heatmap_sample, dims = 1:50, assay = "RNA")

FeaturePlot(heatmap_sample, features = c("CD33", "CLL1", "CD34"), split.by = "response")


dx_pb <- AggregateExpression(blast_object_dx, group.by = "patient", return.seurat = TRUE, assays = c("ADT", "RNA"))
response <- dplyr::select(blast_object_dx, patient, response) %>% 
  distinct() %>% 
  dplyr::mutate(patient = gsub("_", "-", patient))
dx_pb <- dplyr::left_join(dx_pb, response, by = "patient")
dx_pb <- NormalizeData(dx_pb, assay = "ADT", normalization.method = "CLR", margin = 2)
Idents(dx_pb) <- "response"
response_pb_de <- FindMarkers(dx_pb, 
                              ident.1 = "CR", ident.2 = "NR", 
                              assay = "RNA", 
                              max.cells.per.ident = min(table(dx_pb$response)), 
                              test.use = "wilcox", 
                              logfc.threshold = log2(1.2))
mystats <- response_pb_de$avg_log2FC
names(mystats) <- rownames(response_pb_de)
response_es_res <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 100000, nproc = 5, gseaParam = 1) %>%
  dplyr::arrange(desc(NES))

```

# Cytogenetic analysis
## Positional gene sets
```{r cytogenetic_analysis}

pseudobulk <- AggregateExpression(blast_object, assays = c("ADT", "RNA"), group.by = "sample_id")
pseudobulk_rna <- pseudobulk$RNA

myparam <- GSVA::gsvaParam(exprData = pseudobulk_rna, geneSets = pos_gs_list, kcdf = "Poisson", minSize = 5, sparse = TRUE)
cyto_es <- GSVA::gsva(myparam)

cyto_es_filt <- cyto_es[rowSums(cyto_es >= 0) < ncol(cyto_es), ]

pheatmap(cyto_es_filt)


pseudobulk_adt <- pseudobulk$ADT
pseudobulk_adt <- sweep(pseudobulk_adt + 1, MARGIN = 2, STATS = colSums(pseudobulk_adt), FUN = "/") * 1e5
pseudobulk_adt <- log2(pseudobulk_adt)

cor_mat <- matrix(0, nrow(cyto_es_filt), nrow(pseudobulk_adt))
rownames(cor_mat) <- rownames(cyto_es_filt)
colnames(cor_mat) <- rownames(pseudobulk_adt)
for (gset_ind in 1:nrow(cyto_es_filt)) {
  for(marker_ind in 1:nrow(pseudobulk_adt)) {
    
    #cor_mat[gset_ind, marker_ind] <- cor(cyto_es_filt[gset_ind, ], pseudobulk_adt[marker_ind, ], method = "pearson", use =  "pairwise.complete.obs")
    dep_samples <- ifelse(cyto_es_filt[gset_ind, ] < 0, "depleted", "enriched")
    adt_expr <- pseudobulk_adt[marker_ind, ]
    stopifnot(all(names(dep_samples) == names(adt_expr)))
    mytest <- t.test(adt_expr ~ dep_samples)
    
    if (mytest$p.value > 0.1) {
      cor_mat[gset_ind, marker_ind] <- 0
    } else {
      cor_mat[gset_ind, marker_ind] <- mytest$statistic
    }
  }
}

cor_mat[is.na(cor_mat) | cor_mat > 0] <- 0
cor_mat <- cor_mat[rowSums(cor_mat == 0) < ncol(cor_mat), ]

cor_mat <- as.data.frame(cor_mat) %>%
  tibble::rownames_to_column("position") %>%
  dplyr::mutate(chr = as.numeric(gsub("p.*|q.*|chr", "", position)),
                arm = ifelse(grepl("p", position), "p", "q"),
                arm_number = gsub(".*p|.*q", "", position)) %>%
  dplyr::arrange(chr, arm, arm_number) %>%
  dplyr::select(-arm, -arm_number, -chr) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("position")



set.seed(34523532)
myheatmap <- ComplexHeatmap::Heatmap(
  t(cor_mat),
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 2),
  clustering_method_rows = "ward.D2", 
  clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  na_col = "whitesmoke",
  #column_km = 3,
  row_km = 3,
  row_names_side = "left",
  column_gap = unit(2, "mm"),
  row_gap = unit(2, "mm"),
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 11),
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  row_names_max_width = max_text_width(
    rownames(cor_mat), 
    gp = gpar(fontsize = 12)
  )
)

draw(myheatmap, padding = unit(c(2, 20, 2, 2), "mm"), row_dend_side = "right")



```

## InferCNV
```{r infercnv}

aml_patients <- c("Patient_26",
                  "Patient_27",
                  "Patient_28",
                  "Patient_29",
                  "Patient_30",
                  "Patient_31",
                  "Patient_4",
                  "Patient_40",
                  "Patient_41",
                  "Patient_42",
                  "Patient_43",
                  "Patient_44",
                  "Patient_49",
                  "Patient_50",
                  "Patient_51",
                  "Patient_52",
                  "Patient_54",
                  "Patient_56",
                  "Patient_58",
                  "Patient_60",
                  "Patient_61",
                  "Patient_64",
                  "Patient_65",
                  "Patient_66",
                  "Patient_67",
                  "Patient_70")

healthy_samples <- c("Donor_1_Healthy", "Donor_2_Healthy", "Donor_3_Healthy", 
                     "Donor_4_Healthy", "Donor_5_Healthy", "Donor_6_Healthy",
                     "JD_Healthy_Donor_1_Healthy", "JD_Healthy_Donor_2_Healthy", 
                     "JD_Healthy_Donor_3_Healthy", "JD_Healthy_Donor_4_Healthy")

atlas <- readRDS("/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/seurat_2024-04-03/aml_atlas_layered_seurat.rds")

healthy_data_tmp <- readRDS("/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/seurat_2024-04-03/reference.rds")
healthy_cells <- healthy_data_tmp$barcode[healthy_data_tmp$cell_state_mnn_rna_l1 %in% c("Monocyte", "cDC", "pDC", "HSC", "MEP", "GMP", "preB")]
healthy_metadata <- healthy_data_tmp@meta.data
rm(healthy_data_tmp)

atlas <- dplyr::filter(atlas, barcode %in% healthy_cells | barcode %in% blast_object$barcode)

healthy_cells <- atlas$barcode[atlas$barcode %in% healthy_cells]
set.seed(7798)
healthy_cells <- sample(healthy_cells, 3000, replace = FALSE)

for (mypat in aml_patients) {
  
  cat("\n", mypat)
  atlas_sample <- dplyr::filter(atlas, patient %in% mypat | barcode %in% healthy_cells)
  aml_cells_dx <- atlas_sample$barcode[!atlas_sample$barcode %in% healthy_cells & atlas_sample$time %in% "Diagnosis"]
  set.seed(13410)
  aml_cells_sample_dx <- sample(aml_cells_dx, min(length(aml_cells_dx), 3000), replace = FALSE)
  aml_cells_re <- atlas_sample$barcode[!atlas_sample$barcode %in% healthy_cells & atlas_sample$time %in% "Relapse"]
  set.seed(77893)
  aml_cells_sample_re <- sample(aml_cells_re, min(length(aml_cells_re), 3000), replace = FALSE)
  aml_cells_sample <- c(aml_cells_sample_dx, aml_cells_sample_re)
  
  atlas_sample <- subset(atlas_sample, cells = c(healthy_cells, aml_cells_sample))
  atlas_sample_matrix <- as(LayerData(JoinLayers(atlas_sample), assay = "RNA", layer = "counts"), "dgCMatrix")
  
  atlas_sample_metadata <- dplyr::select(atlas_sample@meta.data, barcode, time) %>%
    dplyr::left_join(dplyr::select(healthy_metadata, barcode, cell_state_mnn_rna_l1), by = "barcode") %>%
    dplyr::mutate(time = ifelse(!is.na(cell_state_mnn_rna_l1), as.character(cell_state_mnn_rna_l1), as.character(time))) %>%
    dplyr::select(-cell_state_mnn_rna_l1)
  
  stopifnot(ncol(atlas_sample_matrix) == nrow(atlas_sample_metadata))
  
  metadata_filepath <- glue::glue("/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/infercnv_output_18Nov2024/{mypat}/infercnv_annotation_file_{mypat}.txt")
  if (!dir.exists(dirname(metadata_filepath))) dir.create(dirname(metadata_filepath))
  write.table(atlas_sample_metadata, 
              file =  metadata_filepath, 
              sep = "\t", col.names = FALSE, row.names = FALSE)
  
  infercnv_obj <- infercnv::CreateInfercnvObject(raw_counts_matrix = atlas_sample_matrix,
                                                 annotations_file =  metadata_filepath,
                                                 delim = "\t",
                                                 gene_order_file = "https://data.broadinstitute.org/Trinity/CTAT/cnv/hg38_gencode_v27.txt",
                                                 ref_group_names = atlas_sample_metadata$time[grepl("Healthy", atlas_sample_metadata$barcode)],
                                                 chr_exclude = c("chrY", "chrM")) 
            
  infercnv_obj <- infercnv::run(infercnv_obj,
                                cutoff = 0.1,
                                out_dir = glue::glue("/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/infercnv_output_18Nov2024/{mypat}"), 
                                cluster_by_groups = TRUE, 
                                denoise = TRUE,
                                HMM = TRUE,
                                hclust_method = "complete",
                                k_obs_groups = 2,
                                mask_nonDE_pval = 0.0,
                                tumor_subcluster_partition_method = "leiden",
                                num_threads = 10,
                                output_format = "pdf")
  
}

## upload InferCNV images
myfiles <- list.files(path = "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/infercnv_output_15Nov2024",
                      recursive = TRUE,
                      full.names = TRUE,
                      pattern = "infercnv.pdf")

for (f in myfiles) {
  
 patient = basename(gsub("\\/infercnv.pdf", "", f))
 fname <- paste(patient, "infercnv.pdf", sep = "_")
 system(command = glue::glue("aws s3 cp {f} s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/InferCNV_output/v2/{fname}"))
  
}

```

### Correlate CNV with cell composition
```{r cnv_cell_state_correlation}

infercnv_dir <- "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas/infercnv_output_4Nov2024"

chr <- "^chr11"
patient_dir <- list.files(infercnv_dir, full.names = TRUE)
pat_cnv_list <- NULL
for (pat in patient_dir) {
  
  cell_group <- read.delim(file.path(pat, "17_HMM_predHMMi6.leiden.hmm_mode-subclusters.cell_groupings")) %>%
    dplyr::group_by(cell_group_name) %>%
    dplyr::summarize(cell = n()) %>%
    dplyr::ungroup()
  
  # keep cell subclusters with detected copy number loss in chromosome of interest and sum the number of cells across subclsuters
  # we do counting at subcluster instead of cell level because single cell data is sparse. cell belonging to subcluster
  # with a lot cells with CNV is assumed to have it as well due to its similarity in transcriptome profile
  cnv_regions <- read.delim(file.path(pat, "17_HMM_predHMMi6.leiden.hmm_mode-subclusters.pred_cnv_regions.dat")) %>%
    dplyr::filter(grepl(!!chr, cnv_name), state %in% c(1, 2), grepl("Diagnosis|Relapse", cell_group_name)) %>%
    dplyr::mutate(time = ifelse(grepl("Diagnosis", cell_group_name), "Dx", "Re")) %>%
    dplyr::select(cell_group_name, time) %>% # ignore the number of regions being deleted. as long as there is just one region, it's enough
    dplyr::left_join(cell_group, by = "cell_group_name") %>%
    dplyr::group_by(time) %>%
    dplyr::summarize(pct_cells_with_cn_loss = sum(cell) / 1000) %>%
    dplyr::group_by()

  pat_cnv_list[[basename(pat)]] <-data.frame(patient = gsub("Patient_", "P", basename(pat)), cnv_regions)
  
}

pat_cnv <- do.call("rbind", pat_cnv_list) %>%
  dplyr::right_join(cluster_summary, by = c("patient", "time"))

pat_cnv_filt <- pat_cnv %>%
  dplyr::filter(`Blast cell state` %in% c("HSC")) 

ggscatter(data = pat_cnv_filt,
          x = "pct",
          y = "pct_cells_with_cn_loss",
          color = "time",
          cor.coef = TRUE,
          cor.method = "spearman",
          label = "patient") +
  geom_smooth(method = "lm", lty = 2, lwd = 0.5, se = FALSE)




pat_cnv_filt <- pat_cnv_filt %>%
  dplyr::mutate(pct_binary = factor(ifelse(pct > 90, "HSC-dominant", "Other"), levels = c("HSC-dominant", "Other")),
                sample_id = paste0(patient, substring(time, 1, 1)),
                label = ifelse(pct_binary %in% "HSC-dominant", sample_id, NA))

ggboxplot(data = pat_cnv_filt,
          x = "pct_binary",
          y = "pct_cells_with_cn_loss",
          xlab = "",
          ylab = "Fraction of cells with deletion in chr11",
          add = "jitter",
          fill = "cornflowerblue") +
  stat_compare_means(label.x = 0.85, label.y = 0.95)

```

# Immunophenotypic subtypes 
```{r ip_subs}

hsc_blasts <- dplyr::filter(blast_object, predicted_id %in% "HSC")

DefaultAssay(hsc_blasts) <- "RNA"

hsc_blasts[["RNA"]] <- split(hsc_blasts[["RNA"]], f = hsc_blasts$patient)
hsc_blasts <- NormalizeData(hsc_blasts)
hsc_blasts <- FindVariableFeatures(hsc_blasts)
hsc_blasts <- ScaleData(hsc_blasts)
hsc_blasts <- RunPCA(hsc_blasts)
hsc_blasts <- IntegrateLayers(object = hsc_blasts, 
                              method = HarmonyIntegration, 
                              orig.reduction = "pca",
                              new.reduction = 'harmony', 
                              verbose = FALSE)

hsc_blasts <- RunUMAP(hsc_blasts, reduction = "harmony", dims = 1:50)

DimPlot(hsc_blasts, group.by = "seurat_clusters", reduction = "umap")

Idents(hsc_blasts) <- "seurat_clusters"
hsc_markers <- FindAllMarkers(hsc_blasts, only.pos = TRUE, assay = "ADT")

```

# Cellchat analysis
```{r cellchat_analysis}

aml_sample <- subset(aml_data, cells = sample(colnames(aml_data)[aml_data$time == "Relapse"], 10000))
DefaultAssay(aml_sample) <- "RNA"
aml_sample <- NormalizeData(aml_sample)

data.input <- LayerData(aml_sample, layer = "data", assay = "RNA")
labels <- aml_sample$predicted_id
meta <- data.frame(labels = labels, samples = aml_sample$sample_id, row.names = names(labels)) # create a dataframe of the cell labels
meta$labels <- droplevels(meta$labels)

cellchat <- CellChat::createCellChat(object = as(data.input, "dgCMatrix"), meta = meta, group.by = "labels")

CellChatDB <- CellChatDB.human
# use a subset of CellChatDB for cell-cell communication analysis
CellChatDB.use <- subsetDB(CellChatDB, search = "Secreted Signaling", key = "annotation") # use Secreted Signaling
cellchat@DB <- CellChatDB.use
cellchat <- CellChat::subsetData(cellchat) # This step is necessary even if using the whole database
future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)

cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10)
cellchat <- computeCommunProbPathway(cellchat)
cellchat <- aggregateNet(cellchat)

groupSize <- as.numeric(table(cellchat@idents))
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Number of interactions")


mat <- cellchat@net$weight
par(mfrow = c(3, 4), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = T, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}

netVisual_heatmap(cellchat, signaling = pathways.show, color.heatmap = "Reds")

```





# LSCs
## Find signature for LSCs
```{r lsc_surface_markers}

Idents(blast_object) <- "lsc"
DefaultAssay(blast_object) <- "ADT"

top_table <- FindMarkers(blast_object,
                         latent.vars = "patient",
                         ident.1 = "LSC",
                         logfc.threshold = log2(1.2))


DefaultAssay(blast_object) <- "RNA"
tmp <- subset(blast_object, cells = sample(colnames(blast_object), 5000))
tmp <- CreateSeuratObject(counts = as(LayerData(tmp, assay = "RNA", layer = "counts"), "dgCMatrix"), meta.data = tmp@meta.data)
Idents(tmp) <- "lsc"
top_table_rna <- FindMarkers(tmp,
                             test.use = "MAST",
                             latent.vars = c("patient", "nCount_RNA", "mt.percent"),
                             ident.1 = "LSC",
                             logfc.threshold = log2(1.2))

mystats <- top_table_rna$avg_log2FC
names(mystats) <- rownames(top_table_rna)
mystats <- mystats[is.finite(mystats)]
es_lsc <- fgsea(pathways = pgs_list, stats = mystats)

```

## Differential expression analysis between relapse and relapse groups
### Single cell
```{r de_and_enrichment}

set.seed(13107)
sample_n <- 5000
set1 <- sample(colnames(lsc_object)[lsc_object$time == "Diagnosis"], sample_n)
set2 <- sample(colnames(lsc_object)[lsc_object$time == "Relapse"], sample_n)

tmp <- subset(lsc_object, cells = c(set1, set2))
tmp <- CreateSeuratObject(counts = as(LayerData(tmp, assay = "RNA", layer = "counts"), "dgCMatrix"), meta.data = tmp@meta.data)
keep_features <- rowSums(tmp@assays$RNA@counts)
tmp <- subset(tmp, features = names(keep_features)[keep_features > 50])
tmp <- NormalizeData(tmp, assay = "RNA", normalization.method = "LogNormalize")

Idents(tmp) <- "time"
DefaultAssay(tmp) <- "RNA"
top_table <- FindMarkers(tmp, 
                         test.use = "MAST", 
                         latent.vars = c("patient", "percent.mt"),
                         ident.1 = "Relapse",
                         ident.2 = "Diagnosis",
                         random.seed = 565099, 
                         min.cells.feature = 10,
                         min.cells.group = 10
                         )

# top_table <- FindMarkers(tmp, 
#                          test.use = "wilcox",
#                          ident.1 = "Relapse",
#                          ident.2 = "Diagnosis",
#                          random.seed = 565099, 
#                          min.cells.feature = 10,
#                          min.cells.group = 10
#                          )

top_table <- top_table %>%
  dplyr::arrange(p_val) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene")

EnhancedVolcano(top_table,
                lab = top_table$gene,
                x = 'avg_log2FC',
                y = 'p_val')

#fgsea
mystats <- top_table$avg_log2FC
names(mystats) <- top_table$gene
mystats <- mystats[is.finite(mystats)]
es_res <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 10000) %>%
  dplyr::arrange(pval)

sig_paths <- es_res$pathway[es_res$padj < 0.05]

plotEnrichment(pgs_list[["HALLMARK_TNFA_SIGNALING_VIA_NFKB"]], mystats) + ggtitle("HALLMARK_TNFA_SIGNALING_VIA_NFKB") + theme(plot.title = element_text(size = 9))
plotEnrichment(pgs_list[["HALLMARK_HYPOXIA"]], mystats) + ggtitle("HALLMARK_HYPOXIA") + theme(plot.title = element_text(size = 9))
plotEnrichment(pgs_list[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]], mystats) + ggtitle("HALLMARK_INTERFERON_GAMMA_RESPONSE") + theme(plot.title = element_text(size = 9))
plotEnrichment(pgs_list[["HALLMARK_MYC_TARGETS_V1"]], mystats) + ggtitle("HALLMARK_MYC_TARGETS_V1") + theme(plot.title = element_text(size = 9))

```

```{r de_analysis_per_patient}

lsc_object_split <- SplitObject(lsc_object, split.by = "patient")

de_list <- NULL
es_list <- NULL

for(pat in lsc_object_split) {
  
  Idents(pat) <- "time"
  
  if (length(unique(pat$time)) < 2 | min(table(pat$time)) < 20) next  
  
  top_table <- FindMarkers(pat, 
                           test.use = "wilcox", 
                           ident.1 = "Relapse",
                           ident.2 = "Diagnosis",
                           random.seed = 56739, 
                           #max.cells.per.ident = 1.2 * min(table(pat$time)),
                           min.cells.feature = 10,
                           min.cells.group = 10) %>%
    dplyr::arrange(p_val)
  
  de_list[[unique(pat$patient)]] <- top_table
    
}

for (pat in names(de_list)) {
  
  mystats <- de_list[[pat]]$avg_log2FC
  names(mystats) <- rownames(de_list[[pat]])
  mystats <- mystats[is.finite(mystats)]
  
  es_list[[pat]] <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 10000) %>%
    dplyr::arrange(pval) %>%
    dplyr::mutate(patient = gsub("Patient_", "P", pat))

}

es_df <-  do.call("rbind", es_list)

plotdata <- es_df %>%
  dplyr::filter(padj < 0.1) %>%
  dplyr::mutate(pval = -log10(pval)) %>%
  dplyr::mutate(pval = ifelse(NES > 0, pval, -pval)) %>%
  dplyr::select(patient, pathway, NES) %>%
  dplyr::mutate(pathway = gsub("HALLMARK_", "", pathway)) %>%
  tidyr::pivot_wider(names_from = pathway, values_from = NES, values_fill = 0) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("patient") %>%
  as.matrix() %>% 
  t()

mutation_data <- aml_data@meta.data %>%
  dplyr::filter(time %in% "Diagnosis") %>%
  dplyr::select(patient, npm1, flt3_itd, flt3_tkd, kit, runx1, tp53, asxl1, other_mutations_ngs) %>%
  dplyr::distinct() %>%
  dplyr::mutate(npm1 = ifelse(npm1 %in% "Positive" | grepl("npm1", other_mutations_ngs, ignore.case = TRUE), "Positive", npm1),
                tet2 = case_when(grepl("tet2", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                dnmt3a = case_when(grepl("dnmt3a", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                   is.na(other_mutations_ngs) ~ NA,
                                   TRUE ~ "Undetectable"),
                idh1 = case_when(grepl("idh1", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable"),
                idh2 = case_when(grepl("idh2", other_mutations_ngs, ignore.case = TRUE) ~ "Positive",
                                 is.na(other_mutations_ngs) ~ NA,
                                 TRUE ~ "Undetectable")) %>%
  dplyr::mutate(patient = gsub("Patient_", "P", patient)) %>%
  dplyr::left_join(surv_data, by = "patient") %>%
  dplyr::arrange(desc(time_dx_to_last_follow_up)) %>%
  dplyr::filter(patient %in% colnames(plotdata))

plotdata <- plotdata[, match(mutation_data$patient, colnames(plotdata))]

npm1 <- mutation_data$npm1
names(npm1) <- mutation_data$patient
dnmt3a <- mutation_data$dnmt3a
names(dnmt3a) <- mutation_data$patient
idh1 = mutation_data$idh1
names(idh1) <- mutation_data$patient
idh2 <- mutation_data$idh2
names(idh2) <- mutation_data$patient

overall_survival <- as.numeric(mutation_data$time_dx_to_last_follow_up)
names(overall_survival) <- mutation_data$patient

ha <- HeatmapAnnotation(
  "Overall survival" = overall_survival,
  NPM1 = npm1, 
  DNMT3A = dnmt3a,
  IDH1 = idh1,
  IDH2 = idh2,
  col = list(NPM1 = c("Undetectable" = "black", "Positive" = "magenta"),
             DNMT3A = c("Undetectable" = "black", "Positive" = "magenta"),
             IDH1 = c("Undetectable" = "black", "Positive" = "magenta"),
             IDH2 = c("Undetectable" = "black", "Positive" = "magenta")
             ),
  na_col = "gray"
)

ComplexHeatmap::Heatmap(
  plotdata,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  #clustering_method_rows = "ward.D2", 
  #clustering_method_columns = "ward.D2",
  show_column_names = TRUE,
  top_annotation = ha,
  #column_split = as.factor(gsub(".*-|_relapse", "", colnames(myheat_top))),
  #column_gap = unit(2, "mm"),
  #row_gap = unit(2, "mm"),
  #col = col_fun,
  column_title_rot = 45,
  column_title_gp = grid::gpar(fontsize = 13),
  row_names_gp = grid::gpar(fontsize = 13),
  #row_split = row_split,
  row_title_gp = grid::gpar(fontsize = 0),
  border = TRUE,
  #use_raster = FALSE
)


```

### Pseudobulk
```{r pseudobulk_de}

lsc_pseudobulk <- AggregateExpression(lsc_object, group.by = c("sample_id"), return.seurat = TRUE, assays = "RNA") %>%
  dplyr::mutate(sample_id = gsub("Patient-", "P", sample_id)) %>%
  tidyr::separate(sample_id, into = c("patient", "time"), sep = "-")

feature_sums <- rowSums(LayerData(lsc_pseudobulk, assays = "RNA", layer = "counts"))
lsc_pseudobulk <- subset(lsc_pseudobulk, features = rownames(lsc_pseudobulk@assays$RNA)[feature_sums > 1000])
lsc_pseudobulk <- dplyr::mutate(lsc_pseudobulk, time = factor(time, levels = c("Diagnosis", "Relapse")), patient = factor(patient))

# limma-voom
design_matrix <- model.matrix(~time + patient, data = lsc_pseudobulk@meta.data)

v <- limma::voom(counts = as.matrix(LayerData(lsc_pseudobulk, layer = "counts", assay = "RNA")),
                 design = design_matrix,
                 plot = TRUE,
                 normalize = "quantile")

fit <- limma::lmFit(v, design = design_matrix)
fit <- limma::eBayes(fit)
top_table_limma <- limma::topTable(fit, number = Inf, coef = "timeRelapse", sort.by = "P")

mystats <- top_table_limma$logFC
names(mystats) <- rownames(top_table_limma)
mystats <- mystats[is.finite(mystats)]
es_res <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 10000) %>%
  dplyr::arrange(pval)

es_res

#edgeR
lsc_pseudobulk <- edgeR::Seurat2PB(lsc_object, cluster = "time", sample = "sample_id")
lsc_pseudobulk$samples$patient <- gsub("_relapse|_Relapse", "", lsc_pseudobulk$samples$sample)
keep.genes <- filterByExpr(lsc_pseudobulk, min.count = 10, min.total.count = 100)
lsc_pseudobulk <- lsc_pseudobulk[keep.genes, , keep = FALSE]
lsc_pseudobulk <- normLibSizes(lsc_pseudobulk)

cluster <- as.factor(lsc_pseudobulk$samples$cluster)
patient <- as.factor(lsc_pseudobulk$samples$patient)

design <- model.matrix(~ 0 + cluster + patient)
con <- makeContrasts(clusterRelapse - clusterrelapse, levels = design)
lsc_pseudobulk <- estimateDisp(lsc_pseudobulk, design, robust = TRUE)
fit <- glmQLFit(lsc_pseudobulk, design, robust = TRUE)
qlf <- glmQLFTest(fit, contrast = con)
top_table_edger <- qlf$table
top_table_edger <- top_table_edger %>%
  dplyr::mutate(padj = p.adjust(PValue, method = "BH")) %>%
  dplyr::arrange(PValue)

EnhancedVolcano(top_table_edger,
                lab = top_table_edger %>% rownames,
                x = 'logFC',
                y = 'PValue',
                FCcutoff = 0.5,
                pCutoff = 0.05)


mystats <- top_table_edger$logFC
names(mystats) <- rownames(top_table_edger)
mystats <- mystats[is.finite(mystats)]
es_res <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 10000) %>%
  dplyr::arrange(pval)

es_res
 
```












## Enrichment analysis
### Run FGSEA
```{r enrichment_analysis}


es_list <- list()
for (cell_state in names(scde)) {
  
  tmp <- lapply(names(scde[[cell_state]]), function(arg) {
    
    patient <- arg
    arg <- scde[[cell_state]][[arg]]
    mystats <- arg$avg_log2FC
    names(mystats) <- arg$gene
    
    es_res <- fgsea(pathways = pgs_list, stats = mystats, nPermSimple = 10000) 

    if (nrow(es_res) > 0) {
      es_res <- es_res %>%
        dplyr::arrange(pval) %>%
        dplyr::mutate(predicted_id = cell_state, patient = patient)
      return(es_res)
    }
    
  })
  
  es_list[[cell_state]] <- do.call("rbind", tmp)
  
}

es_df <-  do.call("rbind", es_list)


```

## GSVA on LSC pseudobulk
```{r gsva_lscs}

lsc_pseudobulk <- AggregateExpression(lsc_object, group.by = c("sample_id"))
lsc_pseudobulk_rna <- lsc_pseudobulk$RNA

gsvaPar <- gsvaParam(lsc_pseudobulk_rna, pgs_list)
sse <- GSVA::gsva(gsvaPar)

res <- c()
for (gs in rownames(sse)) {
  
  x <- sse[, grepl("Relapse", colnames(sse))]
  colnames(x) <- gsub("-Relapse", "", colnames(x))
  y <- sse[, grepl("Diagnosis", colnames(sse))]
  colnames(y) <- gsub("-Diagnosis", "", colnames(y))
  int <- intersect(colnames(x), colnames(y))
  res[gs] <- wilcox.test(x[gs, colnames(x) %in% int], y[gs, colnames(y) %in% int], paired = TRUE)$p.value
  
}

res <- sort(res)


```



