---
title: "Inference of antigen density by integrating CITEseq and flow cytometry data"
author: "Matt Ung"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: yeti
    toc: true
    number_sections: true
---

```{r setup, include = FALSE}

library(devtools)
library(flowCore)
library(SeuratObject)
library(BPCells)
library(Seurat)
library(SeuratDisk)
library(dplyr)
library(aws.s3)
library(tibble)
library(tidyr)
library(dotenv)
library(ggplot2)
library(ggpubr)
library(glue)
library(DoubletFinder)
library(tidyseurat)
library(pzfx)
library(preprocessCore)
library(pals)
library(EnhancedVolcano)
library(SeuratWrappers)
library(alphahull)
library(igraph)
library(reticulate)
library(viridis)
library(msigdbr)
library(GSVA)
library(escape)
library(dittoSeq)
library(fgsea)
library(GSEABase)
library(ggbeeswarm)
library(circlize)
library(pheatmap)
library(biomaRt)
library(SingleCellExperiment)
library(Matrix)
library(stringr)
library(ComplexHeatmap)
library(future)
library(circlize)
library(msigdbr)
library(survival)
library(survminer)
library(broom)
library(scales)
library(lubridate)
library(future)
library(ggsci)
library(bazar)
library(flowCore)
library(mclust)
library(doParallel)
library(foreach)
library(scales)
library(tictoc)
library(randomForest)
library(e1071)
library(survminer)
library(robCompositions)

options(Seurat.object.assay.version = "v5")
options(future.globals.maxSize = 1e9)

load_dot_env(file = "/home/rstudio/data/bitbucket/.env")

knitr::opts_knit$set(root.dir = "/home/rstudio/data/project_workspace/internal/sc_rnaseq/meta_analysis/aml_atlas")

Sys.setenv(
  "AWS_ACCESS_KEY_ID" = Sys.getenv("key"),
  "AWS_SECRET_ACCESS_KEY" = Sys.getenv("secret"),
  "AWS_DEFAULT_REGION" = "us-east-1"
)

cl <- makeCluster(6)
registerDoParallel(cl)

```

# Calculate coverage function
```{r calculate_coverage}

calculate_coverage <- function(imputed_abc_df, fmo_threshold_df = NULL, target_threshold = 100, label_threshold = 50, first_antigen, second_antigen) {

  imputed_abc_coverage <- imputed_abc_df %>%
    tibble::rownames_to_column("cell_barcode") %>%
    dplyr::mutate(time = gsub(".*_", "", cell_barcode),
                  patient = gsub("_Diagnosis|_Relapse", "", cell_barcode),
                  patient = gsub(".*-", "", patient)) %>%
    dplyr::select(patient, time, one_of(first_antigen), one_of(second_antigen))
  
  colnames(imputed_abc_coverage) <- c("patient", "time", "first_antigen", "second_antigen")
  
  if (!is.null(fmo_threshold_df)) {
    
    imputed_abc_coverage <- imputed_abc_coverage %>%
      dplyr::left_join(fmo_threshold_df, by = "patient") %>%
      dplyr::group_by(patient, time) %>%
      dplyr::summarize(percent_first_ag = sum(first_antigen > fmo_ag_density_boundary) / n() * 100,
                       percent_second_ag = sum(second_antigen > fmo_ag_density_boundary) / n() * 100) %>%
      dplyr::ungroup()
    
  } else {
    
    imputed_abc_coverage <- imputed_abc_coverage %>%
      dplyr::group_by(patient, time) %>%
      dplyr::summarize(percent_first_ag = sum(first_antigen > target_threshold) / n() * 100,
                       percent_second_ag = sum(second_antigen > target_threshold) / n() * 100) %>%
      dplyr::ungroup()
    
  }
  
  imputed_abc_coverage <- imputed_abc_coverage %>%
    tidyr::unite(col = label, patient, time, sep = "_", remove = FALSE) %>%
    dplyr::mutate(label = ifelse(percent_first_ag < label_threshold | percent_second_ag < label_threshold, label, ""))
    
  return(imputed_abc_coverage)
  
}

```


# Load data objects
```{r load_objects}

aml_data <- aws.s3::s3readRDS(object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/seurat_v5/seurat_2024-04-03/aml_data_v4.rds")

patient_order <- c("P26", "P44", "P52", "P70", "P61", "P31", "P43", "P30", "P64", "P56", "P54", "P27", "P41", "P49", "P51", "P50", "P66", "P29", "P67", "P42", "P65", "P60", "P58", "P28", "P40", "P4") 
patient_id_map <- data.frame(patient = patient_order, 
                             patient_manuscript_id = paste0("P", 1:length(patient_order))) %>%
  dplyr::mutate(patient = factor(patient, levels = patient_order),
                patient_manuscript_id = ifelse(nchar(patient_manuscript_id) == 2, gsub("P", "P0", patient_manuscript_id), patient_manuscript_id),
                patient_manuscript_id = factor(patient_manuscript_id, levels = patient_manuscript_id))

aml_data <- aml_data %>%
  dplyr::left_join(dplyr::mutate(patient_id_map, patient = gsub("^P", "Patient_", patient)), by = "patient") %>%
  dplyr::mutate(sample_manuscript_id = ifelse(sample_status == "Diagnosis", paste0(patient_manuscript_id, "D"), paste0(patient_manuscript_id, "R")))

blast_object <- dplyr::filter(aml_data, cell_state_harmony_rna %in% "Blast")

```

# Load .fcs files
```{r flow_data_exploratory_analysis}

flow_paths <- aws.s3::get_bucket(bucket = "vorbio-datascience", 
                                 prefix = "genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data",
                                 max = Inf)
flow_paths <- lapply(flow_paths, `[`, 1) %>% unlist()
flow_paths <- flow_paths[grepl(".fcs", flow_paths)]
flow_paths <- paste0("s3://vorbio-datascience/", flow_paths[grepl("patients_", flow_paths)])
flow_paths <- flow_paths[!grepl("patients_47_48", flow_paths)]

mynames <- gsub("s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/|.fcs", "", flow_paths)

flow_list <- lapply(flow_paths, function(object_path) aws.s3::s3read_using(FUN = flowCore::read.FCS, 
                                                                           transformation = FALSE, 
                                                                           truncate_max_range = TRUE, 
                                                                           object = object_path))
names(flow_list) <- mynames

flow_list <- lapply(flow_list, function(arg) flowCore::transform(arg, `PE-A` = asinh(`PE-A`)))

flow_list_qb <- flow_list[grepl("quantibrite", names(flow_list), ignore.case = TRUE)]
flow_list_lsc <- flow_list[grepl("LSC", names(flow_list), ignore.case = TRUE)]
flow_list_fmo <- flow_list[grepl("FMO", names(flow_list), ignore.case = TRUE)]
flow_list_ags <- flow_list[grepl("blasts_PE+", names(flow_list), ignore.case = TRUE)]
flow_list_ags_live <- flow_list[grepl("Live", names(flow_list), ignore.case = TRUE)]

sample_df <- lapply(names(flow_list_ags), function(arg) {
  
  fi_matrix <- exprs(flow_list_ags[[arg]])
  
  if (nrow(fi_matrix) == 0) {
    return(NULL)
  } else {
    data.frame(cell = paste0(arg, "-cell_", 1:nrow(fi_matrix)),
               sample = arg,
               fi_matrix)
  }
  
})


```

# Check blast percentage concordance
```{r check_blast_percentage}

blast_counts <- table(aml_data$cell_state_harmony_rna, aml_data$sample_id)
#blast_counts <- table(aml_data$predicted_id_blast, aml_data$sample_id)

col_sums <- colSums(blast_counts)
blast_counts <- sweep(x = blast_counts, MARGIN = 2, STATS = col_sums, FUN = "/")
blast_percent <- blast_counts["Blast", , drop = FALSE] %>% 
  t() %>%
  as.data.frame.matrix() %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::mutate(Blast = Blast * 100)
                      
blast_metadata <- aws.s3::s3read_using(FUN = readr::read_csv,
                                       object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/aml_atlas_sample_metadata.csv") %>%
  dplyr::select(sample_id, blast_percent_vor) %>%
  dplyr::filter(!grepl("Healthy", sample_id)) %>%
  dplyr::filter(!is.na(blast_percent_vor))

blast_data <- dplyr::inner_join(blast_percent, blast_metadata, by = "sample_id")
model_res <- lm(blast_percent_vor ~ Blast, data = blast_data)$residuals
blast_data <- dplyr::mutate(blast_data, residuals = model_res, abs_dev = abs(Blast - blast_percent_vor)) %>%
  dplyr::mutate(`Model` = ifelse(abs(residuals) < 25, "yes", "no"))

deviant_samples <- blast_data$sample_id[blast_data$Model == "no"]

ggpubr::ggscatter(blast_data,
                  x = "Blast",
                  y = "blast_percent_vor",
                  xlab = "CITEseq blast %",
                  ylab = "Flow blast %",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  label = "sample_id",
                  cor.coef = FALSE,
                  label.select = blast_data$sample_id[blast_data$Model == "no"],
                  repel = TRUE,
                  font.label = c(8, "plain"),
                  size = 2) +
  geom_abline(slope = 1, lty = 2)

ggpubr::ggscatter(blast_data,
                  x = "Blast",
                  y = "blast_percent_vor",
                  color = "Model",
                  xlab = "CITEseq blast %",
                  ylab = "Flow blast %",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  cor.coef = TRUE,
                  cor.coef.coord = c(0, 100),
                  palette = c("lightgray", "orchid1"),
                  size = 2) +
  geom_abline(slope = 1, lty = 2)


```

# Compare ADT and FI quantiles
## All blasts
```{r adt_ab_quantiles_blast, fig.width = 12, fig.height = 8}

flow_list_ags_all_blasts <- flow_list[!grepl("_PE+|Quantibrite|LSC|Live|LP|PE_blasts", names(flow_list), ignore.case = TRUE)]

# compute FI quantiles
myprobs <- 0.05
df_list <- NULL
df_quantile_list <- NULL
df_bin_list <- NULL
for (ag in c("CD33", "CLL1", "CD123", "EMR2")) {
  
  flow_list_ags_select <- flow_list_ags_all_blasts[grepl(ag, names(flow_list_ags_all_blasts))]

  for (mynam in names(flow_list_ags_select)) {
    
    patient_id <- gsub(".*\\/", "", mynam)
    patient_id <- gsub("_.*", "", patient_id)
    
    # extract FIs from flowCore object
    fluor <- flow_list_ags_select[[mynam]]@parameters@data
    fluor <- as.character(fluor$name[fluor$desc %in% ag])
    fi <- exprs(flow_list_ags_select[[mynam]])
    fi <- fi[, colnames(fi) %in% fluor]
    if (length(fi) == 0) next
    
    myq <- quantile(fi, probs = seq(0, 1, myprobs), na.rm = TRUE) # compute quantiles
    myq_bin <- NULL
    for (q in 2:length(myq)) {
      myq_bin[paste0("bin_", q-1)] <- mean(fi[fi >= myq[q-1] & fi < myq[q]])
    }
    
    df_list[[mynam]] <- data.frame(patient_id, ag, fi) # raw FI values
    df_quantile_list[[mynam]] <- data.frame(patient_id, ag, quantile = names(myq), fi = myq) # FI quantiles
    df_bin_list[[mynam]] <- data.frame(patient_id, ag, quantile = names(myq_bin), fi = myq_bin)     

  }
}

df_list <- df_list[unlist(lapply(df_list, ncol)) == 3] 
df_quantile_list <- df_quantile_list[unlist(lapply(df_quantile_list, ncol)) == 4] 

# raw FIs
fi_df <- do.call("rbind", df_list)

fi_quant_df <- do.call("rbind", df_quantile_list) %>%
  dplyr::mutate(time = ifelse(grepl("D$", patient_id), "Diagnosis", "Relapse"),
                patient_id = gsub("-.*", "", patient_id),
                patient_id = gsub(" ", "_", patient_id),
                patient_id = paste(patient_id, time, sep = "_")) %>%
  dplyr::select(-time)

fi_mean_df <- do.call("rbind", df_bin_list) %>%
  dplyr::mutate(time = ifelse(grepl("D$", patient_id), "Diagnosis", "Relapse"),
                patient_id = gsub("-.*", "", patient_id),
                patient_id = gsub(" ", "_", patient_id),
                patient_id = paste(patient_id, time, sep = "_")) %>%
  dplyr::select(-time)


# add CD45 flow data (only available for 4 patient samples)
cd45_flow_paths <- c("s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/CD45_43-6D/export_B2 CD45 HI30_Blast.fcs",
                    "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/CD45_43-6R/export_CD45 HI30_Blast.fcs",
                    "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/CD45_70-6D/export_D2 CD45 HI30_Blast.fcs",
                    "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/CD45_70-6R/export_E2 CD45 HI30_Blast.fcs")


cd45_flow_list <- lapply(cd45_flow_paths, 
                         function(object_path) aws.s3::s3read_using(FUN = flowCore::read.FCS, 
                                                                    transformation = FALSE, 
                                                                    truncate_max_range = TRUE, 
                                                                    object = object_path))

mynam <- gsub("export.*", "", cd45_flow_paths)
mynam <- gsub("CD45_", "Patient_", basename(mynam))
mynam <- gsub("-6D", "_Diagnosis", mynam)
mynam <- gsub("-6R", "_Relapse", mynam)
names(cd45_flow_list) <- mynam

cd45_flow_list <- lapply(cd45_flow_list, function(arg) flowCore::transform(arg, `PE-A` = asinh(`PE-A`)))
cd45_flow_list <- lapply(cd45_flow_list, function(arg) exprs(arg)[, "PE-A"])
cd45_flow_list <- lapply(cd45_flow_list, function(arg) {
  tmp <- quantile(arg, probs = seq(0, 1, myprobs), na.rm = TRUE) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("quantile") %>%
    dplyr::rename(fi = ".") %>%
    dplyr::mutate(ag = "CD45")
  return(tmp)
  
})
fi_quant_df_cd45 <- dplyr::bind_rows(cd45_flow_list, .id = "patient_id")
fi_quant_df <- dplyr::bind_rows(fi_quant_df, fi_quant_df_cd45)

# quantile normalize DSB values

# class-specific normalization
# qn_data <- lapply(unique(blast_object$sample_id), function(myid) {
#   tmp_obj <- dplyr::filter(blast_object, sample_id %in% myid)
#   limma::normalizeQuantiles(tmp_obj@assays$ADT_dsb@data)
# })
# qn_data <- do.call("cbind", qn_data)

# whole dataset normalization
qn_data <- limma::normalizeQuantiles(blast_object@assays$ADT_dsb_ebayes@data)
blast_object[["ADT_qn"]] <- CreateAssayObject(data = qn_data)

blast_df <- blast_object@assays$ADT_dsb_ebayes@data[rownames(blast_object@assays$ADT) %in% c("CD33", "CLL1", "EMR2", "CD123", "CD45"), ]
# DefaultAssay(blast_object) <- "RNA"
# blast_df <- FetchData(blast_object, vars = c("CD33", "CLEC12A", "ADGRE2", "IL3RA", "PTPRC"), layer = "data")
# colnames(blast_df) <- c("CD33", "CLL1", "EMR2", "CD123", "CD45")
blast_df <- t(blast_df)

blast_list <- blast_df %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::mutate(patient_id = gsub(".*-", "", cell)) %>%
  dplyr::select(-cell) %>%
  tidyr::pivot_longer(-patient_id, names_to = "ag", values_to = "fi") %>%
  dplyr::group_split(patient_id, ag)

blast_list <- lapply(blast_list, function(arg) {
  
  myq <- quantile(arg$fi, probs = seq(0, 1, myprobs))
  #myq <- mean(arg$fi)
  mydf <- data.frame(patient_id = arg$patient_id[1], ag = arg$ag[1], quantile = names(myq), fi = myq)
  return(mydf)
  
})

blast_df <- do.call("rbind", blast_list)

fi_comdata <- dplyr::right_join(fi_quant_df, blast_df, 
                                by = c("patient_id", "ag", "quantile"), 
                                suffix = c("_flow", "_adt")) 
#%>%
  #dplyr::filter(grepl("Patient_70|Patient_43", patient_id))


# plot quantiles
trim_data_quantiles <- dplyr::filter(
  fi_comdata, 
  !is.na(fi_flow)
)

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "fi_adt",
                  y = "fi_flow",
                  facet.by = "patient_id",
                  color = "ag",
                  xlab = glue::glue("DSB normalized ADT quantiles ({myprobs})"),
                  ylab = glue::glue("log fluorescent intensity quantiles ({myprobs})"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 2),
                  size = 0.4,
                  palette = "lancet") + 
  geom_hline(yintercept = 3.25, lty = 2)

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "fi_adt",
                  y = "fi_flow",
                  color = "ag",
                  ylim = c(0, 12),
                  facet.by = "patient_id",
                  xlab = glue::glue("Median DSB normalized ADT"),
                  ylab = glue::glue("Median log fluorescent intensity"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 2),
                  size = 1.5,
                  palette = "lancet")


# plot quantiles for CD33 across samples
trim_data_quantiles <- dplyr::filter(
  fi_comdata, 
  !is.na(fi_flow)
)

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "fi_adt",
                  y = "fi_flow",
                  #ylim = c(0, 12),
                  color = "ag",
                  #facet.by = "patient_id",
                  xlab = glue::glue("DSB normalized ADT quantiles ({myprobs})"),
                  ylab = glue::glue("log fluorescent intensity quantiles ({myprobs})"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 6),
                  size = 1.5,
                  legend = "none")

# plot median ag expression
trim_data_median <- dplyr::filter(
  fi_comdata, 
  quantile %in% c("50%"),
  !is.na(fi_flow)
)

ggpubr::ggscatter(data = trim_data_median,
                  x = "fi_adt",
                  y = "fi_flow",
                  xlab = glue::glue("Median asinh DSB normalized ADT"),
                  ylab = glue::glue("Median asinh fluorescent intensity"),
                  face.by = "ag",
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 4),
                  size = 1.5,
                  palette = "lancet",
                  add = "reg.line")

ggpubr::ggscatter(data = trim_data_median,
                  x = "fi_adt",
                  y = "fi_flow",
                  color = "ag",
                  facet.by = "patient_id",
                  xlab = glue::glue("Median DSB normalized ADT"),
                  ylab = glue::glue("Median asinh fluorescent intensity"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 2.5),
                  size = 3,
                  palette = "lancet")


```


# Convert FI to antigen density using QuantiBRITE standard curve
```{r fi_vs_quantibrite}

qb_levels <- data.frame(peak = as.factor(1:4), ag_density = c(474, 5359, 23843, 62336))

plot_list <- lapply(names(flow_list_qb), function(arg) {
  data.frame(fcs_file = arg,
             PE = exprs(flow_list_qb[[arg]])[, "PE-A"])
})
names(plot_list) <- names(flow_list_qb)
plot_list <- plot_list[!names(plot_list) %in% "patients_47_48/export_A12 Quantibrite_beads"]
plot_df <- do.call("rbind", plot_list) %>%
  dplyr::mutate(fcs_file = gsub("\\/export.*", "", fcs_file))

rownames(plot_df) <- NULL

gghistogram(data = plot_df,
            x = "PE",
            xlim = c(0, 20),
            fill = "fcs_file",
            color = "white",
            bins = 100,
            legend = "right",
            palette = "simpsons")

ggdensity(data = plot_df,
          x = "PE",
          xlim = c(5, 18),
          xlab = "asinh PE FI",
          color = "fcs_file",
          legend = "right",
          palette = "lancet")


# compute PE peak means for each sample
set.seed(465012)
sample_qb_mean <- lapply(names(plot_list), function(arg) {
  
  mclust_model <- mclust::Mclust(data = plot_list[[arg]]$PE, G = 4)
  cell_class <- data.frame(FI = plot_list[[arg]]$PE, peak = mclust_model$classification)
  
  pe_mean <- sort(mclust_model$parameters$mean, decreasing = FALSE)
  pe_mean <- data.frame(fcs_file = arg, peak = 1:4, pe_mean = pe_mean)
  
  output_df <- dplyr::left_join(cell_class, pe_mean, by = "peak")
  
})

# look at mapping between PE and Ag count at single cell level
plot_df <- do.call("rbind", sample_qb_mean) %>%
  dplyr::mutate(peak = as.factor(peak)) %>%
  dplyr::left_join(qb_levels, by = "peak") %>%
  dplyr::mutate(fcs_file = gsub("\\/export.*", "", fcs_file),
                fcs_file = dplyr::case_when(fcs_file == "patients_26_27_28_29" ~ "P01, P12, P18, P24",
                                            fcs_file == "patients_30_31_67_70" ~ "P04, P06, P08, P19",
                                            fcs_file == "patients_40_41" ~ "P13, P25",
                                            fcs_file == "patients_49_50" ~ "P14, P16",
                                            fcs_file == "patients_4_42_43_44" ~ "P02, P07, P20, P26",
                                            fcs_file == "patients_51_52_54_56" ~ "P03, P10, P11, P15",
                                            fcs_file == "patients_58_64_65_66" ~ "P09, P17, P21, P23",
                                            fcs_file == "patients_60_61_62_63" ~ "P05, P22"))

ggplot(plot_df, aes(x = FI, y = ag_density)) +
  geom_point(shape = 108, size = 3, alpha = 0.1, aes(color = fcs_file)) +
  xlim(c(4, 17)) +
  ylim(c(50, 150000)) +
  xlab("asinh PE fluorescent intensity") +
  ylab("QuantiBRITE bead Ag number") +
  facet_wrap(~fcs_file, ncol = 3) +
  geom_smooth(se = FALSE, method = "lm", lty = 2, width = 0.05, color = "gray") +
  scale_y_log10(labels = scales::comma, breaks = c(100, 1000, 10000, 100000)) +
  scale_color_lancet() +
  theme_bw() +
  theme(axis.text = element_text(size = 10))

# keep means
plot_df <- plot_df %>%
  dplyr::select(-FI) %>%
  dplyr::distinct()

ggscatter(data = plot_df,
          x = "pe_mean",
          y = "ag_density",
          ylim = c(50, 100000),
          xlim = c(8, 15.5),
          xlab = "Mean asinh PE intensity",
          ylab = "Log antigen density",
          color = "fcs_file",
          legend = "right",
          palette = "lancet") +
  scale_y_log10(labels = scales::comma) +
  geom_abline(slope = 0.4501, intercept = -1.8157, lty = 2) +
  geom_hline(yintercept = 50, lty = 3) +
  geom_hline(yintercept = 450, lty = 3)

ggscatter(data = plot_df,
          x = "pe_mean",
          y = "ag_density",
          ylim = c(50, 100000),
          xlim = c(8, 15.5),
          xlab = "Mean asinh PE intensity",
          ylab = "Log antigen density",
          facet.by = "fcs_file",
          color = "fcs_file",
          legend = "right",
          palette = "lancet",
          add = "reg.line",
          add.params = list(linetype = 2, size = 0.1)) +
  scale_y_log10(labels = scales::comma)

qb_models <- plot_df %>% 
  dplyr::group_by(fcs_file) %>%
  dplyr::group_split()

mynam <- sapply(qb_models, function(arg) unique(arg$fcs_file))
qb_models <- lapply(qb_models, function(arg) lm(log10(ag_density) ~ pe_mean, data = arg))
names(qb_models) <- mynam

aggr_qb_model <- lm(log10(ag_density) ~ pe_mean, data = plot_df)

```

# Compare ADT and ABC quantiles
## Calculate FMO Ag density boundaries for all samples
```{r fmo_ag_density}

fmo_boundary <- NULL
for (mynam in names(flow_list_fmo)) {
    
    model_id <- gsub("\\/.*", "", mynam)
    fcs_file <- gsub("\\/.*", "", mynam)
    fi <- exprs(flow_list_fmo[[mynam]])
    fi <- fi[, colnames(fi) %in% "PE-A"]
    
    abc_model <- qb_models[[model_id]]
    abc <- predict(object = abc_model, newdata = data.frame(pe_mean = fi))
    abc <- 10^abc
    upper_boundary <- mean(abc) + 3*sd(abc)
    patient_id <- gsub("patients_", "", fcs_file)
    patient_id <- unlist(strsplit(patient_id, split = "_"))
    patient_id <- paste0("Patient_", patient_id)
    
    fmo_boundary[[fcs_file]] <- data.frame(fcs_file = fcs_file, patient = patient_id, fmo_ag_density_boundary = upper_boundary)
    
}

fmo_df <- do.call("rbind", fmo_boundary) %>%
  dplyr::arrange(fmo_ag_density_boundary)

```

## All blasts per sample per antigen
```{r adt_ag_quantiles_blast, fig.width = 8, fig.height = 7}

flow_list_ags_all_blasts <- flow_list[!grepl("_PE+|Quantibrite|LSC|export|Live", names(flow_list), ignore.case = TRUE)]

file_count <- names(flow_list_ags_all_blasts)
file_count <- gsub(".*/", "", file_count)
file_count <- gsub("_.*", "", file_count)
table(file_count) %>% sort()
stopifnot(all(table(file_count) == 4))

# compute Ag quantiles
myprobs <- 0.05
df_list <- NULL
df_quantile_list <- NULL
df_mean_list <- NULL
df_median_list <- NULL
for (ag in c("CD33", "CLL1", "CD123", "EMR2")) {
  
  flow_list_ags_select <- flow_list_ags_all_blasts[grepl(ag, names(flow_list_ags_all_blasts))]

  for (mynam in names(flow_list_ags_select)) {
    
    model_id <- gsub("\\/.*", "", mynam)
    sample_id <- gsub(".*\\/", "", mynam)
    sample_id <- gsub("_.*", "", sample_id)
    
    # extract FIs from flowCore object
    fluor <- flow_list_ags_select[[mynam]]@parameters@data
    fluor <- as.character(fluor$name[fluor$desc %in% ag])
    if (length(fluor) == 0) fluor = "PE-A"
    fi <- exprs(flow_list_ags_select[[mynam]])
    fi <- fi[, colnames(fi) %in% fluor]
    
    abc_model <- qb_models[[model_id]]
    abc <- predict(object = abc_model, newdata = data.frame(pe_mean = fi))
    
    if (length(abc) == 0) next
    
    # remove negative cells
    #abc <- abc[abc > 1]
    # compute quantiles
    # myclip <- quantile(abc, probs = c(0.1, 0.99))
    # abc <- abc[abc > myclip[1] & abc < myclip[2]]
    myq <- quantile(abc, probs = seq(0, 1, myprobs), na.rm = TRUE)
    df_list[[mynam]] <- data.frame(sample_id, ag, abc) # raw FI values
    df_quantile_list[[mynam]] <- data.frame(sample_id, ag, quantile = names(myq), abc = myq) # ABC quantiles
    df_median_list[[mynam]] <- data.frame(sample_id, ag, quantile = "50%", abc = median(abc))
    
    # quantile-binned averages
    myq_bin <- NULL
    for (q in 2:length(myq)) {
      myq_bin[paste0("bin_", q-1)] <- mean(abc[abc >= myq[q-1] & abc < myq[q]])
    }
      
    df_mean_list[[mynam]] <- data.frame(sample_id, ag, quantile_bin = names(myq_bin), abc = myq_bin) 

  }
}

# raw FIs
fi_df <- do.call("rbind", df_list)

fi_quant_df <- do.call("rbind", df_quantile_list) %>%
  dplyr::mutate(time = ifelse(grepl("D$", sample_id), "Diagnosis", "Relapse"),
                sample_id = gsub("-.*", "", sample_id),
                sample_id = gsub(" ", "_", sample_id),
                sample_id = paste(sample_id, time, sep = "_")) %>%
  dplyr::select(-time)

# quantile normalize adt expression for input into machine learning model
qn_data <- blast_object@assays$ADT_dsb_ebayes@data
qn_data <- limma::normalizeQuantiles(qn_data)
blast_object[["ADT_qn"]] <- CreateAssayObject(data = qn_data)

blast_df <- blast_object@assays$ADT_qn@data[rownames(blast_object@assays$ADT_qn) %in% c("CD33", "CLL1", "EMR2", "CD123"), ]
blast_df <- t(blast_df)

blast_list <- blast_df %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::mutate(sample_id = gsub(".*-", "", cell)) %>%
  dplyr::select(-cell) %>%
  tidyr::pivot_longer(-sample_id, names_to = "ag", values_to = "abc") %>%
  dplyr::group_split(sample_id, ag)

blast_list <- lapply(blast_list, function(arg) {
  
  adt_abc <- arg$abc
  myq <- quantile(adt_abc, probs = seq(0, 1, myprobs))
  mydf <- data.frame(sample_id = arg$sample_id[1], ag = arg$ag[1], quantile = names(myq), abc = myq)
  return(mydf)
  
})

blast_quantiles_df <- do.call("rbind", blast_list)

fi_comdata <- dplyr::right_join(fi_quant_df, blast_quantiles_df, 
                                by = c("sample_id", "ag", "quantile"), 
                                suffix = c("_flow", "_adt")) %>%
  dplyr::mutate(quantile = factor(quantile, levels = unique(.$quantile))) %>% 
  dplyr::mutate(ag_density = 10^abc_flow,
                ag_density = ifelse(ag_density < 1, 0, ag_density)) %>%
  # no flow data for samples "Patient_31_Relapse"   "Patient_52_Diagnosis" "Patient_54_Diagnosis" "Patient_61_Relapse" 
  dplyr::filter(!is.na(abc_flow))

# plot median ag expression
trim_data_median <- dplyr::filter(fi_comdata, quantile %in% c("50%")) 

ggpubr::ggscatter(data = trim_data_median,
                  x = "abc_adt",
                  y = "ag_density",
                  xlab = glue::glue("DSB normalized ADT"),
                  ylab = glue::glue("Antigen density"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 4),
                  cor.coef.coord = c(-1, 4),
                  size = 3,
                  legend = "right") +
  geom_hline(yintercept = min(fmo_df$fmo_ag_density_boundary), lty = 3) +
  geom_hline(yintercept = max(fmo_df$fmo_ag_density_boundary), lty = 3) +
  geom_hline(yintercept = mean(fmo_df$fmo_ag_density_boundary), lty = 2) +
  scale_y_log10(breaks = c(1, 10, 100, 1000, 10000, 100000), labels = c("1", "10", "100", "1,000", "10,000", "100,000")) +
  scale_x_continuous(breaks = -1:15) +
  geom_vline(xintercept = 0, lty = 3) +
  geom_smooth(method = "lm")

ggpubr::ggscatter(data = trim_data_median,
                  x = "abc_adt",
                  y = "abc_flow",
                  xlab = glue::glue("Median asinh DSB normalized ADT"),
                  ylab = glue::glue("Median log10 antigen density"),
                  facet.by = "sample_id",
                  color = "ag",
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 2),
                  size = 1.5,
                  palette = "lancet")

# plot quantiles across samples
trim_data_quantiles <- fi_comdata %>%
  dplyr::mutate(patient = gsub("_Diagnosis|_Relapse", "", sample_id)) %>%
  dplyr::left_join(fmo_df, by = "patient") %>%
  # dplyr::mutate(ag = case_when(ag %in% "EMR2" ~ "ADGRE2",
  #                              ag %in% "CLL1" ~ "CLL-1",
  #                              TRUE ~ ag)) %>%
  dplyr::mutate(ag = factor(ag, levels = c("CD33", "CLL1", "CD123", "EMR2")),
                quantile_group = ifelse(!quantile %in% c("0%", "5%", "10%"), ">10%", as.character(quantile)),
                #quantile_group = ifelse(quantile_group == "0%", ">1%", quantile_group),
                quantile_group = factor(quantile_group, levels = c(">10%", "10%", "5%", "0%")))

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "abc_adt",
                  y = "abc_flow",
                  color = "ag",
                  xlab = glue::glue("DSB normalized ADT quantiles ({myprobs})"),
                  ylab = glue::glue("log10 antigen density quantiles ({myprobs})"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  facet.by = "ag",
                  cor.coeff.args = list(size = 2),
                  size = 1.5)

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "abc_adt",
                  y = "ag_density",
                  alpha = 0.3,
                  color = "sample_id",
                  xlab = glue::glue("Normalized ADT"),
                  ylab = glue::glue("Antigen density"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 4),
                  size = 2,
                  legend = "none",
                  facet.by = "ag") +
  geom_hline(yintercept = min(fmo_df$fmo_ag_density_boundary), lty = 3) +
  geom_hline(yintercept = max(fmo_df$fmo_ag_density_boundary), lty = 3) +
  geom_hline(yintercept = mean(fmo_df$fmo_ag_density_boundary), lty = 1) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10),
                     breaks = c(0, 1, 10, 100, 1000, 10000, 100000), 
                     labels = c("0", "1", "10", "100", "1,000", "10,000", "100,000")) +
  scale_x_continuous(breaks = seq(-1, 20, 2)) +
  geom_vline(xintercept = 0, lty = 3) +
  stat_density_2d(alpha = 0.3, color = "black") +
  geom_smooth(se = FALSE)

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "abc_adt",
                  y = "ag_density",
                  alpha = 0.3,
                  color = "quantile_group",
                  xlab = glue::glue("Normalized ADT"),
                  ylab = glue::glue("Antigen density"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 4),
                  #cor.coef.coord = c(-1.1, 10000),
                  size = 2,
                  legend = "right",
                  facet.by = "ag",
                  palette = c("aquamarine4", "blue", "orange", "red")) +
  geom_hline(yintercept = min(fmo_df$fmo_ag_density_boundary), lty = 3) +
  geom_hline(yintercept = max(fmo_df$fmo_ag_density_boundary), lty = 3) +
  geom_hline(yintercept = mean(fmo_df$fmo_ag_density_boundary), lty = 1) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10),
                     breaks = c(0, 1, 10, 100, 1000, 10000, 100000), 
                     labels = c("0", "1", "10", "100", "1,000", "10,000", "100,000")) +
  scale_x_continuous(breaks = c(seq(-1, 10, 1), seq(12, 20, 2))) +
  geom_vline(xintercept = 1, lty = 3) +
  stat_density_2d(alpha = 0.3, color = "black") +
  geom_smooth(se = FALSE)


ggpubr::ggscatter(data = dplyr::mutate(trim_data_quantiles, 
                                       ag = factor(case_when(ag == "CLL1" ~ "CLL-1",
                                                             ag == "EMR2" ~ "ADGRE2",
                                                             TRUE ~ as.character(ag)), levels = c("CD33", "CLL-1", "CD123", "ADGRE2"))),
                  x = "abc_adt",
                  y = "ag_density",
                  alpha = 0.3,
                  color = "ag",
                  facet.by = "ag",
                  xlab = glue::glue("Normalized ADT"),
                  ylab = glue::glue("QuantiBRITE antigen number"),
                  size = 0.1,
                  legend = "right",
                  palette = c("skyblue3", "hotpink", "purple", "olivedrab3")) +
  #geom_hline(yintercept = mean(fmo_df$fmo_ag_density_boundary), lty = 1) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10),
                     breaks = c(0, 10, 100, 1000, 10000, 100000), 
                     labels = c("0", "10", "100", "1,000", "10,000", "100,000")) +
  scale_x_continuous(breaks = c(-1, 0, 1, 2, 5, 10, 15, 20)) +
  #geom_vline(xintercept = 1, lty = 3) +
  stat_density_2d(alpha = 0.3, color = "black") +
  geom_smooth(se = TRUE, fill = "gray", colour = "black")


ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "abc_adt",
                  y = "abc_flow",
                  color = "sample_id",
                  alpha = 0.5,
                  xlab = glue::glue("DSB normalized ADT"),
                  ylab = glue::glue("log10 antigen density"),
                  cor.coef = TRUE,
                  cor.method = "pearson",
                  cor.coeff.args = list(size = 2),
                  size = 1,
                  legend = "none") +
  geom_vline(xintercept = 0, lty = 3)

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "abc_adt",
                  y = "ag_density",
                  ylim = c(10, 100000),
                  color = "ag",
                  facet.by = "sample_id",
                  xlab = glue::glue("Normalized ADT"),
                  ylab = glue::glue("Antigen density"),
                  cor.coef = FALSE,
                  cor.method = "spearman",
                  cor.coeff.args = list(size = 2),
                  size = 0.4) +
  scale_y_log10(breaks = c(10, 100, 1000, 10000, 100000), labels = c("10", "100", "1,000", "10,000", "100,000"))

ggpubr::ggscatter(data = dplyr::filter(trim_data_quantiles, sample_id %in% c("Patient_26_Diagnosis", "Patient_26_Relapse", "Patient_40_Relapse", "Patient_41_Relapse")),
                  x = "abc_adt",
                  y = "ag_density",
                  color = "ag",
                  facet.by = "sample_id",
                  xlab = glue::glue("DSB normalized ADT"),
                  ylab = glue::glue("log10 antigen density"),
                  cor.coef = TRUE,
                  cor.method = "spearman",
                  cor.coeff.args = list(size = 2),
                  size = 3) +
  geom_vline(xintercept = 0, lty = 3) + 
  scale_y_log10()

summary(lm(abc_flow ~ abc_adt, data = trim_data_median))
summary(lm(abc_flow ~ abc_adt, data = trim_data_quantiles))


```

# Fit model and estimate ABCs for all antigens in blasts
## Use sample-feature quantiles
```{r predict_ag_using_quantiles_as_training_data, fig.height = 9, fig.width = 10}

ab_data <- aws.s3::s3read_using(FUN = readr::read_csv,
                                object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/aml_atlas_sample_metadata.csv") %>%
  dplyr::select(sample_id, contains("antibody")) %>%
  dplyr::distinct() %>%
  dplyr::mutate_at(vars(contains("antibody")), .funs = as.numeric)

training_data <- trim_data_quantiles %>%
  dplyr::left_join(ab_data, by = "sample_id") %>%
  dplyr::mutate(patient = gsub("_Diagnosis|_Relapse", "", sample_id))

training_data_for_model <- dplyr::filter(training_data, ag %in% c("CD33", "CLL1", "EMR2", "CD123")) %>%
  dplyr::select(-fcs_file, -patient, -sample_id, -ag, -quantile, -quantile_group, -ag_density, -fmo_ag_density_boundary)

stopifnot(all(complete.cases(training_data_for_model)))

# train model
set.seed(391622) 
fitControl <- caret::trainControl(method = "repeatedcv", number = 10, repeats = 10)
mymethod <- "rf"
gbmFit1 <- caret::train(
  
  abc_flow ~ .,
  data = training_data_for_model,
  method = mymethod,
  trControl = fitControl,
  verbose = FALSE

)

gbmFit1

input_data <- blast_object@assays$ADT_qn@data %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::mutate(sample_id = gsub(".*-", "", barcode)) %>%
  dplyr::left_join(ab_data, by = "sample_id") %>%
  dplyr::mutate(patient = gsub(".*-|_Diagnosis|_Relapse", "", barcode)) %>%
  dplyr::left_join(fmo_df, by = "patient")

input_data <- input_data[complete.cases(input_data), ]
myfeats <- rownames(blast_object@assays$ADT_qn@data)

# save copy of model
sample_data <- dplyr::select(input_data, sample_id, CD33, contains("antibody")) %>%
  dplyr::rename(abc_adt = CD33)
sample_data <- sample_data[sample(1:nrow(input_data), 100, replace = FALSE), ]
rownames(sample_data) <- paste0("blast_", 1:nrow(sample_data))

# predict antigen density
mypred <- predict(gbmFit1, newdata = sample_data)
mypred <- 10^mypred
ag_density_prediction_model <- list(rf_model = gbmFit1, sample_test_data_CD33 = sample_data, predicted_ag_density = data.frame(CD33_estimated_ag_density = mypred, sample_data))

aws.s3::s3saveRDS(x = ag_density_prediction_model,
                  object = glue::glue("s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/antigen_density_prediction_models/{mymethod}_model_{Sys.Date()}.rds"))

tic()

# load pre-trained model
#mymodel <- aws.s3::s3readRDS(object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/antigen_density_prediction_models/rf_model_2024-09-24_ASH_2024_poster.rds")[[1]]
mymodel <- gbmFit1
imputed_abc <- foreach(arg = myfeats, .combine = "cbind", .inorder = TRUE, .packages = c("randomForest", "e1071",  "dplyr", "caret")) %dopar% {
  
  newdata <- dplyr::select(input_data, all_of(arg), contains("antibody"))
  colnames(newdata)[colnames(newdata) == arg] <- "abc_adt"
  mypred <- predict(mymodel, newdata = newdata)
  mypred <- 10^mypred
  mypred <- data.frame(mypred)
  colnames(mypred) <- arg
  gc()
  return(mypred)
  
}
toc()
rownames(imputed_abc) <- input_data$barcode

aws.s3::s3write_using(imputed_abc, 
                      FUN = write.csv, 
                      object = glue::glue("s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/imputed_ag_density/imputed_ag_density_blasts_{Sys.Date()}.csv"))

#sanity checks
iso_df <- dplyr::select(input_data, starts_with("ISOTYPE"))
iso_df <- data.frame(ADT_type = "Isotype control", mean_ag = rowMeans(iso_df))

myeloid_df  <- dplyr::select(input_data, !starts_with("ISOTYPE"))
myeloid_df <- data.frame(ADT_type = "Myeloid antibody", mean_ag = rowMeans(myeloid_df[, c("CD33", "CLL1", "CD123", "EMR2")]))

lym_df  <- dplyr::select(input_data, !starts_with("ISOTYPE"))
lym_df <- data.frame(ADT_type = "Lymphoid antibody", mean_ag = rowMeans(lym_df[, colnames(lym_df) %in% c("CD3", "CD4", "CD8", "CD19", "CD22")]))

ab_df  <- dplyr::select(input_data, !starts_with("ISOTYPE")) %>%
  dplyr::select(!starts_with("antibody"), -sample_id, -barcode, -patient, -fcs_file, -fmo_ag_density_boundary)
ab_df <- data.frame(ADT_type = "Other antibody", mean_ag = rowMeans(ab_df[, !colnames(ab_df) %in% c("CD33", "CLL1", "CD123", "EMR2", "CD3", "CD4", "CD8", "CD19", "CD22")]))

sanity_data <- bind_rows(myeloid_df, lym_df, ab_df, iso_df) %>%
  dplyr::filter(!is.na(mean_ag)) %>%
  dplyr::mutate(ADT_type = factor(ADT_type, levels = c("Myeloid antibody", "Lymphoid antibody", "Other antibody", "Isotype control")))

gghistogram(sanity_data, 
            x = "mean_ag", 
            xlab = "Average normalized ADT",
            ylab = "Number of cells",
            xlim = c(-1, 10),
            fill = "ADT_type", 
            color = "white", 
            bins = 100,
            palette = c("cornflowerblue", "forestgreen", "purple", "black"))

iso_df <- dplyr::select(imputed_abc, starts_with("ISOTYPE"))
iso_df <- data.frame(ADT_type = "Isotype control", mean_ag = rowMeans(iso_df))

myeloid_df  <- dplyr::select(imputed_abc, !starts_with("ISOTYPE"))
myeloid_df <- data.frame(ADT_type = "Myeloid antibody", mean_ag = rowMeans(myeloid_df[, c("CD33", "CLL1", "CD123", "EMR2")]))

lym_df  <- dplyr::select(imputed_abc, !starts_with("ISOTYPE"))
lym_df <- data.frame(ADT_type = "Lymphoid antibody", mean_ag = rowMeans(lym_df[, colnames(lym_df) %in% c("CD3", "CD4", "CD8", "CD19", "CD22")]))

ab_df  <- dplyr::select(imputed_abc, !starts_with("ISOTYPE"))
ab_df <- data.frame(ADT_type = "Other antibody", mean_ag = rowMeans(ab_df[, !colnames(ab_df) %in% c("CD33", "CLL1", "CD123", "EMR2", "CD3", "CD4", "CD8", "CD19", "CD22")]))

sanity_data <- bind_rows(myeloid_df, lym_df, ab_df, iso_df) %>%
  dplyr::filter(!is.na(mean_ag)) %>%
  dplyr::mutate(ADT_type = factor(ADT_type, levels = c("Myeloid antibody", "Lymphoid antibody", "Other antibody", "Isotype control")))

gghistogram(sanity_data, 
            x = "mean_ag", 
            xlab = "Average predicted antigen density",
            ylab = "Number of cells",
            xlim = c(0, 10000),
            fill = "ADT_type", 
            color = "white", 
            bins = 200,
            palette = c("cornflowerblue", "forestgreen", "purple", "black")) +
  geom_vline(xintercept = 500, lty = 2)

cd33_df  <- dplyr::select(imputed_abc, "CD33")

gghistogram(cd33_df, 
            x = "CD33", 
            main = "Blasts",
            xlab = "Esimated # of CD33 Ags",
            ylab = "Number of cells",
            #xlim = c(0, 60000),
            fill = "blue", 
            bins = 50) +
  geom_vline(xintercept = 500, lty = 2)

# Compare estimated Ag+ rate with flow + FMO
imputed_abc_coverage <- calculate_coverage(imputed_abc_df = imputed_abc, 
                                           fmo_threshold_df = fmo_df, 
                                           first_antigen = "CD33", 
                                           second_antigen = "CLL1") %>%
  tidyr::unite("sample_id", patient, time)

bpos_df <- aws.s3::s3read_using(readr::read_csv,
                                object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/percent_positive_blasts.csv") %>%
  dplyr::mutate(time = ifelse(grepl("D$", vor_patient_id), "Diagnosis", "Relapse"),
                patient = paste0("Patient_", gsub("-.*", "", vor_patient_id))) %>%
  dplyr::select(patient, time, everything()) %>%
  dplyr::select(-vor_patient_id) %>%
  tidyr::pivot_longer(cols = c(CD33, CD123, CLL1, EMR2), values_to = "percent_positive_blast", names_to = "adt") %>%
  tidyr::unite(col = "sample_id", patient, time)

bpos_df_wide <- tidyr::pivot_wider(bpos_df, names_from = "adt", values_from = "percent_positive_blast")

compare_df <- dplyr::left_join(imputed_abc_coverage, bpos_df_wide, by = "sample_id") %>%
  dplyr::rename(percent_cd33 = percent_first_ag, percent_cll1 = percent_second_ag)

p_cd33 <- ggpubr::ggscatter(data = compare_df,
                  main = "% CD33+ blasts",
                  x = "percent_cd33",
                  y = "CD33",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  xlab = "Predicted % positive",
                  ylab = "Flow % positive",
                  repel = TRUE,
                  cor.coef = TRUE) +
  geom_abline(intercept = 0, slope = 1, lty = 2)

p_cll1 <- ggpubr::ggscatter(data = compare_df,
                  main = "% CLL-1 blasts", 
                  x = "percent_cll1",
                  y = "CLL1",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  xlab = "Predicted % positive",
                  ylab = "Flow % positive",
                  repel = TRUE,
                  cor.coef = TRUE) +
  geom_abline(intercept = 0, slope = 1, lty = 2)

# CD123 and EMR2
imputed_abc_coverage <- calculate_coverage(imputed_abc_df = imputed_abc, 
                                           fmo_threshold_df = fmo_df, 
                                           first_antigen = "CD123", 
                                           second_antigen = "EMR2") %>%
  tidyr::unite("sample_id", patient, time)

bpos_df_wide <- tidyr::pivot_wider(bpos_df, names_from = "adt", values_from = "percent_positive_blast")

compare_df <- dplyr::left_join(imputed_abc_coverage, bpos_df_wide, by = "sample_id") %>%
  dplyr::rename(percent_cd123 = percent_first_ag, percent_emr2 = percent_second_ag)

p_cd123 <- ggpubr::ggscatter(data = compare_df,
                  main = "% CD123+ blasts", 
                  x = "percent_cd123",
                  y = "CD123",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  xlab = "Predicted % positive",
                  ylab = "Flow % positive",
                  repel = TRUE,
                  cor.coef = TRUE,
                  cor.coef.coord = c(0, 90)) +
  geom_abline(intercept = 0, slope = 1, lty = 2)

p_emr2 <- ggpubr::ggscatter(data = compare_df,
                  main = "% ADGRE2+ blasts", 
                  x = "percent_emr2",
                  y = "EMR2",
                  xlim = c(1, 100),
                  ylim = c(1, 100),
                  xlab = "Predicted % positive",
                  ylab = "Flow % positive",
                  repel = TRUE,
                  cor.coef = TRUE,
                  cor.coef.coord = c(0, 90)) +
  geom_abline(intercept = 0, slope = 1, lty = 2)


(p_cd33 + p_cll1) / (p_cd123 + p_emr2)


```

## Leave one antigen out analysis
```{r leave_one_ag_out_analysis}

ags <- c("CD33", "CLL1", "EMR2", "CD123", "control")
plot_list <- list()
imputed_abc_ag_out_list <- list()
set.seed(38347)  
for (ag_out in ags) {
  
  training_data_for_model <- dplyr::filter(training_data, !(ag %in% c(ag_out))) %>%
    dplyr::select(-fcs_file, -patient, -sample_id, -ag, -quantile, -quantile_group, -ag_density, -fmo_ag_density_boundary)
  
  stopifnot(all(complete.cases(training_data_for_model)))
  
  # Random forests
  set.seed(85630)                                                         
  fitControl <- caret::trainControl(method = "repeatedcv",
                                    number = 10,
                                    ## repeated ten times
                                    repeats = 10)
  
  mymethod <- "rf"
  gbmFit1 <- caret::train(abc_flow ~ .,
                          data = training_data_for_model,
                          method = mymethod,
                          trControl = fitControl,
                          verbose = FALSE)
  
  # fit model
  tic()
  imputed_abc_ag_out <- foreach(arg = myfeats, .combine = "cbind", .inorder = TRUE, .packages = c("randomForest", "e1071",  "dplyr", "caret")) %dopar% {
    
    newdata <- dplyr::select(input_data, all_of(arg), contains("antibody"))
    colnames(newdata)[colnames(newdata) == arg] <- "abc_adt"
    mypred <- predict(gbmFit1, newdata = newdata)
    mypred <- 10^mypred
    mypred <- data.frame(mypred)
    colnames(mypred) <- arg
    gc()
    return(mypred)
    
  }
  toc()
  rownames(imputed_abc_ag_out) <- input_data$barcode
  
  imputed_abc_ag_out_list[[ag_out]] <- imputed_abc_ag_out
  
  # Compare estimated Ag+ rate with flow + FMO
  imputed_abc_ag_out_coverage <- calculate_coverage(imputed_abc_df = imputed_abc_ag_out, 
                                                    fmo_threshold_df = fmo_df, 
                                                    first_antigen = "CD33", 
                                                    second_antigen = "CLL1") %>%
    tidyr::unite("sample_id", patient, time)
  
  compare_df <- dplyr::left_join(imputed_abc_ag_out_coverage, bpos_df_wide, by = "sample_id") %>%
    dplyr::rename(percent_cd33 = percent_first_ag, percent_cll1 = percent_second_ag)
  
  p_cd33 <- ggpubr::ggscatter(data = compare_df,
                    main = "% CD33+ blasts",
                    x = "percent_cd33",
                    y = "CD33",
                    xlim = c(1, 100),
                    ylim = c(1, 100),
                    xlab = "Predicted % positive",
                    ylab = "Flow % positive",
                    repel = TRUE,
                    cor.coef = TRUE) +
    geom_abline(intercept = 0, slope = 1, lty = 2)
  
  p_cll1 <- ggpubr::ggscatter(data = compare_df,
                    main = "% CLL-1 blasts", 
                    x = "percent_cll1",
                    y = "CLL1",
                    xlim = c(1, 100),
                    ylim = c(1, 100),
                    xlab = "Predicted % positive",
                    ylab = "Flow % positive",
                    repel = TRUE,
                    cor.coef = TRUE) +
    geom_abline(intercept = 0, slope = 1, lty = 2)
  
  # CD123 and EMR2
  imputed_abc_coverage <- calculate_coverage(imputed_abc_df = imputed_abc, 
                                             fmo_threshold_df = fmo_df, 
                                             first_antigen = "CD123", 
                                             second_antigen = "EMR2") %>%
    tidyr::unite("sample_id", patient, time)
  
  bpos_df_wide <- tidyr::pivot_wider(bpos_df, names_from = "adt", values_from = "percent_positive_blast")
  
  compare_df <- dplyr::left_join(imputed_abc_coverage, bpos_df_wide, by = "sample_id") %>%
    dplyr::rename(percent_cd123 = percent_first_ag, percent_emr2 = percent_second_ag)
  
  p_cd123 <- ggpubr::ggscatter(data = compare_df,
                    main = "% CD123+ blasts", 
                    x = "percent_cd123",
                    y = "CD123",
                    xlim = c(1, 100),
                    ylim = c(1, 100),
                    xlab = "Predicted % positive",
                    ylab = "Flow % positive",
                    repel = TRUE,
                    cor.coef = TRUE,
                    cor.coef.coord = c(0, 90)) +
    geom_abline(intercept = 0, slope = 1, lty = 2)
  
  p_emr2 <- ggpubr::ggscatter(data = compare_df,
                    main = "% ADGRE2+ blasts", 
                    x = "percent_emr2",
                    y = "EMR2",
                    xlim = c(1, 100),
                    ylim = c(1, 100),
                    xlab = "Predicted % positive",
                    ylab = "Flow % positive",
                    repel = TRUE,
                    cor.coef = TRUE,
                    cor.coef.coord = c(0, 90)) +
    geom_abline(intercept = 0, slope = 1, lty = 2)
  
  if (ag_out == "CD33") {
    plot_list[[ag_out]] <- p_cd33
  } else if (ag_out == "CLL1") {
    plot_list[[ag_out]] <- p_cll1
  } else if (ag_out == "EMR2") {
    plot_list[[ag_out]] <- p_emr2
  } else {
    plot_list[[ag_out]] <- p_cd123
  }

}

(plot_list[["CD33"]] + plot_list[["CLL1"]]) / (plot_list[["CD123"]] + plot_list[["EMR2"]])

```

# Compare estimated Ag density with flow
## Original flow data plots
```{r compare_estimated_ag_density}

mypath <- "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/20240425_Percent Positivity Matched Blasts Diagnosis vs  Relapse bone marrow .csv"
flow_ag_data_percent <- aws.s3::s3read_using(readr::read_csv, object = mypath)

colnames(flow_ag_data_percent) <- gsub(" ", "_", colnames(flow_ag_data_percent))
flow_ag_data_percent <- flow_ag_data_percent %>%
  dplyr::mutate(patient= paste0("Patient_", patient),) %>%
  tidyr::pivot_longer(-patient, names_to = "sample", values_to = "flow_antigen_percent_positive") %>%
  tidyr::separate(sample, c("ag", "timepoint")) %>%
  dplyr::mutate(ag = factor(ag, levels = c("CD33", "CLL1", "CD123", "EMR2")),
                timepoint = ifelse(timepoint == "Diagnosis", "Dx", "Re"))

select_pat1 <- dplyr::filter(flow_ag_data_percent, ag == "CD33", timepoint == "Re", flow_antigen_percent_positive < 50)$patient
select_pat2 <- dplyr::filter(flow_ag_data_percent, ag == "CD33", timepoint == "Dx", flow_antigen_percent_positive < 50)$patient
select_pat3 <- dplyr::filter(flow_ag_data_percent, ag == "CLL1", timepoint == "Re", flow_antigen_percent_positive > 50)$patient
select_pat4 <- dplyr::filter(flow_ag_data_percent, ag == "CLL1", timepoint == "Dx", flow_antigen_percent_positive > 50)$patient
select_pat <- intersect(intersect(intersect(select_pat1, select_pat2), select_pat3), select_pat4)

# flow_ag_data_percent <- flow_ag_data_percent %>%
#   dplyr::mutate(highlight = factor(ifelse(patient %in% select_pat, "yes", "none"), levels = c("none", "yes")))
flow_ag_data_percent <- flow_ag_data_percent %>%
  dplyr::mutate(highlight = factor(ifelse(patient %in% select_pat, patient, "none"), levels = c("none", select_pat)))

ggboxplot(data = flow_ag_data_percent,
          x = "timepoint",
          y = "flow_antigen_percent_positive",
          xlab = "",
          ylab = "% Antigen+ on AML Blasts",
          ylim = c(0, 100),
          color = "black",
          fill = "white",
          outlier.shape = NA,
          legend.position = "none",
          palette = "lancet") +
  geom_point(aes(color = ag), position = position_jitter(seed = 42, width = 0.2)) +
  facet_grid(~ag) +
  theme_bw() + 
  theme(legend.position = "none", 
        axis.text = element_text(size = 15), 
        axis.title = element_text(size = 17), 
        strip.text = element_text(size = 15),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        strip.background = element_rect(fill="moccasin")) +
  scale_y_continuous(labels = c("0%", "25%", "50%", "75%", "100%")) +
  scale_color_manual(values = c("firebrick2", "cornflowerblue", "mediumorchid1", "seagreen3"))

ggboxplot(data = flow_ag_data_percent,
          x = "timepoint",
          y = "flow_antigen_percent_positive",
          xlab = "",
          ylab = "% Antigen+ on AML Blasts",
          ylim = c(0, 100),
          fill = "white",
          outlier.shape = NA,
          legend.position = "none",
          palette = "jco") +
  geom_point(aes(color = highlight, alpha = highlight, size = highlight), position = position_jitter(seed = 42, width = 0.2)) +
  facet_grid(~ag) +
  theme_bw() + 
  theme(legend.position = "none", 
        axis.text = element_text(size = 15), 
        axis.title = element_text(size = 17), 
        strip.text = element_text(size = 15),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        strip.background = element_rect(fill="moccasin")) +
  scale_y_continuous(labels = c("0%", "25%", "50%", "75%", "100%")) +
  scale_color_manual(values = c("lightgray", "cyan3", "gold1")) +
  scale_alpha_manual(values = c(0.3, 1, 1)) +
  scale_size_manual(values = c(1.5, 4, 4))

# ag density
mypath <- "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/20240425_Antigen Density Matched Blasts Diagnosis vs  Relapse bone marrow .csv"
flow_ag_data <- aws.s3::s3read_using(readr::read_csv, object = mypath)
colnames(flow_ag_data) <- gsub(" ", "_", colnames(flow_ag_data))
flow_ag_data <- flow_ag_data %>%
  dplyr::mutate(patient = paste0("Patient_", patient),
                highlight = ifelse(CD33_Relapse < 1000 & CLL1_Relapse > 1000, patient, "none")) %>%
  tidyr::pivot_longer(-c(patient, highlight), names_to = "sample", values_to = "flow_antigen_density") %>%
  tidyr::separate(sample, c("ag", "timepoint")) %>%
  dplyr::mutate(ag = factor(ag, levels = c("CD33", "CLL1", "CD123", "EMR2")))

ggboxplot(data = flow_ag_data,
          x = "timepoint",
          y = "flow_antigen_density",
          xlab = "",
          ylab = "# of Antigens per Cell (ABC)",
          ylim = c(100, 100000),
          color = "black",
          fill = "white",
          outlier.shape = NA,
          legend.position = "none",
          palette = "lancet") +
  geom_point(aes(color = ag), position = position_jitter(seed = 42, width = 0.2)) +
  scale_y_log10(breaks = c(100, 500, 1000, 5000, 10000, 50000),
                labels = c("100", "500", "1,000", "5,000", "10,000", "50,000")) + 
  facet_grid(~ag) +
  theme_bw() + 
  theme(legend.position = "none", 
        axis.text = element_text(size = 15), 
        axis.title = element_text(size = 17), 
        strip.text = element_text(size = 15),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        strip.background = element_rect(fill="moccasin")) +
  scale_x_discrete(labels = c("Dx", "Re")) +
  geom_hline(yintercept = 1000, lty = 2) +
  scale_color_manual(values = c("firebrick2", "cornflowerblue", "mediumorchid1", "seagreen3"))



ggboxplot(data = flow_ag_data,
          x = "timepoint",
          y = "flow_antigen_density",
          xlab = "",
          ylab = "# of Antigens per Cell (ABC)",
          ylim = c(100, 100000),
          fill = "white",
          outlier.shape = NA,
          legend.position = "none",
          palette = "lancet") +
  geom_point(aes(color = highlight, alpha = highlight, size = highlight), position = position_jitter(seed = 42, width = 0.2)) +
  facet_grid(~ag) +
  theme_bw() + 
  scale_y_log10(breaks = c(100, 500, 1000, 5000, 10000, 50000),
                labels = c("100", "500", "1,000", "5,000", "10,000", "50,000")) + 
  facet_grid(~ag) +
  theme_bw() + 
  theme(legend.position = "right", 
        axis.text = element_text(size = 15), 
        axis.title = element_text(size = 17), 
        strip.text = element_text(size = 15),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        strip.background = element_rect(fill="moccasin")) +
  scale_x_discrete(labels = c("Dx", "Re")) +
  geom_hline(yintercept = 1000, lty = 2) +
  scale_color_manual(values = c("lightgray", "cyan3", "gold1")) +
  scale_alpha_manual(values = c(0.3, 1, 1)) +
  scale_size_manual(values = c(1.5, 4, 4))

```

## Assess correlation
```{r assess_accorrelation} 

abc_estimate <- imputed_abc[, colnames(imputed_abc) %in% c("CD33", "CLL1", "CD123", "EMR2")]

# imputed_abc_ag_out_list <- imputed_abc_ag_out_list[1:4]
# abc_estimate <- lapply(names(imputed_abc_ag_out_list), function(arg) imputed_abc_ag_out_list[[arg]][, arg, drop = FALSE])
# abc_estimate <- do.call("cbind", abc_estimate)
# colnames(abc_estimate) <- names(imputed_abc_ag_out_list)

abc_estimate <- abc_estimate %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  tidyr::pivot_longer(-barcode, values_to = "estimated_antigen_density", names_to = "ag") %>%
  dplyr::mutate(patient_id = gsub(".*-", "", barcode),
                patient_id = gsub("_Diagnosis|_Relapse", "", patient_id),
                timepoint = gsub(".*_", "", barcode)) %>%
  dplyr::left_join(fmo_df, by = c("patient_id" = "patient")) %>%
  dplyr::filter(estimated_antigen_density > fmo_ag_density_boundary) %>% # calculate avg Ag count in Ag+ blasts only
  dplyr::select(patient_id, timepoint, ag, estimated_antigen_density) %>%
  dplyr::group_by(patient_id, timepoint, ag) %>%
  dplyr::summarize(estimated_antigen_density = mean(estimated_antigen_density)) %>%
  dplyr::ungroup() %>%
  dplyr::inner_join(flow_ag_data, by = c("patient_id" = "patient", "ag", "timepoint")) %>%
  tidyr::unite(sample_id, patient_id, timepoint, sep = "_") %>%
  dplyr::mutate(deviation = abs(estimated_antigen_density - flow_antigen_density)) %>% 
  dplyr::arrange(desc(deviation)) %>%
  dplyr::mutate(ag = case_when(ag %in% "EMR2" ~ "ADGRE2",
                               ag %in% "CLL1" ~ "CLL-1",
                               TRUE ~ as.character(ag)),
                ag = factor(ag, levels = c("CD33", "CLL-1", "CD123", "ADGRE2")))

myrange <- c(min(c(abc_estimate$estimated_antigen_density, abc_estimate$flow_antigen_density), na.rm = TRUE), 
             max(c(abc_estimate$estimated_antigen_density, abc_estimate$flow_antigen_density), na.rm = TRUE))

plotdata <- dplyr::filter(abc_estimate, !is.na(flow_antigen_density))
ggscatter(data = plotdata,
          x = "estimated_antigen_density",
          y = "flow_antigen_density",
          xlab = "Estimated mean antigen number per blast",
          ylab = "Measured mean antigen number per blast",
          xlim = c(100, 100000),
          ylim = c(100, 100000),
          facet.by = "ag",
          color = "ag",
          cor.coef = TRUE,
          cor.method = "pearson",
          #cor.coeff.args = list(p.accuracy = 1e-19),
          palette = c("skyblue3", "hotpink", "purple", "olivedrab3"),
          size = 1.4,
          cor.coef.coord = c(log10(100), log10(65000))
        ) +
  scale_x_log10(breaks = c(100, 1000, 10000, 100000), labels = c("100", "1,000", "10,000", "100,000")) + 
  scale_y_log10(breaks = c(100, 1000, 10000, 100000), labels = c("100", "1,000", "10,000", "100,000")) +
  geom_abline(slope = 1, intercept = 0, lty = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

test_samples <- blast_data$sample_id[abs(blast_data$residuals) > 25]
abc_estimate <- dplyr::mutate(abc_estimate,
                              test_sample = ifelse(sample_id %in% test_samples, "Excluded from model training", "Included in model training"),
                              test_sample = factor(test_sample, levels = c("Included in model training", "Excluded from model training")))

ggscatter(data = abc_estimate,
          x = "estimated_antigen_density",
          y = "flow_antigen_density",
          xlab = "Estimated Ag density",
          ylab = "QuantiBRITE Ag density",
          xlim = myrange,
          ylim = myrange,
          facet.by = "test_sample",
          color = "ag",
          cor.coef = TRUE,
          cor.method = "pearson",
          palette = c("skyblue3", "orchid1", "purple", "olivedrab3"),
          size = 2
        ) +
  scale_x_continuous(breaks = c(1000, 5000, 10000, 15000, 20000)) +
  geom_abline(slope = 1, intercept = 0, lty = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

abc_estimate_remove_outlier_samples <- dplyr::filter(abc_estimate, !sample_id %in% c("Patient_50_Diagnosis", "Patient_41_Relapse"))
ggscatter(data = abc_estimate_remove_outlier_samples,
          x = "estimated_antigen_density",
          y = "flow_antigen_density",
          xlab = "Estimated Ag density",
          ylab = "QuantiBRITE Ag density",
          xlim = myrange,
          ylim = myrange,
          facet.by = "test_sample",
          color = "ag",
          cor.coef = TRUE,
          cor.method = "pearson",
          palette = c("skyblue3", "orchid1", "purple", "olivedrab3"),
          size = 2
        ) +
  scale_x_continuous(breaks = c(1000, 5000, 10000, 15000, 20000)) +
  geom_abline(slope = 1, intercept = 0, lty = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
                                                     
```

# Systematic evaluation of multiplex potential for all possible antigen pairs
```{r multiplex_potential, width = 17, height = 17}

ag_thr <- 1000
all_ags <- colnames(imputed_abc)
all_ags <- all_ags[!grepl("isotype", all_ags, ignore.case = TRUE)]
all_ags <- all_ags[!all_ags %in% c("CD34", "CD8A", "CD8", "CD19", "CD4", "CD3", "CD44", "CD99", "CD71", "CD93", "CD9", "CD34", "CD47", "CD49F", "CD56", "CD82", "CD58", "CD82", "CD119", "MESOTHELIN", "CD10", "GPR56", "HLA-DR", "CD13", "CD15", "CD26", "CD90", "CD117", "CD119", "CD133", "CD155", "CD41", "CD22")]
ncells <- nrow(imputed_abc)
multi_matrix <- matrix(NA, length(all_ags), length(all_ags))
rownames(multi_matrix) <- all_ags
colnames(multi_matrix) <- all_ags
multi_matrix_alt <- multi_matrix
multi_matrix_euc <- multi_matrix
p1_matrix <- p2_matrix <- multi_matrix
for (ag1 in 1:(length(all_ags) - 1)) {
  for (ag2 in (ag1 + 1):length(all_ags)) {
    
   ag1_fraction <- sum(imputed_abc[, which(colnames(imputed_abc) == all_ags[ag1])] > ag_thr) / ncells
   ag2_fraction <- sum(imputed_abc[, which(colnames(imputed_abc) == all_ags[ag2])] > ag_thr) / ncells
   
   if (ag2_fraction > ag1_fraction) {
     primary_ag <- all_ags[ag2]
     secondary_ag <- all_ags[ag1]
   } else {
     primary_ag <- all_ags[ag1]
     secondary_ag <- all_ags[ag2]
   }
   
   p1 <- sum(imputed_abc[, primary_ag] > ag_thr) / ncells
   p2 <- sum(imputed_abc[, primary_ag] < ag_thr & imputed_abc[, secondary_ag] > ag_thr) / ncells
   p3 <- sum(imputed_abc[, primary_ag] < ag_thr & imputed_abc[, secondary_ag] < ag_thr) / ncells
   
   metric <- 2*p2
   multi_matrix[ag2, ag1] <- metric
   multi_matrix_alt[ag2, ag1] <- log10(p1 + p2) + log10(p1) + log10(p2)
   multi_matrix_euc[ag2, ag1] <- 1 - (sqrt((p1 - 0.5)^2 + (p2 - 0.5)^2) / sqrt(0.5^2 + 0.5^2))
   p1_matrix[ag2, ag1] <- p1
   p2_matrix[ag2, ag1] <- p2
   
  }
}

x <- multi_matrix %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("ag1") %>%
  tidyr::pivot_longer(-ag1, names_to = "ag2", values_to = "simple") %>%
  dplyr::filter(!is.na(simple))

y <- multi_matrix_alt %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("ag1") %>%
  tidyr::pivot_longer(-ag1, names_to = "ag2", values_to = "log") %>%
  dplyr::filter(!is.na(log))

z <- multi_matrix_euc %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("ag1") %>%
  tidyr::pivot_longer(-ag1, names_to = "ag2", values_to = "euclidean") %>%
  dplyr::filter(!is.na(euclidean))

p1 <- p1_matrix %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("ag1") %>%
  tidyr::pivot_longer(-ag1, names_to = "ag2", values_to = "p1") %>%
  dplyr::filter(!is.na(p1))

p2 <- p2_matrix %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("ag1") %>%
  tidyr::pivot_longer(-ag1, names_to = "ag2", values_to = "p2") %>%
  dplyr::filter(!is.na(p2))

test <- dplyr::full_join(x, y, by = c("ag1", "ag2")) %>%
  dplyr::left_join(z, by = c("ag1", "ag2")) %>%
  dplyr::left_join(p1, by = c("ag1", "ag2")) %>%
  dplyr::left_join(p2, by = c("ag1", "ag2"))

plot_ly(test,
         x = ~p1,
         y = ~p2,
         z = ~simple,
         type = "scatter3d",
         mode = "markers",
         marker = list(size = 3))

plot_ly(test,
         x = ~p1,
         y = ~p2,
         z = ~log,
         type = "scatter3d",
         mode = "markers",
         marker = list(size = 3))

plot_ly(test,
         x = ~p1,
         y = ~p2,
         z = ~euclidean,
         type = "scatter3d",
         mode = "markers",
         marker = list(size = 3))



multi_matrix_heatdata <- multi_matrix
multi_matrix_heatdata[upper.tri(multi_matrix_heatdata)] <- t(multi_matrix_heatdata)[upper.tri(multi_matrix_heatdata)]
row_order <- hclust(dist(multi_matrix_heatdata))$order
col_order <- hclust(dist(t(multi_matrix_heatdata)))$order
multi_matrix_heatdata <- multi_matrix_heatdata[rev(row_order), rev(row_order)]

#multi_matrix_heatdata <- multi_matrix_heatdata[rowMeans(multi_matrix_heatdata, na.rm = TRUE) > 0.25, ]
multi_matrix_heatdata <- multi_matrix_heatdata[, rownames(multi_matrix_heatdata)]

multi_matrix_heatdata[upper.tri(multi_matrix_heatdata, diag = TRUE)] <- NA

split_lev <- case_when(colnames(multi_matrix_heatdata) == "CD33" ~ 1,
                       colnames(multi_matrix_heatdata) == "CLL-1" ~ 1,
                       TRUE ~ 0)
ag_split <- NULL
counter <- 0
ag_split[1] <- counter
for (i in 2:length(split_lev)) {
  if (split_lev[i] != split_lev[i-1]) {
    counter <- counter + 1
  }
  ag_split[i] <- counter
}

ComplexHeatmap::Heatmap(multi_matrix_heatdata, 
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                        #column_split = ag_split,
                        row_split = ag_split,
                        column_split = ag_split,
                        row_title = NULL,
                        column_title = NULL,
                        cluster_row_slices = FALSE,
                        cluster_column_slices = FALSE,
                        na_col = "white",
                        col = c("white", "red2"),
                        row_names_side = "left",
                        rect_gp = grid::gpar(col = "white", lwd = 1),
                        heatmap_legend_param = list(title = "Multi-targeting index"))


```




# Analyze CD33 and CLL1 co-expression
## Hexbin
```{r cd33_cll1_coexpression}

abc_coexpr <- imputed_abc[, colnames(imputed_abc) %in% c("CD33", "CLL1", "CD123", "EMR2")]
abc_coexpr <- abc_coexpr %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  tidyr::pivot_longer(-barcode, values_to = "estimated_antigen_density", names_to = "ag") %>%
  dplyr::mutate(patient_id = gsub(".*-", "", barcode),
                patient_id = gsub("_Diagnosis|_Relapse", "", patient_id),
                timepoint = gsub(".*_", "", barcode),
                ag = factor(ag, levels = c("CD33", "CLL1", "CD123", "EMR2"))) %>%
  dplyr::left_join(fmo_df, by = c("patient_id" = "patient")) %>%
  dplyr::filter(ag %in% c("CD33", "CLL1")) %>%
  dplyr::select(-fcs_file, -patient_id) %>%
  dplyr::mutate(positive = ifelse(estimated_antigen_density >= 1000, "yes", "no")) %>%
  dplyr::select(-estimated_antigen_density, -fmo_ag_density_boundary) %>%
  tidyr::pivot_wider(names_from = ag, values_from = positive) %>%
  dplyr::mutate(coexpressed = dplyr::case_when(CLL1 == "yes" & CD33 == "yes" ~ "CD33+ CLL-1+",
                                               CLL1 == "yes" & CD33 == "no" ~ "CD33neg CLL-1+",
                                               CLL1 == "no" & CD33 == "yes" ~ "CD33+ CLL-1neg",
                                               TRUE ~ "CD33neg CLL-1neg")) %>%
  dplyr::mutate(coexpressed = factor(coexpressed, levels = c("CD33+ CLL-1+", "CD33+ CLL-1neg", "CD33neg CLL-1+", "CD33neg CLL-1neg")))

diagnosis_table <- abc_coexpr %>%
  dplyr::filter(timepoint %in% "Diagnosis")
diagnosis_table <- table(diagnosis_table$coexpressed)
diagnosis_table <- diagnosis_table / sum(diagnosis_table) * 100

relapse_table <- abc_coexpr %>%
  dplyr::filter(timepoint %in% "Relapse")
relapse_table <- table(relapse_table$coexpressed)
relapse_table <- relapse_table / sum(relapse_table) * 100

blast_object <- dplyr::left_join(blast_object, dplyr::select(abc_coexpr, barcode, coexpressed), by = "barcode")
DimPlot(blast_object, 
        reduction = "umap", 
        group.by = "coexpressed", 
        split.by = "time", 
        cols = c("orange", "yellowgreen", "steelblue3", "forestgreen")) + 
  NoAxes()


# follow-up request from TC 2024Nov20

relapse_table_patient <- abc_coexpr %>%
  dplyr::filter(timepoint %in% "Relapse") %>%
  dplyr::mutate(patient = gsub(".*-", "", barcode)) %>%
  dplyr::select(patient, barcode, coexpressed) %>%
  dplyr::group_by(patient, coexpressed) %>%
  dplyr::summarize(count = n()) %>%
  dplyr::group_by(patient) %>%
  dplyr::mutate(percent = count / sum(count) * 100) %>%
  dplyr::ungroup() %>%
  dplyr::select(-count) %>%
  tidyr::pivot_wider(values_from = percent, names_from = coexpressed, values_fill = 0) %>%
  janitor::clean_names() %>%
  dplyr::arrange(desc(cd33neg_cll_1)) %>%
  dplyr::mutate(patient = gsub("_Relapse", "", patient),
                multiplex = ifelse(cd33neg_cll_1 > 10, "Yes", "No")) 

ggbarplot(relapse_table_patient,
          x = "patient",
          y = "cd33neg_cll_1",
          ylab = "% CD33neg CLL-1+ blasts",
          fill = "multiplex",
          palette = "jco",
          legend.position = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(breaks = seq(0, 100, 10))
  

# summary table

table_patient <- abc_coexpr %>%
  dplyr::mutate(sample_id = gsub(".*-", "", barcode)) %>%
  dplyr::select(sample_id, barcode, coexpressed) %>%
  dplyr::group_by(sample_id, coexpressed) %>%
  dplyr::summarize(count = n()) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(percent = count / sum(count) * 100) %>%
  dplyr::ungroup() %>%
  dplyr::select(-count) %>%
  tidyr::pivot_wider(values_from = percent, names_from = coexpressed, values_fill = 0)




bardata <- abc_coexpr %>%
  dplyr::mutate(sample_id = gsub(".*-", "", barcode)) %>%
  dplyr::select(sample_id, barcode, coexpressed) %>%
  dplyr::group_by(sample_id, coexpressed) %>%
  dplyr::summarize(count = n()) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(percent = count / sum(count) * 100) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(patient = gsub("_Diagnosis|_Relapse|atient_", "", sample_id),
                time = ifelse(grepl("Diagnosis", sample_id), "D", "R")) %>%
  dplyr::left_join(patient_id_map, by = "patient")

ggplot(bardata, aes(x = time, y = count, fill = coexpressed)) + 
  geom_bar(position = "fill", stat = "identity", width = 0.7) + 
  xlab("") +
  ylab("Fraction of blasts") +
  facet_grid(~patient_manuscript_id) +
  #scale_fill_manual(values = mycols) +
  theme_light() +
  theme(axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14)) +
  scale_fill_manual(values = c("orange", "yellowgreen", "steelblue3", "forestgreen"))

quad1 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CLL1 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad2 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CLL1 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad3 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CLL1 < 1000) / nrow(imputed_abc) * 100, 2), "%")
quad4 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CLL1 < 1000) / nrow(imputed_abc) * 100, 2), "%")

max_limit <- max(dplyr::select(imputed_abc, CD33, CLL1))

ggplot(data = imputed_abc, aes(x = CD33, y = CLL1)) +
  geom_hex(bins = 30) +  # Adjust bins for desired resolution
  #geom_density_2d(alpha = 0.3, color = "gray") +
  ggtitle("Estimated # of antigens on blast surface") + xlab("CD33") + ylab("CLL-1") +
  scale_x_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  scale_y_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  geom_vline(xintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_hline(yintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_vline(xintercept = 1000, color = "black") +
  geom_hline(yintercept = 1000, color = "black") +
  scale_fill_gradient(low = "white", high = "red3", name = "# of blasts") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = 60000, y = 60000, label = quad1, color = "black", size = 5) +
  annotate("text", x = 200, y = 60000, label = quad2, color = "black", size = 5) +
  annotate("text", x = 200, y = 150, label = quad3, color = "black", size = 5) +
  annotate("text", x = 60000, y = 150, label = quad4, color = "black", size = 5)


# CD33 with CD305
quad1 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD305 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad2 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD305 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad3 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD305 < 1000) / nrow(imputed_abc) * 100, 2), "%")
quad4 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD305 < 1000) / nrow(imputed_abc) * 100, 2), "%")

max_limit <- max(dplyr::select(imputed_abc, CD33, CD305))

ggplot(data = imputed_abc, aes(x = CD33, y = CD305)) +
  geom_hex(bins = 30) +  # Adjust bins for desired resolution
  ggtitle("Estimated # of antigens on blast surface") + xlab("CD33") + ylab("LAIR1") +
  scale_x_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  scale_y_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  geom_vline(xintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_hline(yintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_vline(xintercept = 1000, color = "black") +
  geom_hline(yintercept = 1000, color = "black") +
  scale_fill_gradient(low = "white", high = "red3", name = "# of blasts") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = 60000, y = 60000, label = quad1, color = "black", size = 5) +
  annotate("text", x = 200, y = 60000, label = quad2, color = "black", size = 5) +
  annotate("text", x = 200, y = 150, label = quad3, color = "black", size = 5) +
  annotate("text", x = 60000, y = 150, label = quad4, color = "black", size = 5)


# CD33 with CD49D
quad1 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD49D > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad2 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD49D > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad3 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD49D < 1000) / nrow(imputed_abc) * 100, 2), "%")
quad4 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD49D < 1000) / nrow(imputed_abc) * 100, 2), "%")

max_limit <- max(dplyr::select(imputed_abc, CD33, CD49D))

ggplot(data = imputed_abc, aes(x = CD33, y = CD49D)) +
  geom_hex(bins = 30) +  # Adjust bins for desired resolution
  ggtitle("Estimated # of antigens on blast surface") + xlab("CD33") + ylab("ITG4A") +
  scale_x_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  scale_y_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  geom_vline(xintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_hline(yintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_vline(xintercept = 1000, color = "black") +
  geom_hline(yintercept = 1000, color = "black") +
  scale_fill_gradient(low = "white", high = "red3", name = "# of blasts") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = 60000, y = 60000, label = quad1, color = "black", size = 5) +
  annotate("text", x = 200, y = 60000, label = quad2, color = "black", size = 5) +
  annotate("text", x = 200, y = 150, label = quad3, color = "black", size = 5) +
  annotate("text", x = 60000, y = 150, label = quad4, color = "black", size = 5)

# CD33 with CD205
quad1 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD205 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad2 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD205 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad3 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD205 < 1000) / nrow(imputed_abc) * 100, 2), "%")
quad4 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD205 < 1000) / nrow(imputed_abc) * 100, 2), "%")

max_limit <- max(dplyr::select(imputed_abc, CD33, CD205))

ggplot(data = imputed_abc, aes(x = CD33, y = CD205)) +
  geom_hex(bins = 30) +  # Adjust bins for desired resolution
  ggtitle("Estimated # of antigens on blast surface") + xlab("CD33") + ylab("DEC-205") +
  scale_x_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  scale_y_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  geom_vline(xintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_hline(yintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_vline(xintercept = 1000, color = "black") +
  geom_hline(yintercept = 1000, color = "black") +
  scale_fill_gradient(low = "white", high = "red3", name = "# of blasts") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = 60000, y = 60000, label = quad1, color = "black", size = 5) +
  annotate("text", x = 200, y = 60000, label = quad2, color = "black", size = 5) +
  annotate("text", x = 200, y = 150, label = quad3, color = "black", size = 5) +
  annotate("text", x = 60000, y = 150, label = quad4, color = "black", size = 5)

# CD33 with CD244
quad1 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD244 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad2 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD244 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad3 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD244 < 1000) / nrow(imputed_abc) * 100, 2), "%")
quad4 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD244 < 1000) / nrow(imputed_abc) * 100, 2), "%")

max_limit <- max(dplyr::select(imputed_abc, CD33, CD244))

ggplot(data = imputed_abc, aes(x = CD33, y = CD244)) +
  geom_hex(bins = 30) +  # Adjust bins for desired resolution
  ggtitle("Estimated # of antigens on blast surface") + xlab("CD33") + ylab("CD244") +
  scale_x_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  scale_y_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  geom_vline(xintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_hline(yintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_vline(xintercept = 1000, color = "black") +
  geom_hline(yintercept = 1000, color = "black") +
  scale_fill_gradient(low = "white", high = "red3", name = "# of blasts") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = 60000, y = 60000, label = quad1, color = "black", size = 5) +
  annotate("text", x = 200, y = 60000, label = quad2, color = "black", size = 5) +
  annotate("text", x = 200, y = 150, label = quad3, color = "black", size = 5) +
  annotate("text", x = 60000, y = 150, label = quad4, color = "black", size = 5)




# CD33 with CD16
quad1 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD16 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad2 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD16 > 1000) / nrow(imputed_abc) * 100, 2), "%")
quad3 <- paste0(round(sum(imputed_abc$CD33 < 1000 & imputed_abc$CD16 < 1000) / nrow(imputed_abc) * 100, 2), "%")
quad4 <- paste0(round(sum(imputed_abc$CD33 > 1000 & imputed_abc$CD16 < 1000) / nrow(imputed_abc) * 100, 2), "%")

max_limit <- max(dplyr::select(imputed_abc, CD33, CD16))

ggplot(data = imputed_abc, aes(x = CD33, y = CD16)) +
  geom_hex(bins = 30) +  # Adjust bins for desired resolution
  ggtitle("Estimated # of antigens on blast surface") + xlab("CD33") + ylab("CD16") +
  scale_x_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  scale_y_log10(limits = c(100, max_limit), breaks = c(100, 1000, 5000, 10000, 100000), labels = c("\u2264100", "1000", "5000", "10,000", "100,000")) +
  geom_vline(xintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_hline(yintercept = c(5000, 10000), linetype = "dotted", color = "black") +
  geom_vline(xintercept = 1000, color = "black") +
  geom_hline(yintercept = 1000, color = "black") +
  scale_fill_gradient(low = "white", high = "red3", name = "# of blasts") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = 60000, y = 60000, label = quad1, color = "black", size = 5) +
  annotate("text", x = 200, y = 60000, label = quad2, color = "black", size = 5) +
  annotate("text", x = 200, y = 150, label = quad3, color = "black", size = 5) +
  annotate("text", x = 60000, y = 150, label = quad4, color = "black", size = 5)






DefaultAssay(blast_object) <- "RNA"
CD33_rna <- FetchData(blast_object, vars = "CD33", layer = "data")
DefaultAssay(blast_object) <- "ADT"
CD33_adt <- FetchData(blast_object, vars = "CD33", layer = "data")
mydata <- data.frame("rna" = CD33_rna$CD33, "adt" = CD33_adt$CD33)

ggscatter(data = mydata[sample(1:nrow(mydata), 10000), ],
          title = "CD33",
          x = "rna",
          y = "adt",
          size = 0.1) +
  #scale_y_log10() +
  stat_density_2d(bins = 50)

DefaultAssay(blast_object) <- "ADT"
CD33 <- FetchData(blast_object, vars = "CD33", layer = "data")
CLL1 <- FetchData(blast_object, vars = "CLL1", layer = "data")
mydata <- data.frame("CD33" = CD33$CD33, "CLL1" = CLL1$CLL1)

ggscatter(data = mydata[sample(1:nrow(mydata), 5000), ],
          x = "CD33",
          y = "CLL1",
          size = 0.02) +
  stat_density_2d(bins = 10)


```

## Barplot
```{r barplot_coexpression}

find_possible_pairs <- function(features) {
  stopifnot(length(features) > 1)
  mypairs <- list()
  for (i in 1:(length(features) - 1)) {
    for (j in (i + 1):length(features)) {
      mypairs <- c(mypairs, list(c(features[i], features[j])))
    }
  }
  return(mypairs)
}

pairs <- find_possible_pairs(features = c("CD11A", "CD4"))

#selected_patient_data <- imputed_abc[grepl("Patient_70_Relapse", rownames(imputed_abc)), ]
selected_patient_data <- imputed_abc
#ag_thresholds <- c(300, 400, 500, 600, 700, 800, 900, 1000, 1100, 1200, 1300, 1400, 1500)
ag_thresholds <- seq(500, 10000, 500)
plotdata_list <- lapply(ag_thresholds, function(thr) {
  
  plotdata <- lapply(pairs, function(arg) {
    
    ag1 <- arg[1]
    ag2 <- arg[2]
    
    # union_counts <- sum(selected_patient_data[, arg[1]] > thr & selected_patient_data[, arg[2]] > thr)
    # ag1 <- sum(selected_patient_data[, arg[1]] > thr & !selected_patient_data[, arg[2]] > thr)
    # ag2 <- sum(!selected_patient_data[, arg[1]] > thr & selected_patient_data[, arg[2]] > thr)
    # uncovered <- sum(!selected_patient_data[, arg[1]] > thr & !selected_patient_data[, arg[2]] > thr)
    # 
    # myout <- c(union_counts, ag1, ag2, uncovered)
    
    uncovered <- sum((selected_patient_data[, ag1] + selected_patient_data[, ag2]) < thr)
    ag1_cover <- sum(selected_patient_data[, ag1] > thr)
    lift <- sum((selected_patient_data[, ag1] < thr) & ((selected_patient_data[, ag1] + selected_patient_data[, ag2]) > thr))
    myout <- c(uncovered, ag1_cover, lift)
    stopifnot(sum(myout) == nrow(selected_patient_data))
    myout <- myout / sum(myout) * 100
    return(myout)
    
  })
  
  plotdata <- do.call("rbind", plotdata)
  #colnames(plotdata) <- c("co-expressed", "CD33", "CLL-1", "below threshold")
  colnames(plotdata) <- c("Below threshold", "ag1", "ag2")
  plotdata <- as.data.frame(plotdata) %>%
    dplyr::mutate(pair = lapply(pairs, function(arg) paste(arg, collapse = " | ")) %>% unlist()) %>%
    tidyr::pivot_longer(-pair, names_to = "Coverage") %>%
    dplyr::mutate(threshold = thr)

})

plotdata <- do.call("rbind", plotdata_list)

plotdata <- plotdata %>%
  dplyr::arrange(threshold) %>%
  dplyr::mutate(Coverage = factor(Coverage, levels = c("Below threshold", "ag2", "ag1")),
                threshold = factor(paste(">", threshold), levels = paste(">", unique(threshold))))
levels(plotdata$Coverage) <- c("Below threshold", glue::glue("Expanded by {unlist(pairs)[2]}"), glue::glue("Covered by {unlist(pairs)[1]}"))

ggbarplot(data = plotdata,
          x = "threshold",
          xlab = "Antigen number per blast",
          y = "value",
          ylab = glue::glue("% of AML blasts"),
          fill = "Coverage",
          color = "white",
          legend = "top",
          size = 1.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(limits =  c(glue::glue("Covered by {unlist(pairs)[1]}"), glue::glue("Expanded by {unlist(pairs)[2]}"), "Below threshold"),
                    values = c("mediumturquoise", "purple", "white")) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  labs(fill = "")

```

# Generate UMAP plots showing Ag density at different thresholds
```{r generate_umap_ag_density_thresholds}

imputed_abc_ags <- imputed_abc[, c("CD33", "CLL1")]
colnames(imputed_abc_ags) <- c("CD33", "CLL-1")
stopifnot(all(rownames(imputed_abc_ags) == colnames(blast_object)))

# Patients from TC analysis that benefit from multiplex tx is 31, 40, and 60, 
sample_object <- blast_object %>% dplyr::filter(patient == "Patient_44")
Idents(sample_object) <- "time"
DefaultAssay(sample_object) <- "RNA"
sample_object <- FindVariableFeatures(sample_object)
# myfeats <- FindMarkers(sample_object, ident.1 = "Diagnosis") %>%
#   dplyr::arrange(abs(avg_log2FC)) %>%
#   dplyr::filter(p_val_adj < 1e-6) %>%
#   dplyr::slice(1:1000)
# myfeats <- rownames(myfeats)

sample_object <- ScaleData(sample_object)
sample_object <- RunPCA(sample_object)
sample_object <- RunUMAP(sample_object, dims = 1:50, seed.use = 81298)
DimPlot(sample_object, group.by = "time", reduction = "umap")

plot_list <- lapply(c(1000, 2000, 3000), function(arg) {
  
  tmp <- imputed_abc_ags %>%
    dplyr::mutate(count = rowSums(imputed_abc_ags > arg)) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("barcode") %>%
    dplyr::mutate(label = case_when(count == 0 ~ "Below threshold",
                                    count > 1 ~ "Multiple antigens",
                                    CD33 > arg ~ "CD33 only",
                                    `CLL-1` > arg ~ "CLL-1 only")) %>%
    dplyr::group_split(label)
  
  mynam <- lapply(tmp, function(arg) unique(arg$label)) %>% unlist
  tmp <- lapply(tmp, function(arg) dplyr::pull(arg, barcode))
  names(tmp) <- mynam
  tmp <- tmp[c("Below threshold", "Multiple antigens", "CD33 only", "CLL-1 only", "CD123 only", "ADGRE2 only")]
  names(tmp) <- paste(rev(1:length(tmp)), names(tmp), sep = " ")
  
  DimPlot(sample_object, 
          cells.highlight = tmp, 
          cols.highlight = c("yellow", "skyblue3", "hotpink", "purple", "olivedrab3"),
          reduction = "umap",
          sizes.highlight = 0.3,
          alpha = 0.6,
          group.by = "time",
          label = TRUE) +
   NoAxes() + ggtitle(NULL) + theme_bw() + guides(x = "none", y = "none") + xlab("") + ylab("")

})

plot_list[[1]] / plot_list[[2]] / plot_list[[3]]
  

```

# Test model on control cell populations
## T cell population
```{r test_model_tcell}

mymodel <- aws.s3::s3read_using(FUN = readRDS, 
                                object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/antigen_density_prediction_models/rf_model_2024-09-24.rds")

# test model on T cells to validate
input_data <- dplyr::filter(aml_data, cell_state_harmony_rna %in% c("CD4 T cell", "CD8 T cell"))@assays$ADT_dsb_qc_filter@data %>%
  limma::normalizeQuantiles() %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::mutate(sample_id = gsub(".*-", "", barcode)) %>%
  dplyr::left_join(ab_data, by = "sample_id") %>%
  dplyr::mutate(patient = gsub(".*-|_Diagnosis|_Relapse", "", barcode)) %>%
  dplyr::left_join(fmo_df, by = "patient") %>%
  dplyr::filter(sample_id %in% mymodel$trainingData$sample_id)

stopifnot(all(complete.cases(input_data)))
myfeats <- rownames(blast_object@assays$ADT_dsb_qc_filter@data)

tic()
imputed_abc_tcell <- foreach(arg = myfeats, .combine = "cbind", .inorder = TRUE, .packages = c("randomForest", "e1071",  "dplyr")) %dopar% {
  
  newdata <- dplyr::select(input_data, all_of(arg), contains("antibody"))
  colnames(newdata)[colnames(newdata) == arg] <- "abc_adt"
  mypred <- predict(mymodel[[1]], newdata = newdata)
  #mypred <- predict(og_mod, newdata = newdata)
  mypred <- 10^mypred
  mypred <- data.frame(mypred)
  colnames(mypred) <- arg
  gc()
  return(mypred)
  
}
toc()
rownames(imputed_abc_tcell) <- input_data$barcode

#sanity checks
iso_df <- dplyr::select(input_data, starts_with("ISOTYPE"))
iso_df <- data.frame(ADT_type = "Isotype control", mean_ag = rowMeans(iso_df))

myeloid_df  <- dplyr::select(input_data, !starts_with("ISOTYPE"))
myeloid_df <- data.frame(ADT_type = "Myeloid antibody", mean_ag = rowMeans(myeloid_df[, c("CD33", "CLL1", "CD123", "EMR2")]))

lym_df  <- dplyr::select(input_data, !starts_with("ISOTYPE"))
lym_df <- data.frame(ADT_type = "Lymphoid antibody", mean_ag = rowMeans(lym_df[, colnames(lym_df) %in% c("CD3", "CD4", "CD8", "CD19", "CD22")]))

ab_df  <- dplyr::select(input_data, !starts_with("ISOTYPE")) %>%
  dplyr::select(!starts_with("antibody"), -sample_id, -barcode, -patient, -fcs_file, -fmo_ag_density_boundary)
ab_df <- data.frame(ADT_type = "Other antibody", mean_ag = rowMeans(ab_df[, !colnames(ab_df) %in% c("CD33", "CLL1", "CD123", "EMR2", "CD3", "CD4", "CD8", "CD19", "CD22")]))

sanity_data <- bind_rows(myeloid_df, lym_df, ab_df, iso_df) %>%
  dplyr::filter(!is.na(mean_ag)) %>%
  dplyr::mutate(ADT_type = factor(ADT_type, levels = c("Myeloid antibody", "Lymphoid antibody", "Other antibody", "Isotype control")))

gghistogram(sanity_data, 
            x = "mean_ag", 
            xlab = "Average normalized ADT",
            ylab = "Number of cells",
            xlim = c(-1, 10),
            fill = "ADT_type", 
            color = "white", 
            bins = 100,
            palette = c("cornflowerblue", "forestgreen", "purple", "black"))

iso_df <- dplyr::select(imputed_abc_tcell, starts_with("ISOTYPE"))
iso_df <- data.frame(ADT_type = "Isotype control", mean_ag = rowMeans(iso_df))

myeloid_df  <- dplyr::select(imputed_abc_tcell, !starts_with("ISOTYPE"))
myeloid_df <- data.frame(ADT_type = "Myeloid antibody", mean_ag = rowMeans(myeloid_df[, c("CD33", "CLL1", "CD123", "EMR2")]))

lym_df  <- dplyr::select(imputed_abc_tcell, !starts_with("ISOTYPE"))
lym_df <- data.frame(ADT_type = "Lymphoid antibody", mean_ag = rowMeans(lym_df[, colnames(lym_df) %in% c("CD3", "CD4", "CD8", "CD19", "CD22")]))

ab_df  <- dplyr::select(imputed_abc_tcell, !starts_with("ISOTYPE"))
ab_df <- data.frame(ADT_type = "Other antibody", mean_ag = rowMeans(ab_df[, !colnames(ab_df) %in% c("CD33", "CLL1", "CD123", "EMR2", "CD3", "CD4", "CD8", "CD19", "CD22")]))

sanity_data <- bind_rows(myeloid_df, lym_df, ab_df, iso_df) %>%
  dplyr::filter(!is.na(mean_ag)) %>%
  dplyr::mutate(ADT_type = factor(ADT_type, levels = c("Other antibody", "Myeloid antibody", "Isotype control", "Lymphoid antibody")))

gghistogram(sanity_data, 
            x = "mean_ag", 
            xlab = "Average predicted antigen density",
            ylab = "Number of cells",
            xlim = c(0, 60000),
            fill = "ADT_type", 
            color = "white", 
            bins = 100,
            palette = c("purple", "cornflowerblue", "black", "forestgreen")) +
  geom_vline(xintercept = 500, lty = 2)

cd3_df  <- dplyr::select(imputed_abc_tcell, "CD3")

gghistogram(cd3_df, 
            x = "CD3", 
            main = "T cells",
            xlab = "Esimated # of CD3 Ags",
            ylab = "Number of cells",
            #xlim = c(0, 60000),
            fill = "darkgreen", 
            bins = 50) +
  geom_vline(xintercept = 500, lty = 2)

```

## B cell population
```{r test_model_bcell}

# test model on T cells to validate
input_data <- dplyr::filter(aml_data, cell_state_harmony_rna %in% c("B cell"))@assays$ADT_dsb_qc_filter@data %>%
  limma::normalizeQuantiles() %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::mutate(sample_id = gsub(".*-", "", barcode)) %>%
  dplyr::left_join(ab_data, by = "sample_id") %>%
  dplyr::mutate(patient = gsub(".*-|_Diagnosis|_Relapse", "", barcode)) %>%
  dplyr::left_join(fmo_df, by = "patient") %>%
  dplyr::filter(sample_id %in% og_mod$trainingData$sample_id)

stopifnot(all(complete.cases(input_data)))
myfeats <- rownames(blast_object@assays$ADT_dsb_qc_filter@data)

tic()
imputed_abc_tcell <- foreach(arg = myfeats, .combine = "cbind", .inorder = TRUE, .packages = c("randomForest", "e1071",  "dplyr")) %dopar% {
  
  newdata <- dplyr::select(input_data, all_of(arg), contains("antibody"))
  colnames(newdata)[colnames(newdata) == arg] <- "abc_adt"
  mypred <- predict(gbmFit1, newdata = newdata)
  #mypred <- predict(og_mod, newdata = newdata)
  mypred <- 10^mypred
  mypred <- data.frame(mypred)
  colnames(mypred) <- arg
  gc()
  return(mypred)
  
}
toc()
rownames(imputed_abc_tcell) <- input_data$barcode

#sanity checks
iso_df <- dplyr::select(input_data, starts_with("ISOTYPE"))
iso_df <- data.frame(ADT_type = "Isotype control", mean_ag = rowMeans(iso_df))

myeloid_df  <- dplyr::select(input_data, !starts_with("ISOTYPE"))
myeloid_df <- data.frame(ADT_type = "Myeloid antibody", mean_ag = rowMeans(myeloid_df[, c("CD33", "CLL1", "CD123", "EMR2")]))

lym_df  <- dplyr::select(input_data, !starts_with("ISOTYPE"))
lym_df <- data.frame(ADT_type = "Lymphoid antibody", mean_ag = rowMeans(lym_df[, colnames(lym_df) %in% c("CD3", "CD4", "CD8", "CD19", "CD22")]))

ab_df  <- dplyr::select(input_data, !starts_with("ISOTYPE")) %>%
  dplyr::select(!starts_with("antibody"), -sample_id, -barcode, -patient, -fcs_file, -fmo_ag_density_boundary)
ab_df <- data.frame(ADT_type = "Other antibody", mean_ag = rowMeans(ab_df[, !colnames(ab_df) %in% c("CD33", "CLL1", "CD123", "EMR2", "CD3", "CD4", "CD8", "CD19", "CD22")]))

sanity_data <- bind_rows(myeloid_df, lym_df, ab_df, iso_df) %>%
  dplyr::filter(!is.na(mean_ag)) %>%
  dplyr::mutate(ADT_type = factor(ADT_type, levels = c("Myeloid antibody", "Lymphoid antibody", "Other antibody", "Isotype control")))

gghistogram(sanity_data, 
            x = "mean_ag", 
            xlab = "Average normalized ADT",
            ylab = "Number of cells",
            xlim = c(-1, 10),
            fill = "ADT_type", 
            color = "white", 
            bins = 100,
            palette = c("cornflowerblue", "forestgreen", "purple", "black"))

iso_df <- dplyr::select(imputed_abc_tcell, starts_with("ISOTYPE"))
iso_df <- data.frame(ADT_type = "Isotype control", mean_ag = rowMeans(iso_df))

myeloid_df  <- dplyr::select(imputed_abc_tcell, !starts_with("ISOTYPE"))
myeloid_df <- data.frame(ADT_type = "Myeloid antibody", mean_ag = rowMeans(myeloid_df[, c("CD33", "CLL1", "CD123", "EMR2")]))

lym_df  <- dplyr::select(imputed_abc_tcell, !starts_with("ISOTYPE"))
lym_df <- data.frame(ADT_type = "Lymphoid antibody", mean_ag = rowMeans(lym_df[, colnames(lym_df) %in% c("CD3", "CD4", "CD8", "CD19", "CD22")]))

ab_df  <- dplyr::select(imputed_abc_tcell, !starts_with("ISOTYPE"))
ab_df <- data.frame(ADT_type = "Other antibody", mean_ag = rowMeans(ab_df[, !colnames(ab_df) %in% c("CD33", "CLL1", "CD123", "EMR2", "CD3", "CD4", "CD8", "CD19", "CD22")]))

sanity_data <- bind_rows(myeloid_df, lym_df, ab_df, iso_df) %>%
  dplyr::filter(!is.na(mean_ag)) %>%
  dplyr::mutate(ADT_type = factor(ADT_type, levels = c("Other antibody", "Myeloid antibody", "Isotype control", "Lymphoid antibody")))

gghistogram(sanity_data, 
            x = "mean_ag", 
            xlab = "Average predicted antigen density",
            ylab = "Number of cells",
            xlim = c(0, 60000),
            fill = "ADT_type", 
            color = "white", 
            bins = 100,
            palette = c("purple", "cornflowerblue", "black", "forestgreen")) +
  geom_vline(xintercept = 500, lty = 2)

cd19_df  <- dplyr::select(imputed_abc_tcell, "CD19")

gghistogram(cd19_df, 
            x = "CD19", 
            main = "T cells",
            xlab = "Esimated # of CD3 Ags",
            ylab = "Number of cells",
            #xlim = c(0, 60000),
            fill = "darkgreen", 
            bins = 50) +
  geom_vline(xintercept = 500, lty = 2)

```

## Test on engineered cell lines
```{r engineered_cell_lines}

eng_cell_data_matrix <- aws.s3::s3read_using(FUN = Seurat::Read10X_h5, 
                                      object = "s3://vorbio-analysis/single_cell/internal/DME-240603_VH00296_91_AACHTTVHV/aggregated/outs/count/filtered_feature_bc_matrix.h5")
 
raw_feature <- aws.s3::s3read_using(FUN = Seurat::Read10X_h5,
                                    object = "s3://vorbio-analysis/single_cell/internal/DME-240603_VH00296_91_AACHTTVHV/aggregated/outs/count/raw_feature_bc_matrix.h5")

aggr_csv <- aws.s3::s3read_using(FUN = readr::read_csv,
                                 object = "s3://vorbio-analysis/single_cell/internal/DME-240603_VH00296_91_AACHTTVHV/aggregated/outs/aggregation.csv") %>%
  dplyr::mutate(suffix = as.character(1:4))

metrics <- aws.s3::get_bucket_df(bucket = "s3://vorbio-analysis",
                                       prefix = "single_cell/internal/DME-240603_VH00296_91_AACHTTVHV/") %>%
  dplyr::filter(grepl("metrics_summary.csv", Key))

metrics_list <- lapply(metrics$Key, function(arg) aws.s3::s3read_using(FUN = readr::read_csv, object = glue::glue("s3://vorbio-analysis/{arg}")))
metrics <- do.call("rbind", metrics_list)
metrics <- dplyr::mutate(metrics, suffix = as.character(1:4))

eng_cell_data <- CreateSeuratObject(counts = eng_cell_data_matrix[["Gene Expression"]], assay = "RNA", min.cells = 10)

eng_cell_data[["percent.mt"]] <- Seurat::PercentageFeatureSet(eng_cell_data, pattern = "^MT-")
eng_cell_data[["percent.malat1"]] <- Seurat::PercentageFeatureSet(eng_cell_data, pattern = "MALAT1")
eng_cell_data[["percent.rpl_rps"]] <- Seurat::PercentageFeatureSet(eng_cell_data, pattern = "^RPL|^RPS")

select_features <- rownames(eng_cell_data@assays$RNA)[!grepl("^MT-|^RPS|^RPL|MALAT1", rownames(eng_cell_data@assays$RNA))]
eng_cell_data <- subset(eng_cell_data, features = select_features)
eng_cell_data <- dplyr::mutate(eng_cell_data, 
                               barcode = colnames(eng_cell_data),
                               suffix = gsub(".*-", "", barcode)) %>%
  dplyr::left_join(aggr_csv, by = "suffix") %>%
  dplyr::left_join(metrics, by = "suffix")

adt_matrix <- eng_cell_data_matrix[["Antibody Capture"]]
rownames(adt_matrix) <- gsub("\\..*", "", rownames(adt_matrix))
eng_cell_data[["ADT"]] <- CreateAssayObject(counts = adt_matrix, assay = "ADT")

eng_cell_data <- dplyr::filter(eng_cell_data,
                               nCount_RNA > 1000,
                               nFeature_RNA > 500,
                               percent.mt < 15)

eng_cell_data <- NormalizeData(eng_cell_data, assay = "RNA")
eng_cell_data <- NormalizeData(eng_cell_data, assay = "ADT", normalization.method = "CLR", margin = 2)


# select droplets based on RNA and ADT library size
bg_adt_matrix <- raw_feature[["Antibody Capture"]]
bg_rna_matrix <- raw_feature[["Gene Expression"]]
adt_size <- Matrix::colSums(bg_adt_matrix)
rna_size <- Matrix::colSums(bg_rna_matrix)
names(rna_size) <- gsub(".*\\.", "", names(rna_size))
overlap_bc <- intersect(names(adt_size), names(rna_size))
adt_size <- adt_size[names(adt_size) %in% overlap_bc]
rna_size <- rna_size[names(rna_size) %in% overlap_bc]
stopifnot(all(names(adt_size) == names(rna_size)))

qc_df <- data.frame(rna_size = rna_size, adt_size = adt_size) %>%
  dplyr::filter(rna_size > 0 & adt_size > 0)

mysample <- qc_df[sample(1:nrow(qc_df), 20000), ]
ggscatter(data = mysample,
          x = "adt_size",
          y = "rna_size",
          alpha = 0.1) +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 10000)) +
  scale_y_log10() +
  stat_density2d()

# keep only empty droplets for background matrix
bg_adt_matrix <- bg_adt_matrix[, !colnames(bg_adt_matrix) %in% colnames(eng_cell_data)]
# remove droplets with zero signal
bg_adt_matrix <- bg_adt_matrix[, colSums(bg_adt_matrix) > 10]
# remove droplets near filter threshold based on histogram of droplet count sums
bg_adt_matrix <- bg_adt_matrix[, colSums(bg_adt_matrix) < 1000]
rownames(bg_adt_matrix) <- gsub("\\..*", "", rownames(bg_adt_matrix))
rownames(bg_adt_matrix) <- gsub("_", "-", rownames(bg_adt_matrix))
# normalize and de-noise with dsb
dsb_norm_mat <- dsb::DSBNormalizeProtein(
  
  cell_protein_matrix = eng_cell_data@assays$ADT@counts, 
  empty_drop_matrix = bg_adt_matrix, 
  denoise.counts = TRUE, 
  use.isotype.control = TRUE, 
  isotype.control.name.vec = c("Isotype-IgG1K", "Isotype-Rat", "Isotype-Armenian-Hamster"),
  pseudocount.use = 10
  
)

eng_cell_data[["ADT_dsb"]] <- CreateAssayObject(data = dsb_norm_mat)
eng_cell_data[["ADT_dsb_qn"]]  <- CreateAssayObject(limma::normalizeQuantiles(eng_cell_data@assays$ADT_dsb@data))

ab_data <- eng_cell_data@meta.data %>%
  as.data.frame() %>%
  janitor::clean_names() %>%
  dplyr::mutate_all(function(arg) gsub("%", "", arg)) %>%
  dplyr::select(contains("antibody")) %>%
  dplyr::mutate_at(vars(contains("antibody")), .funs = as.numeric)

input_data <- eng_cell_data@assays$ADT_dsb_qn@data %>% t() %>% as.data.frame()
myfeats <- colnames(input_data)
input_data <- cbind(input_data, ab_data) %>% as.data.frame()

model <- aws.s3::s3readRDS(object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/antigen_density_prediction_models/gbm_model_2024-04-24.rds")[[1]]
tic()
imputed_abc <- foreach::foreach(arg = myfeats, .combine = "cbind", .inorder = TRUE, .packages = c("randomForest", "e1071",  "dplyr", "caret")) %dopar% {
  
  newdata <- dplyr::select(input_data, all_of(arg), contains("antibody"))
  colnames(newdata)[colnames(newdata) == arg] <- "abc_adt"
  mypred <- predict(model, newdata = newdata)
  mypred <- 10^mypred
  mypred <- data.frame(mypred)
  colnames(mypred) <- arg
  gc()
  return(mypred)
  
}
toc()

rownames(imputed_abc) <- rownames(input_data)

eng_cell_data[["predicted_ABC"]] <- CreateAssayObject(data = t(as.matrix(imputed_abc)))

DefaultAssay(eng_cell_data) <- "predicted_ABC"
eng_cell_data@assays$predicted_ABC@counts <- eng_cell_data@assays$predicted_ABC@data
VlnPlot(eng_cell_data, group.by = 'sample_id', features = 'CD33', pt.size = 0, assay = 'predicted_ABC')


DefaultAssay(eng_cell_data) <- "ADT"
eng_cell_data <- eng_cell_data %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors() %>%
  FindClusters() %>%
  RunUMAP(dims = 1:20)

DimPlot(eng_cell_data, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoAxes() + NoLegend()
FeaturePlot(eng_cell_data, features = "CD33", reduction = "umap")

```

# Evaluate model performance on new Ags
## Calculate correlation
```{r compare_with_QB}

myags <- c("CD33", "CD305", "CD49D", "CD205", "CD244")

qb_abc <- aws.s3::s3read_using(FUN = readr::read_csv, 
                               object = "s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/quantibrite_model_ag_count_validation_blast_20241217.csv") %>%
  dplyr::mutate(sample_id = paste0("Patient_", sample_id),
                time = ifelse(grepl("D$", sample_id), "Diagnosis", "Relapse"),
                sample_id = gsub("-.*", "", sample_id)) %>%
  tidyr::unite(sample_id, sample_id, time, sep = "_") %>%
  tidyr::pivot_longer(-sample_id, names_to = "antigen", values_to = "counts") %>%
  dplyr::mutate(platform = "QuantiBRITE", antigen = toupper(antigen)) %>%
  dplyr::filter(antigen %in% myags)

grep_pattern <- paste(unique(qb_abc$sample_id), collapse = "|")

pat_imputed_abc <- imputed_abc[grepl(grep_pattern, rownames(imputed_abc)), colnames(imputed_abc) %in% myags]
pat_imputed_abc <- pat_imputed_abc %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::mutate(sample_id = gsub(".*-", "", barcode)) %>%
  tidyr::pivot_longer(c(-barcode, -sample_id), names_to = "antigen", values_to = "counts") %>%
  dplyr::group_by(sample_id, antigen) %>%
  dplyr::filter(counts > 250) %>% # Set FMO cutoff; This number is equivalent to average FMO Ags/cell across batches
  dplyr::summarize(counts = mean(counts)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(antigen = factor(antigen, levels = myags),
                platform = "CITEseq estimation")

plotdata <- dplyr::bind_rows(pat_imputed_abc, qb_abc) %>%
  dplyr::mutate(antigen = case_when(antigen %in% "CD33" ~ "CD33",
                                    antigen %in% "CD305" ~ "LAIR1",
                                    antigen %in% "CD49D" ~ "ITGA4",
                                    antigen %in% "CD205" ~ "DEC-205",
                                    antigen %in% "CD244" ~ "CD244")) %>%
  dplyr::mutate(antigen = factor(antigen, levels = c("CD33", "LAIR1", "ITGA4", "DEC-205", "CD244")),
                patient = gsub("atient|_|Diagnosis|Relapse", "", sample_id)) %>%
  dplyr::left_join(patient_id_map, by = "patient") %>%
  dplyr::mutate(sample_manuscript_id = ifelse(grepl("Diagnosis", sample_id), paste0(patient_manuscript_id, "D"), paste0(patient_manuscript_id, "R"))) %>%
  dplyr::select(sample_manuscript_id, antigen, platform, counts)

ggpar(ggbarplot(plotdata,
          x = "antigen",
          y = "counts",
          xlab = "",
          ylab = "Mean # of antigens/cell",
          fill = "platform",
          facet.by = "sample_manuscript_id",
          color = "white",
          position = position_dodge(width = 0.5),
          palette = "jco",
          ncol = 5,
          legend = "top") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_log10(), legend = "top")

ggbarplot(dplyr::filter(plotdata, !antigen %in% "CD33", sample_manuscript_id %in% c("P10R", "P14D")),
          x = "antigen",
          y = "counts",
          xlab = "",
          ylab = "Avg # of antigens per cell",
          #ylim = c(1, 20000),
          fill = "platform",
          facet.by = "sample_manuscript_id",
          color = "white",
          position = position_dodge(width = 0.5),
          palette = "jco",
          ncol = 5) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 10000, lty = 2) +
  scale_y_log10()

scatterdata <- tidyr::pivot_wider(plotdata, names_from = "platform", values_from = "counts")

colnames(scatterdata) <- c("Sample ID", "Ag", "model", "qb")
mybreaks <- c(0, 5000, 10000, 15000, 20000, 25000, 30000, 40000)
mylabels <- c("0", "5,000", "10,000", "15,000", "20,000", "25,000", "30,000", "40,000")
ggscatter(data = scatterdata,
          x = "model",
          y = "qb",
          xlim = c(0, 40000),
          ylim = c(0, 40000),
          xlab = "Mean estimated # antigens/cell",
          ylab = "Mean measured # antigens/cell",
          color = "Sample ID",
          shape = "Ag",
          size = 4,
          alpha = 0.7,
          cor.coef = TRUE,
          cor.method = "pearson",
          cor.coef.coord = c(1000, 28000),
          cor.coeff.args = list(p.accuracy = 1e-21),
          palette = "npg") +
  geom_abline(slope = 1, lty = 2) +
  scale_x_continuous(breaks = mybreaks, labels = mylabels) +
  scale_y_continuous(breaks = mybreaks, labels = mylabels) +
  theme_bw() +
  theme(axis.text = element_text(angle = 45, hjust = 1, size = 13),
        axis.title = element_text(size = 13)) 

```
## Retrain model to include validation data
```{r retrain_model}



```




# Apply antigen density prediction approach on early cell line CITEseq data
## Get cell line flow data and convert to Ag density using QuantiBRITE
```{r cell_line_test}

# april flow cytometry data
flow_paths_april <- aws.s3::get_bucket(bucket = "vorbio-datascience",
                                       prefix = "genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/cell_line/210416",
                                       max = Inf)
flow_paths_april <- lapply(flow_paths_april, `[`, 1) %>% unlist()
flow_paths_april <- flow_paths_april[grepl(".fcs", flow_paths_april)]
flow_paths_april <- paste0("s3://vorbio-datascience/", flow_paths_april)
mynames <- gsub("s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/cell_line/|.fcs", "", flow_paths_april)

flow_list_april <- lapply(flow_paths_april, function(object_path) aws.s3::s3read_using(FUN = flowCore::read.FCS,
                                                                                       transformation = FALSE,
                                                                                       truncate_max_range = TRUE,
                                                                                       object = object_path))
flow_list_april <- lapply(flow_list_april, function(arg) flowCore::transform(arg, `Y586-A` = asinh(`Y586-A`)))
names(flow_list_april) <- mynames
flow_qb_april <- flow_list_april[grepl("Quantibrite", names(flow_list_april))] %>% unlist()
flow_list_april <- flow_list_april[!grepl("Quantibrite", names(flow_list_april))]

# may flow cytometry data
flow_paths_may <- aws.s3::get_bucket(bucket = "vorbio-datascience",
                                       prefix = "genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/cell_line/210525",
                                       max = Inf)
flow_paths_may <- lapply(flow_paths_may, `[`, 1) %>% unlist()
flow_paths_may <- flow_paths_may[grepl(".fcs", flow_paths_may)]
flow_paths_may <- paste0("s3://vorbio-datascience/", flow_paths_may)
mynames <- gsub("s3://vorbio-datascience/genomics/internal_data/single_cell_meta_analysis/aml_hematology_atlas/flow_data/cell_line/|.fcs", "", flow_paths_may)

flow_list_may <- lapply(flow_paths_may, function(object_path) aws.s3::s3read_using(FUN = flowCore::read.FCS,
                                                                                         transformation = FALSE,
                                                                                         truncate_max_range = TRUE,
                                                                                         object = object_path))
flow_list_may <- lapply(flow_list_may, function(arg) flowCore::transform(arg, `Y586-A` = asinh(`Y586-A`)))
names(flow_list_may) <- mynames
flow_qb_may <- flow_list_may[grepl("Quantibrite", names(flow_list_may))] %>% unlist()
flow_list_may <- flow_list_may[!grepl("Quantibrite", names(flow_list_may))]

# Convert FI to antigen density
flow_qb_cell_line <- c(flow_qb_april, flow_qb_may)

plot_list <- lapply(names(flow_qb_cell_line), function(arg) {
  data.frame(fcs_file = arg,
             PE = exprs(flow_qb_cell_line[[arg]])[, "Y586-A"])
})
names(plot_list) <- names(flow_qb_cell_line)

plot_df <- do.call("rbind", plot_list) %>%
  dplyr::mutate(fcs_file = gsub("\\/export.*", "", fcs_file))
rownames(plot_df) <- NULL

ggdensity(data = plot_df,
          x = "PE",
          xlab = "asinh PE",
          color = "fcs_file",
          legend = "right",
          palette = "simpsons")

# compute PE peak means for each sample
set.seed(465012)
sample_qb_mean <- lapply(names(plot_list), function(arg) {
  
  pe_mean <- sort(mclust::Mclust(data = plot_list[[arg]]$PE, G = 4)$parameters$mean, decreasing = FALSE)
  data.frame(fcs_file = arg, peak = 1:4, pe_mean = pe_mean)
  
})

plot_df <- do.call("rbind", sample_qb_mean) %>%
  dplyr::left_join(qb_levels, by = "peak") %>%
  dplyr::mutate(fcs_file = gsub("\\/export.*", "", fcs_file))

ggscatter(data = plot_df,
          x = "pe_mean",
          y = "ag_density",
          ylim = c(50, 100000),
          xlab = "Mean asinh PE intensity",
          ylab = "Log antigen density",
          color = "fcs_file",
          legend = "right",
          palette = "lancet",
           size = 3) +
  scale_y_log10(labels = scales::comma) +
  geom_smooth(se = FALSE, method = "lm", lty = 2, color = "black", linewidth = 0.5)

ggscatter(data = plot_df,
          x = "pe_mean",
          y = "ag_density",
          xlab = "Mean asinh PE intensity",
          ylab = "Log antigen density",
          facet.by = "fcs_file",
          legend = "right",
          palette = "lancet") +
  scale_y_log10()

qb_models_cell_line <- plot_df %>% 
  dplyr::group_by(fcs_file) %>%
  dplyr::group_split()

mynam <- sapply(qb_models_cell_line, function(arg) unique(arg$fcs_file))
qb_models_cell_line <- lapply(qb_models_cell_line, function(arg) lm(log10(ag_density) ~ pe_mean, data = arg))
names(qb_models_cell_line) <- mynam

```

## Ag density prediction and validation
```{r ag_density_cell_line_validation}

# compute Ag quantiles
myprobs <- 0.1
df_list <- NULL
df_quantile_list <- NULL
df_mean_list <- NULL
df_median_list <- NULL
for (ag in c("CD33", "CLL1", "CD123")) {
  
  flow_list_ags_select <- flow_list_april[grepl(ag, names(flow_list_april))]

  for (mynam in names(flow_list_ags_select)) {
    
    model_id <- gsub("\\/.*", "", mynam)
    sample_id <- gsub(".*\\/", "", mynam)
    sample_id <- gsub("_Live", "", sample_id)
    sample_id <- gsub(".*_", "", sample_id)
    
    # extract FIs from flowCore object
    fluor <- flow_list_ags_select[[mynam]]@parameters@data
    fluor <- as.character(fluor$name[fluor$desc %in% ag])
    if (length(fluor) == 0) fluor = "Y586-A"
    fi <- exprs(flow_list_ags_select[[mynam]])
    fi <- fi[, colnames(fi) %in% fluor]
    
    abc_model <- qb_models_cell_line[[model_id]]
    abc <- predict(object = abc_model, newdata = data.frame(pe_mean = fi))
    
    if (length(abc) == 0) next
    
    # compute quantiles
    myq <- quantile(abc, probs = seq(0, 1, myprobs), na.rm = TRUE)
    df_list[[mynam]] <- data.frame(sample_id, ag, abc) # raw FI values
    df_quantile_list[[mynam]] <- data.frame(sample_id, ag, quantile = names(myq), abc = myq) # ABC quantiles
    df_median_list[[mynam]] <- data.frame(sample_id, ag, quantile = "50%", abc = median(abc))
    
    # quantile-binned averages
    myq_bin <- NULL
    for (q in 2:length(myq)) {
      myq_bin[paste0("bin_", q-1)] <- mean(abc[abc >= myq[q-1] & abc < myq[q]])
    }
      
    df_mean_list[[mynam]] <- data.frame(sample_id, ag, quantile_bin = names(myq_bin), abc = myq_bin) 

  }
}

# raw FIs
fi_df <- do.call("rbind", df_list)

fi_quant_df <- do.call("rbind", df_quantile_list)



#### Prepare CITEseq ADT data ####
# cell line paths
#experiment <- "single_cell/internal/SCSEQ-NTD-210525"
experiment <- "single_cell/internal/SCSEQ-NTD-210416"

cell_line_paths <- aws.s3::get_bucket(bucket = "vorbio-analysis", prefix = experiment, max = Inf)
cell_line_paths <- lapply(cell_line_paths, `[`, 1) %>% unlist()
cell_line_paths <- cell_line_paths[grepl("Jurkat|Molm-13|HL-60|THP-1", cell_line_paths, ignore.case = TRUE)]
cell_line_paths <- cell_line_paths[!grepl("reanalyzed|aggregated", cell_line_paths, ignore.case = TRUE)]

bg_matrix <- lapply(c("Jurkat", "Molm-13", "HL-60", "THP-1"), function(arg) {
  
  cell_line_matrix <- cell_line_paths[grepl("raw_feature_bc_matrix.h5", cell_line_paths) & grepl(arg, cell_line_paths, ignore.case = TRUE)]
  cell_line_matrix <- cell_line_matrix[grepl("outs", cell_line_matrix)]
  cell_line_matrix <- aws.s3::s3read_using(FUN = Read10X_h5, 
                                           object = paste0("s3://vorbio-analysis/", cell_line_matrix))[["Antibody Capture"]]
  colnames(cell_line_matrix) <- paste(colnames(cell_line_matrix), arg, sep = "_")
  return(cell_line_matrix)
  
})
bg_matrix <- do.call("cbind", bg_matrix)

cell_protein_matrix <- lapply(c("Jurkat", "Molm-13", "HL-60", "THP-1"), function(arg) {
  
  cell_line_matrix <- cell_line_paths[grepl("filtered_feature_bc_matrix.h5", cell_line_paths) & grepl(arg, cell_line_paths)]
  cell_line_matrix <- aws.s3::s3read_using(FUN = Read10X_h5, 
                                           object = paste0("s3://vorbio-analysis/", cell_line_matrix))[["Antibody Capture"]]
  colnames(cell_line_matrix) <- paste(colnames(cell_line_matrix), arg, sep = "_")
  return(cell_line_matrix)
  
})
cell_protein_matrix <- do.call("cbind", cell_protein_matrix)

# keep only empty droplets for background matrix
bg_matrix <- bg_matrix[, !colnames(bg_matrix) %in% colnames(cell_protein_matrix)]
# remove droplets with zero signal
bg_matrix <- bg_matrix[, colSums(bg_matrix) > nrow(bg_matrix)]
# remove droplets near filter threshold based on histogram of droplet count sums
#bg_matrix <- bg_matrix[, colSums(bg_matrix) < 1000]

stopifnot(all(rownames(bg_matrix) %in% rownames(cell_protein_matrix)))

# normalize and de-noise with dsb
dsb_norm_mat <- dsb::DSBNormalizeProtein(
  
  cell_protein_matrix = cell_protein_matrix, 
  empty_drop_matrix = bg_matrix, 
  denoise.counts = TRUE, 
  use.isotype.control = TRUE, 
  #isotype.control.name.vec = c("IgG1_K_Isotype_TotalSeqC", "IgG2a_K_Isotype_TotalSeqC"),
  isotype.control.name.vec = c("IgG1_K_Isotype_TotalSeqC"),
  pseudocount.use = 1,
  quantile.clipping = TRUE,

)

qn_data <- limma::removeBatchEffect(x = dsb_norm_mat, covariates = colSums(cell_protein_matrix))
#qn_data <- limma::normalizeQuantiles(qn_data)
blast_df <- dsb_norm_mat[rownames(qn_data) %in% c("CD33_P67.6_TotalSeqC", "CD123_6H6_TotalSeqC", "CD371_50C1_TotalSeqC"), ]
blast_df <- t(blast_df)
colnames(blast_df) <- c("CD33", "CD123", "CLL1")

blast_list <- blast_df %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::mutate(sample_id = gsub(".*_|-", "", cell)) %>%
  dplyr::select(-cell) %>%
  tidyr::pivot_longer(-sample_id, names_to = "ag", values_to = "abc") %>%
  dplyr::group_split(sample_id, ag)

blast_list <- lapply(blast_list, function(arg) {
  
  adt_abc <- arg$abc
  #adt_abc <- adt_abc[adt_abc > 0]
  myq <- quantile(adt_abc, probs = seq(0, 1, myprobs))
  mydf <- data.frame(sample_id = arg$sample_id[1], ag = arg$ag[1], quantile = names(myq), abc = myq)
  return(mydf)
  
})

blast_quantiles_df <- do.call("rbind", blast_list)

fi_comdata <- dplyr::right_join(fi_quant_df, blast_quantiles_df, 
                                by = c("sample_id", "ag", "quantile"), 
                                suffix = c("_flow", "_adt"))

trim_data_quantiles <- filter(
  fi_comdata, 
  !is.na(abc_flow)
) %>% dplyr::mutate(ag_density = 10^abc_flow)

ggpubr::ggscatter(data = trim_data_quantiles,
                  x = "abc_adt",
                  y = "ag_density",
                  color = "sample_id",
                  alpha = 0.3,
                  xlab = glue::glue("DSB normalized ADT"),
                  ylab = glue::glue("Antigen density"),
                  cor.coef = TRUE,
                  cor.method = "spearman",
                  cor.coeff.args = list(size = 4),
                  cor.coef.coord = c(4, 0.00001),
                  size = 3,
                  facet.by = "ag") +
  geom_hline(yintercept = 100, lty = 3) +
  scale_y_log10(breaks = c(1, 10, 100, 1000, 10000, 100000), labels = c("1", "10", "100", "1,000", "10,000", "100,000")) +
  scale_x_continuous(breaks = -1:15) +
  geom_vline(xintercept = 0, lty = 3) +
  stat_density_2d(alpha = 0.2) 

# cell_line_ab_metrics <- c("s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210525/210525_VH00296_5_AAAFG2THV_THP-1/outs/metrics_summary.csv",
#                           "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210525/210525_VH00296_5_AAAFG2THV_HL-60/outs/metrics_summary.csv",
#                           "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210525/210525_VH00296_5_AAAFG2THV_Molm-13/outs/metrics_summary.csv",
#                           "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210525/210525_VH00296_5_AAAFG2THV_Jurkat/outs/metrics_summary.csv")

cell_line_ab_metrics <- c("s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_THP-1/outs/metrics_summary.csv",
                          "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_HL-60/outs/metrics_summary.csv",
                          "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_Molm-13/outs/metrics_summary.csv",
                          "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_Jurkat/outs/metrics_summary.csv")

cell_line_ab_metrics <- lapply(ab_metrics, function(arg) aws.s3::s3read_using(object = arg, FUN = readr::read_csv))
cell_line_ab_metrics <- do.call("rbind", cell_line_ab_metrics) %>%
  janitor::clean_names() %>%
  dplyr::mutate_all(.funs = function(arg) gsub("%", "", arg)) %>%
  dplyr::mutate_if(is.character, .funs = as.numeric) %>%
  dplyr::mutate(sample_id = c("THP1", "HL60", "Molm13", "Jurkat"))

training_data <- cell_line_ab_metrics %>% 
  dplyr::right_join(fi_comdata, by = c("sample_id"))

training_data_for_model <- training_data %>%
  dplyr::select(-sample_id, -ag, -quantile)

set.seed(69420)                                                         
myfit <- randomForest::randomForest(abc_flow ~ ., data = training_data_for_model, ntree = 1000, type = "regression")









# Test on separate CITEseq dataset (SCSEQ-NTD-210416)
# experiment <- "single_cell/internal/SCSEQ-NTD-210416"
# 
# cell_line_paths <- aws.s3::get_bucket(bucket = "vorbio-analysis", prefix = experiment, max = Inf)
# cell_line_paths <- lapply(cell_line_paths, `[`, 1) %>% unlist()
# cell_line_paths <- cell_line_paths[grepl("Jurkat|Molm-13|HL-60|THP-1", cell_line_paths, ignore.case = TRUE)]
# cell_line_paths <- cell_line_paths[!grepl("reanalyzed|aggregated", cell_line_paths, ignore.case = TRUE)]
# 
# bg_matrix <- lapply(c("Jurkat", "Molm-13", "HL-60", "THP-1"), function(arg) {
#   
#   cell_line_matrix <- cell_line_paths[grepl("raw_feature_bc_matrix.h5", cell_line_paths) & grepl(arg, cell_line_paths, ignore.case = TRUE)]
#   cell_line_matrix <- cell_line_matrix[grepl("outs", cell_line_matrix)]
#   cell_line_matrix <- aws.s3::s3read_using(FUN = Read10X_h5, 
#                                            object = paste0("s3://vorbio-analysis/", cell_line_matrix))[["Antibody Capture"]]
#   colnames(cell_line_matrix) <- paste(colnames(cell_line_matrix), arg, sep = "_")
#   return(cell_line_matrix)
#   
# })
# bg_matrix <- do.call("cbind", bg_matrix)
# 
# cell_protein_matrix <- lapply(c("Jurkat", "Molm-13", "HL-60", "THP-1"), function(arg) {
#   
#   cell_line_matrix <- cell_line_paths[grepl("filtered_feature_bc_matrix.h5", cell_line_paths) & grepl(arg, cell_line_paths)]
#   cell_line_matrix <- aws.s3::s3read_using(FUN = Read10X_h5, 
#                                            object = paste0("s3://vorbio-analysis/", cell_line_matrix))[["Antibody Capture"]]
#   colnames(cell_line_matrix) <- paste(colnames(cell_line_matrix), arg, sep = "_")
#   return(cell_line_matrix)
#   
# })
# cell_protein_matrix <- do.call("cbind", cell_protein_matrix)
# 
# # keep only empty droplets for background matrix
# bg_matrix <- bg_matrix[, !colnames(bg_matrix) %in% colnames(cell_protein_matrix)]
# # remove droplets with zero signal
# bg_matrix <- bg_matrix[, colSums(bg_matrix) > nrow(bg_matrix)]
# # remove droplets near filter threshold based on histogram of droplet count sums
# #bg_matrix <- bg_matrix[, colSums(bg_matrix) < 1000]
# 
# stopifnot(all(rownames(bg_matrix) %in% rownames(cell_protein_matrix)))
# 
# # normalize and de-noise with dsb
# dsb_norm_mat <- dsb::DSBNormalizeProtein(
#   
#   cell_protein_matrix = cell_protein_matrix, 
#   empty_drop_matrix = bg_matrix, 
#   denoise.counts = FALSE, 
#   use.isotype.control = FALSE, 
#   #isotype.control.name.vec = c("IgG1_K_Isotype_TotalSeqC", "IgG2a_K_Isotype_TotalSeqC"),
#   #isotype.control.name.vec = c("IgG1_K_Isotype_TotalSeqC"),
#   pseudocount.use = 1,
#   quantile.clipping = TRUE,
# 
# )
# 
# cell_line_ab_metrics <- c("s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_THP-1/outs/metrics_summary.csv",
#                           "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_HL-60/outs/metrics_summary.csv",
#                           "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_Molm-13/outs/metrics_summary.csv",
#                           "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/210416_VH00296_5_AAAFG2THV_Jurkat/outs/metrics_summary.csv")
# 
# cell_line_ab_metrics <- lapply(ab_metrics, function(arg) aws.s3::s3read_using(object = arg, FUN = readr::read_csv))
# cell_line_ab_metrics <- do.call("rbind", cell_line_ab_metrics) %>%
#   janitor::clean_names() %>%
#   dplyr::mutate_all(.funs = function(arg) gsub("%", "", arg)) %>%
#   dplyr::mutate_if(is.character, .funs = as.numeric) %>%
#   dplyr::mutate(sample_id = c("THP1", "HL60", "Molm13", "Jurkat"))

#dsb_norm_mat_qn <- limma::normalizeQuantiles(dsb_norm_mat)
#dsb_norm_mat_qn <- qn_data <- limma::removeBatchEffect(x = dsb_norm_mat, covariates = colSums(cell_protein_matrix))

input_data <- blast_df %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::mutate(sample_id = gsub(".*_|-", "", barcode)) %>%
  dplyr::left_join(cell_line_ab_metrics, by = "sample_id")

tic()
myfeats <- c("CD33", "CLL1", "CD123")
imputed_abc <- foreach(arg = myfeats, .combine = "cbind", .inorder = TRUE, .packages = c("randomForest", "e1071",  "dplyr")) %dopar% {
  
  newdata <- input_data
  colnames(newdata)[colnames(newdata) == arg] <- "abc_adt"
  mypred <- predict(myfit, newdata = newdata)
  mypred <- 10^mypred
  mypred <- data.frame(mypred)
  colnames(mypred) <- arg
  return(mypred)
  
}
toc()

imputed_abc <- imputed_abc %>%
  dplyr::mutate(barcode = input_data$barcode) %>%
  dplyr::mutate(sample_id = gsub(".*_|-", "", barcode))

boxplot(CD33 ~ sample_id, data = imputed_abc)
boxplot(CD123 ~ sample_id, data = imputed_abc)
boxplot(CLL1 ~ sample_id, data = imputed_abc)

boxplot(IgG1_K_Isotype_TotalSeqC ~ sample_id, data = imputed_abc)
boxplot(IgG2a_K_Isotype_TotalSeqC ~ cell_line, data = imputed_abc)


# aws.s3::s3read_using(FUN = pzfx::read_pzfx, 
#                      object = "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210416/20210416_CD33_CLL1_CD123_quantibrite analysis_cell lines.pzfx")
# 
# aws.s3::s3read_using(FUN = pzfx::read_pzfx, 
#                      object = "s3://vorbio-analysis/single_cell/internal/SCSEQ-NTD-210525/20210514_CD33_CLL1_CD123_quantibrite analysis_cell lines.pzfx")

```
